### 2015-09-21 paper.Rmd
Create the paper Rmd file and other stuff

---
title: "School accessibility in Sao Paulo"
author: "Ana Moreno and Robin Lovelace"
date: "Monday, September 21, 2015"
output: pdf_document
---

# Introduction

Blah blah

### 2015-09-22 R/load_OD_2007.R
Upload and scan OD 2007 data

@@ -1,0 +1,12 @@
+# Aim: Read (.mdb) OD 2007 data
+
+setwd("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/")
+
+data<-read.table("OD_2007.txt", sep=",")
+
+colnames(data)<-c("ZONA",  "MUNI_DOM",	"CO_DOM_X",	"CO_DOM_Y",	"ID_DOM",	"F_DOM",	"FE_DOM",	"DOM",	"CD_ENTRE",	"DATA",	"TIPO_DOM",	"NO_MORAD",	"TOT_FAM",	"ID_FAM",	"F_FAM",	"FE_FAM",	"FAMILIA",	"NO_MORAF",	"CONDMORA",	"QT_RADIO",	"QT_GEL1",	"QT_GEL2",	"QT_TVCOR",	"QT_FREEZ",	"QT_VIDEO",	"QT_BANHO",	"QT_MOTO",	"QT_AUTO",	"QT_ASPIR",	"QT_MLAVA",	"QT_EMPRE",	"QT_MICRO",	"QT_BICICLE",	"NAO_DCL_IT",	"CRITERIO_B",	"ANO_AUTO1",	"ANO_AUTO2",	"ANO_AUTO3",	"RENDA_FA",	"CD_RENFA",	"ID_PESS",	"F_PESS",	"FE_PESS",	"PESSOA",	"SIT_FAM",	"IDADE",	"SEXO",	"ESTUDA",	"GRAU_INS",	"CD_ATIVI",	"CO_REN_I",	"VL_REN_I",	"ZONA_ESC",	"MUNIESC",	"CO_ESC_X",	"CO_ESC_Y",	"TIPO_ESC",	"ZONATRA1",	"MUNITRA1",	"CO_TR1_X",	"CO_TR1_Y",	"TRAB1_RE",	"TRABEXT1",	"OCUP1",	"SETOR1",	"VINC1",	"ZONATRA2",	"MUNITRA2",	"CO_TR2_X",	"CO_TR2_Y",	"TRAB2_RE",	"TRABEXT2",	"OCUP2",	"SETOR2",	"VINC2",	"N_VIAG",	"FE_VIA",	"DIA_SEM",	"TOT_VIAG",	"ZONA_O",	"MUNI_O",	"CO_O_X",	"CO_O_Y",	"ZONA_D",	"MUNI_D",	"CO_
[…]

### 2015-09-22 paper.Rmd
Add abstract!

@@
 title: "School accessibility  in Sao Paulo"
+title: "School accessibility: a new index with a case study of enrollment and academic performance in Sao Paulo"
 author: "Ana Moreno and Robin Lovelace"
@@
 
+# Abstract
+
+Travel to school is an everyday reality for millions of young people around the world.
+The mode, duration, levels
+
+Access to high quality quality schools is highly variable across geographical
+space so travel options are vital for ensuring a more equitable supply of
+educational opportunity to diverse groups. Conversely, poor accessibility
+to deprived area can reinforce social inequalities, with long-term implications.
+
+Based on these emerging findings, this paper sets out a school accessibility index.
+This index is based on current travel patterns, local travel infrastructure serving the school
+catchment and the provision of travel services such as public transport links.
+Using data from the Household Travel Survey (2007 and 2012), the School Census
+and electronic card data, we apply this metric to schools in Sao Paulo.
+
+Its impact on the social background of school attendees.
+A wider aim of this paper is to explore the potential links between accessibility
+and academic acheivement and opportunities.
+
 # Introduction

### 2015-09-24 paper.Rmd
Update abstract - 1st draft

@@
 Travel to school is an everyday reality for millions of young people around the world.
-The mode, duration, levels
-
+The mode, duration, safety, comfort and pollution levels of this trip has huge
+impacts on the future generation, yet has received relatively little academic
+attention.
 Access to high quality quality schools is highly variable across geographical
-space so travel options are vital for ensuring a more equitable supply of
+space, making travel options are vital for ensuring a more equitable supply of
 educational opportunity to diverse groups. Conversely, poor accessibility
 to deprived area can reinforce social inequalities, with long-term implications.
-
-Based on these emerging findings, this paper sets out a school accessibility index.
+Based on this emerging context, this paper develops a school accessibility index
+for local areas.
 This index is based on current travel patterns, local travel infrastructure serving the school
 catchment and the provision of travel services such as public transport links.
-Using data from the Household Travel Survey (2007 and 2012), the School Census
-and electronic card data, we apply this metric to schools in Sao Paulo.
-
-Its impact on the social background of school attendees.
-A wider aim of this paper is to explore the potential links between accessibility
-and academic acheivement and opportunities.
-
-# Introduction
-
-Blah blah
+Using data from the Sao Paulo Household Travel Survey (2007 and 2012), School Census
+and electronic card data, we demonstrate how this index can be used
+measure current provision and evaluate future interventions.
+The results suggest that that school accessibility can
+impact the social background of school attendees and
+academic performance.
+We conclude that extending the concept of local accessibility indicators
+to the 'school commute' can ensure that transport decision makers take
+equality and health impacts on young people into account.

### 2015-09-24 paper.Rmd

@@
 ---
-title: "School accessibility  in Sao Paulo"
-title: "School accessibility: a new index with a case study of enrollment and academic performance in Sao Paulo"
+title: "School accessibility: a new index applied to enrollment and performance inequalities in Sao Paulo"
 author: "Ana Moreno and Robin Lovelace"
-date: "Monday, September 21, 2015"
+date: "`r format(Sys.time(), '%F')`"
 output: pdf_document
+bibliography: school-access-bib.bib
 ---
@@
 equality and health impacts on young people into account.
+
+# Introduction
+
+**Contents:**
+
+- Motivations
+- General measures of accessibility
+- School accessibility
+
+...
+
+The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
+
+This early work made the distinction between accessibility indeces that apply to zones or single 'desire lines':
+"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area"
+[@allen1993accesibility].
+
+# Method: measuring school accessibility
+
+# Results
+
+# Discussion
+
+# Conclusion
+
+# References

### 2015-09-28 paper.Rmd
Add equation

@@
 ---
-title: "School accessibility: a new index applied to enrollment and performance inequalities in Sao Paulo"
+title: 'School accessibility: a new index applied to enrollment and performance inequalities
+  in Sao Paulo'
 author: "Ana Moreno and Robin Lovelace"
-date: "`r format(Sys.time(), '%F')`"
 output: pdf_document
@@
 ---
+
+```{r, echo=FALSE}
+#date: '`r format(Sys.time(), ''%F'')`'
+```
 
@@
 
+$$
+\tau=2
+$$
+
 # Results

### 2015-09-29 paper.Rmd
Final version abstract JoTG

@@
 ---
-title: 'School accessibility: a new index applied to enrollment and performance inequalities
-  in Sao Paulo'
-author: "Ana Moreno and Robin Lovelace"
-output: pdf_document
+title: 'School accessibility: a new index applied to the case of São Paulo'
+author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação
+  Getulio Vargas) and Robin Lovelace (University of Leeds)
+output: word_document
 bibliography: school-access-bib.bib
@@
 
-Travel to school is an everyday reality for millions of young people around the world.
-The mode, duration, safety, comfort and pollution levels of this trip has huge
-impacts on the future generation, yet has received relatively little academic
-attention.
-Access to high quality quality schools is highly variable across geographical
-space, making travel options are vital for ensuring a more equitable supply of
-educational opportunity to diverse groups. Conversely, poor accessibility
-to deprived area can reinforce social inequalities, with long-term implications.
-Based on this emerging context, this paper develops a school accessibility index
-for local areas.
-This index is based on current travel patterns, local travel infrastructure serving the school
-catchment and the provision of travel services such as public transport links.
-Using data from the Sao Paulo Household Travel Survey (2007 and 2012), School Census
-and electronic card data, we demonstrate how this index can be used
-measure current provision and evaluate future interventions.
-The results suggest that that school accessibility can
-impact the social background of school attendees and
-academic performance.
-We conclude that extending the concept of local accessibility indicators
-to the 'school commute' can ensure that transport decision makers take
-equality and health impacts on young people into account.
+A skewed distribution of schools towards wealthy areas can result in a mismatch between the provision of educational opportunity and the need for access to education.
+This mismatch can be further reinforced by poor transport provision in lower income areas.
+In this paper, we develop a school accessibility index that captures both the unequal distribution of schools and transport services across space.
+Our index combines information on the spatial distribution of children and adolescents by income categories, the location of existing schools, the travel patterns of students, and the travel infrastructure serving the school catchment area into a single measure of school accesibility.
+We use our index to measure school accessibility by areas across São Paulo, based on information from Population and School Censuses, Household Travel Surveys and electronic transit card records.
+The results illustrate how existing inequalities are amplified by variable accessibility to schools across income groups and geographical space. 
+We conclude that extending the concept of local accessibility indicators to the 'school commute' can guide the development of policies to reduce educational inequalities.
 
-# Introduction
+<!-- # Introduction
 
@@
 
+Travel to school is an everyday reality for millions of young people around the world.
+The mode, duration, safety, comfort and pollution levels of this trip has huge
+impacts on the future generation, yet has received relatively little academic
+attention.
+
+Travel options are vital for ensuring a more equitable supply of
+educational opportunity to diverse groups. Conversely, poor accessibility
+to deprived area can reinforce social inequalities, with long-term implications.
+Based on this emerging context, this paper develops a school accessibility index
+for local areas.
+
 The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
 
-This early work made the distinction between accessibility indeces that apply to zones or single 'desire lines':
+This early work made the distinction between accessibility indexes that apply to zones or single 'desire lines':
 "relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area"
@@
 # References
+
+-->

### 2015-10-04 paper.Rmd
Various updates

--- 2015-09-28 paper.Rmd
@@
 author: "Ana Moreno and Robin Lovelace"
-output: pdf_document
+output: word_document
 bibliography: school-access-bib.bib

### 2016-02-08 R/load_schools_data.R
Update schools loading

@@ -2,7 +2,13 @@
 
 library(raster)
 
-schools <- shapefile("../SaoPaulo/Data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.shp")
+file.rename("/tmp/data/escolas_SP_2008/ESCOLAS_DE_2008_GEORREFERENCIADAS.dbf", "/tmp/data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.dbf")
+schools <- shapefile("/tmp/data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.shp")
 
+proj4string(schools) <- CRS("+init=epsg:4326")
+
+library(tmap)
+library(leaflet)
+leaflet() %>% addTiles() %>% addCircles(data = schools)
 plot(schools)
 head

### 2016-02-08 R/load_schools_data.R
Add data loading scripts

@@ -2,11 +2,29 @@
 
 library(raster)
 
+# 1 loading data
 file.rename("/tmp/data/escolas_SP_2008/ESCOLAS_DE_2008_GEORREFERENCIADAS.dbf", "/tmp/data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.dbf")
 schools <- shapefile("/tmp/data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.shp")
-
 proj4string(schools) <- CRS("+init=epsg:4326")
 
+zones <- shapefile("/tmp/data/Shapefiles SPMR 2010/aep10.shp")
+zones <- spTransform(zones, proj4string(schools))
+
+od <- 
+
+plot(zones)
+points(schools, col = "red")
+schools <- schools[zones,] # select only those in zones
+points(schools, col = "green")
+nrow(schools)
+df <- schools@data
+View(df)
+plot(df$ALE_1_4 + df$ALE_5_8, df$TOT_MATR)
+cor(df$ALE_1_4 + df$ALE_5_8, df$TOT_MATR, use = "complete.obs")
+
+# which schools to use
+
+# Visualising data
 library(tmap)
 library(leaflet)
 leaflet() %>% addTiles() %>% addCircles(data = schools)

### 2016-02-09 R/load_OD_2007.R
analyze OD data

@@ -1,13 +1,16 @@
 # Aim: Read (.mdb) OD 2007 data
 
-od_zone <- shapefile("/tmp/data/OD-2007/Mapas OD/ZONAS_od07.shp")
-od_flow <- read.table("/tmp/Data/OD-2007/OD_2007.txt", sep=",")
+
+library(raster)
+
+od_zone <- shapefile("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
+od_flow <- read.table("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/OD_2007.txt", sep=",")
 
 colnames(od_flow) <- c("ZONA",  "MUNI_DOM",	"CO_DOM_X",	"CO_DOM_Y",	"ID_DOM",	"F_DOM",	"FE_DOM",	"DOM",	"CD_ENTRE",	"DATA",	"TIPO_DOM",	"NO_MORAD",	"TOT_FAM",	"ID_FAM",	"F_FAM",	"FE_FAM",	"FAMILIA",	"NO_MORAF",	"CONDMORA",	"QT_RADIO",	"QT_GEL1",	"QT_GEL2",	"QT_TVCOR",	"QT_FREEZ",	"QT_VIDEO",	"QT_BANHO",	"QT_MOTO",	"QT_AUTO",	"QT_ASPIR",	"QT_MLAVA",	"QT_EMPRE",	"QT_MICRO",	"QT_BICICLE",	"NAO_DCL_IT",	"CRITERIO_B",	"ANO_AUTO1",	"ANO_AUTO2",	"ANO_AUTO3",	"RENDA_FA",	"CD_RENFA",	"ID_PESS",	"F_PESS",	"FE_PESS",	"PESSOA",	"SIT_FAM",	"IDADE",	"SEXO",	"ESTUDA",	"GRAU_INS",	"CD_ATIVI",	"CO_REN_I",	"VL_REN_I",	"ZONA_ESC",	"MUNIESC",	"CO_ESC_X",	"CO_ESC_Y",	"TIPO_ESC",	"ZONATRA1",	"MUNITRA1",	"CO_TR1_X",	"CO_TR1_Y",	"TRAB1_RE",	"TRABEXT1",	"OCUP1",	"SETOR1",	"V
[…]

### 2016-02-09 paper.Rmd

--- 2015-09-28 paper.Rmd
@@
 ---
-title: 'School accessibility: a new index applied to enrollment and performance inequalities
-  in Sao Paulo'
-author: "Ana Moreno and Robin Lovelace"
-output: pdf_document
+title: 'The unequal distribution of access to education and transport:
+  insights from for São Paulo'
+author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação
+  Getulio Vargas) and Robin Lovelace (University of Leeds)
+output: 
+  pdf_document: 
+    keep_tex: yes
 bibliography: school-access-bib.bib
@@
 
-Travel to school is an everyday reality for millions of young people around the world.
-The mode, duration, safety, comfort and pollution levels of this trip has huge
-impacts on the future generation, yet has received relatively little academic
-attention.
-Access to high quality quality schools is highly variable across geographical
-space, making travel options are vital for ensuring a more equitable supply of
-educational opportunity to diverse groups. Conversely, poor accessibility
-to deprived area can reinforce social inequalities, with long-term implications.
-Based on this emerging context, this paper develops a school accessibility index
-for local areas.
-This index is based on current travel patterns, local travel infrastructure serving the school
-catchment and the provision of travel services such as public transport links.
-Using data from the Sao Paulo Household Travel Survey (2007 and 2012), School Census
-and electronic card data, we demonstrate how this index can be used
-measure current provision and evaluate future interventions.
-The results suggest that that school accessibility can
-impact the social background of school attendees and
-academic performance.
-We conclude that extending the concept of local accessibility indicators
-to the 'school commute' can ensure that transport decision makers take
-equality and health impacts on young people into account.
+In many large Latin American cities such as São Paulo, growing social
+and economic inequalities are embedded through unfair education and
+transport systems. Good schools are mostly concentrated in wealthy
+areas, while transport links to school are deficient in deprived areas,
+exacerbating the issue. Inequalities in educational and transport
+infrastructure are mutually reinforcing: the right to mobility is
+intrinsically linked to the right to education. This is manifested by
+the overlap between recent protests against unwanted changes to public
+education and the social movements contesting increases in public
+transport fares. Another manifestation is to be found in the concept of
+school accessibility. This paper sheds light on the transport-education
+inequality nexus with reference to a new school accessibility measure
+applied São Paulo. By capturing both the unequal distribution of schools
+and transport services across space, the index allows embedded
+inequalities to be better understood and, with political will,
+contested. Our index combines information on the spatial distribution of
+children and adolescents by income categories, the location of existing
+schools, the travel patterns of students, and the travel infrastructure
+serving the school catchment area into a single measure. The index is
+used to measure school accessibility locally across São Paulo, using
+data sources from Population and School Censuses, commonly available in
+Latin American cities. The results illustrate how existing inequalities
+are amplified by variable accessibility to schools across income groups
+and geographical space. We conclude that extending the concept of local
+accessibility indicators to education can help to both contest and
+constructively tackle embedded social inequalities.
 
@@
 
+Travel to school is an everyday reality for millions of young people around the world.
+The mode, duration, safety, comfort and pollution levels of this trip has huge
+impacts on the future generation, yet has received relatively little academic
+attention.
+
+Travel options are vital for ensuring a more equitable supply of
+educational opportunity to diverse groups. Conversely, poor accessibility
+to deprived area can reinforce social inequalities, with long-term implications.
+Based on this emerging context, this paper develops a school accessibility index
+for local areas.
+
 The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
@@
 
-$$
-\tau=2
+We calculate three different accessibility measures for different modes: cumulative, gravitiy-based, and competitive.
+
+The cumulative-opportunity measure for mode $M$ (Boisjoly and El-Geneidy, 2016) is defined as:
+
+$$ CO_{i}^M= \sum_{j=1}^n(S_{j}f(C_{ij}) $$
+
+$$ f(C_{ij}) = \left\{ 
+                \begin{array}{ll}
+                  1 if C_{ij}<=t\\
+                  0 if C_{ij}<t
+                \end{array}
+              \right.
   $$
 
+This measure counts the number of schools available from one area within a certain travel time by mode $M$ threshold.
+$C_{ij}$ is the travel cost (measured in time) between the centroid of zone i and the centroid of zone j.
+
+Gravity-based accessibility measure
+
+In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found that the cumulative-opportunity measure (using a 45 minute threshold) is highly correlated with a gravity-based accessibility measure ([@handy1997measuring], [@hansen1959accessibility]), defined as:
+
+$$ A_{i}^M= \sum_{j=1}^n(S_{j}e^{\beta*C_{ij}}) $$
+
+Where $A_{i}^M$ is the gravity accessibility at the centroid of area $i$ to all schools at area $j$ using mode $M$, and $\beta$ is a negative exponential cost function.
+The cost function, based on a negative exponential decay curve, includes reported number of school trips in the 2007 Origin-Destination Survey.
+(Here we can also use information from the school census on number of students per school to know how many students the area attracts).
+The proposed accessibility measure captures the fact that more proximate schools (S) are more attractive than those located further away.
+
+Competitive accessibility measure
+
+The measures proposed above take into account the sptail distribution of schooling opportunities, but not the local demand for schooling.
+This is particularly relevant for our equity analysis, since it could be the case that higher areas are disproportionally served with respect to the number of potentail students living within a certain travel distance, whereas the opposite holds for lower-income areas.
+In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (the demand), and the sum of individuals within the school grade age-group living in each area (the supply) in the following competitive accessibility measure, first proposed by [@shen1998location] and adapted by El-Geneidy et al (2015):
+
+$$ CA_{i}^M= \sum_{j=1}^n\frac{P_{j}-f(C_{ij})}{\sum_{k=1}^n(Y_{k}-f(C_{kj}))}$$
+
+Where the numerator discounts the number of pupils in area $j$ $P_{j}$ by how far area $i$ is from area $j$ using the same function as before, and the denominator discounts the number of students (young adults) living in zone $k$ $Y_{k}$ by how far they are from area $j$.
+In this way, the discounted number of study places at each area is divided by the potential students available to fill those places, and then summed in order to obtain a single accessibility measure for each area $i$.
+
+In order to assess equality in access to schooling, we stratify our population of students in two student groups (we could do it for more income groups): the students living with a head of household who earns 1 minimum wage or less, and those living with a head of household who earns more than 1 minimum wage.
+The accessibility measures are based on student's travel times by different modes using data from the 2007 Origin-Destination survey.
+Alternatively, for the case of public transport, wet use actual transit times in 2015 for comparison (bilhete unico data?).
+
 # Results
+
+- Descriptive statistics about modal choice and travel time by AEPs and level of income
+- Spatial Distribution of Schools and Transit-Dependent Students at the AEP Level
 
@@
 # References
+
+<!--
+
+  -->

### 2016-02-10 paper.tex
load data

--- 2016-02-09 paper.Rmd
@@
----
-title: 'The unequal distribution of access to education and transport:
-  insights from for São Paulo'
-author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação
-  Getulio Vargas) and Robin Lovelace (University of Leeds)
-output: 
-  pdf_document: 
-    keep_tex: yes
-bibliography: school-access-bib.bib
----
+\documentclass[]{article}
+\usepackage{lmodern}
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \ifxetex
+    \usepackage{mathspec}
+    \usepackage{xltxtra,xunicode}
+  \else
+    \usepackage{fontspec}
+  \fi
+  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
+  \newcommand{\euro}{€}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\usepackage[margin=1in]{geometry}
+\ifxetex
+  \usepackage[setpagesize=false, % page size defined by xetex
+              unicode=false, % unicode breaks when used with xetex
+              xetex]{hyperref}
+\else
+  \usepackage[unicode=true]{hyperref}
+\fi
+\hypersetup{breaklinks=true,
+            bookmarks=true,
+            pdfauthor={Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação Getulio Vargas) and Robin Lovelace (University of Leeds)},
+            pdftitle={The unequal distribution of access to education and transport: insights from for São Paulo},
+            colorlinks=true,
+            citecolor=blue,
+            urlcolor=blue,
+            linkcolor=magenta,
+            pdfborder={0 0 0}}
+\urlstyle{same}  % don't use monospace font for urls
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\setcounter{secnumdepth}{0}
 
-```{r, echo=FALSE}
-#date: '`r format(Sys.time(), ''%F'')`'
-```
+%%% Use protect on footnotes to avoid problems with footnotes in titles
+\let\rmarkdownfootnote\footnote%
+\def\footnote{\protect\rmarkdownfootnote}
 
-# Abstract
+%%% Change title format to be more compact
+\usepackage{titling}
+
+% Create subtitle command for use in maketitle
+\newcommand{\subtitle}[1]{
+  \posttitle{
+    \begin{center}\large#1\end{center}
+    }
+}
+
+\setlength{\droptitle}{-2em}
+  \title{The unequal distribution of access to education and transport: insights
+from for São Paulo}
+  \pretitle{\vspace{\droptitle}\centering\huge}
+  \posttitle{\par}
+  \author{Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos
+(Fundação Getulio Vargas) and Robin Lovelace (University of Leeds)}
+  \preauthor{\centering\large\emph}
+  \postauthor{\par}
+  \date{}
+  \predate{}\postdate{}
+
+\begin{document}
+
+\maketitle
+
+\section{Abstract}\label{abstract}
 
@@
 
-# Introduction
+\section{Introduction}\label{introduction}
 
-**Contents:**
+\textbf{Contents:}
 
-- Motivations
-- General measures of accessibility
-- School accessibility
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Motivations
+\item
+  General measures of accessibility
+\item
+  School accessibility
+\end{itemize}
 
-...
+\ldots{}
 
-Travel to school is an everyday reality for millions of young people around the world.
-The mode, duration, safety, comfort and pollution levels of this trip has huge
-impacts on the future generation, yet has received relatively little academic
-attention.
+Travel to school is an everyday reality for millions of young people
+around the world. The mode, duration, safety, comfort and pollution
+levels of this trip has huge impacts on the future generation, yet has
+received relatively little academic attention.
 
 Travel options are vital for ensuring a more equitable supply of
-educational opportunity to diverse groups. Conversely, poor accessibility
-to deprived area can reinforce social inequalities, with long-term implications.
-Based on this emerging context, this paper develops a school accessibility index
-for local areas.
+educational opportunity to diverse groups. Conversely, poor
+accessibility to deprived area can reinforce social inequalities, with
+long-term implications. Based on this emerging context, this paper
+develops a school accessibility index for local areas.
 
-The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
+The first well-known attempt to define accessibility quantitatively was
+by Ingram (1971), which presented a range of measures related to
+distance (Euclidean and network), barriers and different functions
+representing distance decay.
 
-This early work made the distinction between accessibility indeces that apply to zones or single 'desire lines':
-"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area"
-[@allen1993accesibility].
+This early work made the distinction between accessibility indeces that
+apply to zones or single `desire lines': ``relative accessibility is
+defined as a measure of the effort of overcoming spatial separation
+between two points, while the integral accessibility is defined as a
+measure of the effort of overcoming spatial separation between a point
+and all other points within an area'' (Allen, Liu, and Singer 1993).
 
-# Method: measuring school accessibility
+\section{Method: measuring school
+accessibility}\label{method-measuring-school-accessibility}
 
-We calculate three different accessibility measures for different modes: cumulative, gravitiy-based, and competitive.
+We calculate three different accessibility measures for different modes:
+cumulative, gravitiy-based, and competitive.
 
-The cumulative-opportunity measure for mode $M$ (Boisjoly and El-Geneidy, 2016) is defined as:
+The cumulative-opportunity measure for mode \(M\) (Boisjoly and
+El-Geneidy, 2016) is defined as:
 
-$$ CO_{i}^M= \sum_{j=1}^n(S_{j}f(C_{ij}) $$
+\[ CO_{i}^M= \sum_{j=1}^n(S_{j}f(C_{ij}) \]
 
-$$ f(C_{ij}) = \left\{ 
+\[ f(C_{ij}) = \left\{ 
                 \begin{array}{ll}
@@
               \right.
-  $$
+  \]
 
-This measure counts the number of schools available from one area within a certain travel time by mode $M$ threshold.
-$C_{ij}$ is the travel cost (measured in time) between the centroid of zone i and the centroid of zone j.
+This measure counts the number of schools available from one area within
+a certain travel time by mode \(M\) threshold. \(C_{ij}\) is the travel
+cost (measured in time) between the centroid of zone i and the centroid
+of zone j.
 
@@
 
-In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found that the cumulative-opportunity measure (using a 45 minute threshold) is highly correlated with a gravity-based accessibility measure ([@handy1997measuring], [@hansen1959accessibility]), defined as:
+In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found
+that the cumulative-opportunity measure (using a 45 minute threshold) is
+highly correlated with a gravity-based accessibility measure
+(({\textbf{???}}), ({\textbf{???}})), defined as:
 
-$$ A_{i}^M= \sum_{j=1}^n(S_{j}e^{\beta*C_{ij}}) $$
+\[ A_{i}^M= \sum_{j=1}^n(S_{j}e^{\beta*C_{ij}}) \]
 
-Where $A_{i}^M$ is the gravity accessibility at the centroid of area $i$ to all schools at area $j$ using mode $M$, and $\beta$ is a negative exponential cost function.
-The cost function, based on a negative exponential decay curve, includes reported number of school trips in the 2007 Origin-Destination Survey.
-(Here we can also use information from the school census on number of students per school to know how many students the area attracts).
-The proposed accessibility measure captures the fact that more proximate schools (S) are more attractive than those located further away.
+Where \(A_{i}^M\) is the gravity accessibility at the centroid of area
+\(i\) to all schools at area \(j\) using mode \(M\), and \(\beta\) is a
+negative exponential cost function. The cost function, based on a
+negative exponential decay curve, includes reported number of school
+trips in the 2007 Origin-Destination Survey. (Here we can also use
+information from the school census on number of students per school to
+know how many students the area attracts). The proposed accessibility
+measure captures the fact that more proximate schools (S) are more
+attractive than those located further away.
 
@@
 
-The measures proposed above take into account the sptail distribution of schooling opportunities, but not the local demand for schooling.
-This is particularly relevant for our equity analysis, since it could be the case that higher areas are disproportionally served with respect to the number of potentail students living within a certain travel distance, whereas the opposite holds for lower-income areas.
-In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (the demand), and the sum of individuals within the school grade age-group living in each area (the supply) in the following competitive accessibility measure, first proposed by [@shen1998location] and adapted by El-Geneidy et al (2015):
+The measures proposed above take into account the sptail distribution of
+schooling opportunities, but not the local demand for schooling. This is
+particularly relevant for our equity analysis, since it could be the
+case that higher areas are disproportionally served with respect to the
+number of potentail students living within a certain travel distance,
+whereas the opposite holds for lower-income areas. In order to assess
+the mismatch between the demand and supply for schooling, we use the sum
+of students in schools in each area (the demand), and the sum of
+individuals within the school grade age-group living in each area (the
+supply) in the following competitive accessibility measure, first
+proposed by ({\textbf{???}}) and adapted by El-Geneidy et al (2015):
 
-$$ CA_{i}^M= \sum_{j=1}^n\frac{P_{j}-f(C_{ij})}{\sum_{k=1}^n(Y_{k}-f(C_{kj}))}$$
+\[ CA_{i}^M= \sum_{j=1}^n\frac{P_{j}-f(C_{ij})}{\sum_{k=1}^n(Y_{k}-f(C_{kj}))}\]
 
-Where the numerator discounts the number of pupils in area $j$ $P_{j}$ by how far area $i$ is from area $j$ using the same function as before, and the denominator discounts the number of students (young adults) living in zone $k$ $Y_{k}$ by how far they are from area $j$.
-In this way, the discounted number of study places at each area is divided by the potential students available to fill those places, and then summed in order to obtain a single accessibility measure for each area $i$.
+Where the numerator discounts the number of pupils in area \(j\)
+\(P_{j}\) by how far area \(i\) is from area \(j\) using the same
+function as before, and the denominator discounts the number of students
+(young adults) living in zone \(k\) \(Y_{k}\) by how far they are from
+area \(j\). In this way, the discounted number of study places at each
+area is divided by the potential students available to fill those
+places, and then summed in order to obtain a single accessibility
+measure for each area \(i\).
 
-In order to assess equality in access to schooling, we stratify our population of students in two student groups (we could do it for more income groups): the students living with a head of household who earns 1 minimum wage or less, and those living with a head of household who earns more than 1 minimum wage.
-The accessibility measures are based on student's travel times by different modes using data from the 2007 Origin-Destination survey.
-Alternatively, for the case of public transport, wet use actual transit times in 2015 for comparison (bilhete unico data?).
+In order to assess equality in access to schooling, we stratify our
+population of students in two student groups (we could do it for more
+income groups): the students living with a head of household who earns 1
+minimum wage or less, and those living with a head of household who
+earns more than 1 minimum wage. The accessibility measures are based on
+student's travel times by different modes using data from the 2007
+Origin-Destination survey. Alternatively, for the case of public
+transport, wet use actual transit times in 2015 for comparison (bilhete
+unico data?).
 
-# Results
+\section{Results}\label{results}
 
-- Descriptive statistics about modal choice and travel time by AEPs and level of income
-- Spatial Distribution of Schools and Transit-Dependent Students at the AEP Level
+\begin{itemize}
+\itemsep1pt\parskip0pt\parsep0pt
+\item
+  Descriptive statistics about modal choice and travel time by AEPs and
+  level of income
+\item
+  Spatial Distribution of Schools and Transit-Dependent Students at the
+  AEP Level
+\end{itemize}
 
-# Discussion
+\section{Discussion}\label{discussion}
 
-# Conclusion
+\section{Conclusion}\label{conclusion}
 
-# References
+\section{References}\label{references}
 
-<!--
+Allen, W Bruce, Dong Liu, and Scott Singer. 1993. ``Accesibility
+Measures of US Metropolitan Areas.'' \emph{Transportation Research Part
+B: Methodological} 27 (6). Elsevier: 439--49.
 
-  -->
+Ingram, David R. 1971. ``The Concept of Accessibility: A Search for an
+Operational Form.'' \emph{Regional Studies} 5 (2). Taylor \& Francis:
+101--7.
+
+\end{document}

### 2016-02-10 R/load_census_sc_2010.R

@@ -1,0 +1,24 @@
+# Aim: Read 2010 Census data at the setor censitario level
+
+library(raster)
+library(foreign)
+library(dplyr)
+
+sc_zone <- shapefile("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/Shapefiles SPMR 2010/RM_Sao_Paulo_sc2010_CEM/SC2010_CEM_RMSAO.shp")
+
+data<-read.dbf("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/Shapefiles SPMR 2010/RM_Sao_Paulo_sc2010_CEM/SC2010_CEM_RMSAO_p2.dbf")
+
+keep<-c("CODSETTX", "P11_040" , "P11_041",  "P11_042", "P11_043", "P11_044", "P11_045", "P11_046", "P11_047", "P11_048", "P11_049", "P11_050", "P11_051", "P11_052", "P11_053", "P11_054", "P12_040" , "P12_041",  "P12_042", "P12_043", "P12_044", "P12_045", "P12_046", "P12_047", "P12_048", "P12_049", "P12_050", "P12_051", "P12_052", "P12_053", "P12_054")
+
+data<-data[colnames(data) %in% keep]
+colnames(data)[1]<-"CODSETOR"
+
+sc_zone@data<-left_join(sc_zone@data, data, by = "CODSETOR")
+
+sc_zone@data[is.na(sc_zone@data)]<-0
+
+# Number of individuals aged 6-14 (fundamental school age range) in 2008 (equivalent to individuals 8-16 in 2010)
+sc_zone@data$fund_2008 <-sc_zone@data$P11_040 + sc_zone@data$P11_041 + sc_zone@data$P11_042 + sc_zone@data$P11_043 + sc_zone@data$P11_044 + sc_zone@data$P11_045 + sc_zone@data$P11_046 + sc_zone@data$P11_047 + sc_zone@data$P11_048 + sc_zone@data$P12_040 + sc_zone@data$P12_041 + sc_zone@data$P12_042 + sc_zone@data$P12_043 + sc_zone@data$P12_044 + sc_zone@data$P12_045 + sc_zone@data$P12_046 + sc_zone@data$P12_047 + sc_zone@data$P12_048 
+
+# Number of individuals aged 15-18 (fundamental school age range) in 2008 (equivalent to individuals 17-20 in 2010)
[…]

### 2016-02-10 R/load_school_census.R

@@ -1,0 +1,5 @@
+# Load micro-data Censos Escolar 2000
+
+library(SAScii)
+
+escolas2000<-read.SAScii( "C:/Users/juanita82/Dropbox/Escolas/micro_censo_escolar2000/DADOS_CENSOESC/DADOS_CENSOESC.txt", "C:/Users/juanita82/Dropbox/Escolas/micro_censo_escolar2000/Import/INPUT_SAS_CENSOESC.sas", beginline = 6, buffersize = 50, zipped = F , n = -1 , intervals.to.print = 10 , lrecl = NULL , skip.decimal.division = NULL )

### 2016-02-10 R/agg_flows.R
Add aggregation script - subsetting flows

@@ -1,0 +1,22 @@
+# Aggregate trips
+# Aim: find the number of students by mode travelling on each home:school route in the data
+
+# od_flows with source("load_OD_2007")
+
+# stage 1: identify unique lines
+nrow(od_flow)
+sel_u <- !duplicated(od_flow[c("CO_O_X", "CO_O_Y")])
+od_flowu <- od_flow[sel_u,]
+cents_o <- zones
+cents_d <- schools
+head(zones)
+
+
+library(stplanr)
+od2line()
+
+# stage 2: snap lines to school/residential areas
+
+# stage 3: aggregate flows
+
+# stage 4 convert into lines

### 2016-02-10 R/load_OD_2007.R

@@ -4,7 +4,9 @@
 library(raster)
 
 od_zone <- shapefile("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
+# od_zone <- shapefile("D://data-2-irc/sao-paulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
 od_flow <- read.table("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/OD_2007.txt", sep=",")
+# od_flow <- read.table("D://data-2-irc/sao-paulo/Data/OD-2007/OD_2007.txt", sep=",")
 
 colnames(od_flow) <- c("ZONA",  "MUNI_DOM",	"CO_DOM_X",	"CO_DOM_Y",	"ID_DOM",	"F_DOM",	"FE_DOM",	"DOM",	"CD_ENTRE",	"DATA",	"TIPO_DOM",	"NO_MORAD",	"TOT_FAM",	"ID_FAM",	"F_FAM",	"FE_FAM",	"FAMILIA",	"NO_MORAF",	"CONDMORA",	"QT_RADIO",	"QT_GEL1",	"QT_GEL2",	"QT_TVCOR",	"QT_FREEZ",	"QT_VIDEO",	"QT_BANHO",	"QT_MOTO",	"QT_AUTO",	"QT_ASPIR",	"QT_MLAVA",	"QT_EMPRE",	"QT_MICRO",	"QT_BICICLE",	"NAO_DCL_IT",	"CRITERIO_B",	"ANO_AUTO1",	"ANO_AUTO2",	"ANO_AUTO3",	"RENDA_FA",	"CD_RENFA",	"ID_PESS",	"F_PESS",	"FE_PESS",	"PESSOA",	"SIT_FAM",	"IDADE",	"SEXO",	"ESTUDA",	"GRAU_INS",	"CD_ATIVI",	"CO_REN_I",	"VL_REN_I",	"ZONA_ESC",	"MUNIESC",	"CO_ESC_X",	"CO_ESC_Y",	"TIPO_ESC",	"ZONATRA1",	"MUNITRA1",	"CO_TR1_X",	"CO_TR1_Y",	"TRAB1_RE",	"TRABEXT1",	"OCUP1",	"SETOR1",	"VINC
[…]

### 2016-02-10 R/load_schools_data.R

@@ -4,13 +4,11 @@ library(raster)
 
 # 1 loading data
 file.rename("/tmp/data/escolas_SP_2008/ESCOLAS_DE_2008_GEORREFERENCIADAS.dbf", "/tmp/data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.dbf")
-schools <- shapefile("/tmp/data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.shp")
+schools <- shapefile("D://data-2-irc/sao-paulo/data/escolas_SP_2008/Escolas_de_2008_georreferenciadas.shp")
 proj4string(schools) <- CRS("+init=epsg:4326")
 
-zones <- shapefile("/tmp/data/Shapefiles SPMR 2010/aep10.shp")
+zones <- shapefile("D://data-2-irc/sao-paulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
 zones <- spTransform(zones, proj4string(schools))
-
-od <- 
 
 plot(zones)
 points(schools, col = "red")
@@ -30,3 +28,4 @@ library(leaflet)
 leaflet() %>% addTiles() %>% addCircles(data = schools)
 plot(schools)
 head
+

### 2016-02-10 paper.Rmd
add data section and refs

@@
 - School accessibility
+- Literature review
 
+References transport and social exclusion
+[@hernandez2015mobilities] and [@hernandez2015mobilities] provides a general framework to undertand transport-related social exclusion in emerging economies
+
+Evidence on access inequalities in the Brazilian context [@maia2016access]
+
+Active travel children and adolescents in the SPMR [@de2015changes]
 ...
@@
 
+# Area of study
+
+The The S\~{a}o Paulo Metropolitan Region (SPMR) is a large metropolis located in the South West of Brazil.
+It extends for over 7,700 squared kilometers and groups 39 different minucipalities.
+The estimated population in 2010 was close to 21 million.
+...
+ADD MAP, etc.
+
+#Data
+
+For building our school accessibility index, we use data from three different sources.
+The first one is the 2008 School Census of the Brazilian National Institute of Educational Studies and Surveys (INEP), which provides information on the (universe) of public educational institutions, including their exact location (postal address) and the number of students enrolled in primary and secondary education.
+A clarification about the educational system in Brazil is in place here.
+Primary education \textit{ensino fundamental} comprises nine years, divided on a first year of basic literacy \textit{clase de alfabetiza\c{c}\~{a}}, followed by eight `` series''.
+Primary education is mandatory for children aged 6-14.
+Secondary education lasts three years.
+In principle, the age range of secondary education students is 15-18.
+For each school, have geo-localized each postal adress for a total of XX public primary and XX secondary schools in the SPRM.
+
+Our second source is information the 2010 Population Census of Brazil at the \textit{clase de alfabetiza\c{c}\~{a}}, compiled and freely distributed by the Brazilian Institute of Statistics (IBGE).
+This spatial unit is equivalent to enumeration areas defined for surveying purposes, and contain on average 400 households.
+There is a total of 30,815 areas in the SPMR.
+IBGE also provides the digital networks containing the boundaries of the enumeration areas.
+From the census microdata, we have data on the number of inhabitants by age for each enumeration area.
+As the data was compiled in 2010 and we are interested in the number of children in primary school age (ages 6-14), and adolescents in secondary school age (ages 15-18), we calculate the count of children in the age range 8-16 (who were 6 and 14 in 2008) and adolescents in the age range 17-20 (aged 15-18 in 2008).
+The underlying assumption is that the location of children and adolescents did not change dramatically or systematically between 2008 and 2010, which is a reasonable assumption given thatr residential mobility is relatively low.
+using this procedure, we obtain a total number of children in primary school age of 2,731,999 (approximately 13 percent of the population) and a total of 1,730,252 tenneagers in secondary school age (approximately 8 percent of the population).
+
+Out last source is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the S\~{a}o Paulo Metropolitan Transport Authority M\^{e}tro.
+we have information at the travel flow-level.
+geographic coordinates are provided for the origin and destination of each trip, as well as additional information such as trip purpose (includng education), trip duration, number of transfers and mode.
+The O-D Survey also provides the exact location of schools.
+The survey also provides relevant information on individuals undertaking the trips, such as their age, income range and educational status (including student).
+We focus on trips made by primary and secondary schools with an indicated motive ``education'' at the destination.
+The total number of trips (excluding missing values) was 11,845.
+Of these, 5.5 percent were multimodal.
+In those cases, we assing the mode of the trip leg with the largest duration.
+We add all trip leg durations (in minutes) to obtain the total trip duration.
+
 # Method: measuring school accessibility
@@
 
-The cumulative-opportunity measure for mode $M$ (Boisjoly and El-Geneidy, 2016) is defined as:
+The cumulative-opportunity measure for mode $M$ [@boisjoly2016] is defined as:
 

### 2016-02-10 R/load_census_sc_2010.R

@@ -18,7 +18,7 @@ sc_zone@data<-left_join(sc_zone@data, data, by = "CODSETOR")
 sc_zone@data[is.na(sc_zone@data)]<-0
 
 # Number of individuals aged 6-14 (fundamental school age range) in 2008 (equivalent to individuals 8-16 in 2010)
-sc_zone@data$fund_2008 <-sc_zone@data$P11_040 + sc_zone@data$P11_041 + sc_zone@data$P11_042 + sc_zone@data$P11_043 + sc_zone@data$P11_044 + sc_zone@data$P11_045 + sc_zone@data$P11_046 + sc_zone@data$P11_047 + sc_zone@data$P11_048 + sc_zone@data$P12_040 + sc_zone@data$P12_041 + sc_zone@data$P12_042 + sc_zone@data$P12_043 + sc_zone@data$P12_044 + sc_zone@data$P12_045 + sc_zone@data$P12_046 + sc_zone@data$P12_047 + sc_zone@data$P12_048 
+sc_zone@data$fund_2008 <- sc_zone@data$P11_042 + sc_zone@data$P11_043 + sc_zone@data$P11_044 + sc_zone@data$P11_045 + sc_zone@data$P11_046 + sc_zone@data$P11_047 + sc_zone@data$P11_048 + sc_zone@data$P12_040 + sc_zone@data$P12_041 + sc_zone@data$P12_042 + sc_zone@data$P12_043 + sc_zone@data$P12_044 + sc_zone@data$P12_045 + sc_zone@data$P12_046 + sc_zone@data$P12_047 + sc_zone@data$P12_048 + sc_zone@data$P11_049 + sc_zone@data$P11_050
 
 # Number of individuals aged 15-18 (fundamental school age range) in 2008 (equivalent to individuals 17-20 in 2010)
-sc_zone@data$media_2008 <-sc_zone@data$P11_048 + sc_zone@data$P11_049 + sc_zone@data$P11_050 + sc_zone@data$P11_051 + sc_zone@data$P11_052 + sc_zone@data$P11_053 + sc_zone@data$P11_054 +   sc_zone@data$P12_048 + sc_zone@data$P12_049 + sc_zone@data$P12_050 + sc_zone@data$P12_051 + sc_zone@data$P12_052 + sc_zone@data$P12_053 + sc_zone@data$P12_054 
[…]

### 2016-02-10 paper.Rmd
Update paper + flow aggregation code (work in progress)

@@
 
-The The S\~{a}o Paulo Metropolitan Region (SPMR) is a large metropolis located in the South West of Brazil.
+The São Paulo Metropolitan Region (SPMR) is a large metropolis located in the South West of Brazil.
 It extends for over 7,700 squared kilometers and groups 39 different minucipalities.
@@
 A clarification about the educational system in Brazil is in place here.
-Primary education \textit{ensino fundamental} comprises nine years, divided on a first year of basic literacy \textit{clase de alfabetiza\c{c}\~{a}}, followed by eight `` series''.
+Primary education *ensino fundamental* comprises nine years, divided on a first year of basic literacy *clase de alfabetização*, followed by eight "series".
 Primary education is mandatory for children aged 6-14.
@@
 
-Our second source is information the 2010 Population Census of Brazil at the \textit{clase de alfabetiza\c{c}\~{a}}, compiled and freely distributed by the Brazilian Institute of Statistics (IBGE).
+Our second source is information the 2010 Population Census of Brazil at the *clase de alfabetização*, compiled and freely distributed by the Brazilian Institute of Statistics (IBGE).
 This spatial unit is equivalent to enumeration areas defined for surveying purposes, and contain on average 400 households.
@@
 
-Out last source is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the S\~{a}o Paulo Metropolitan Transport Authority M\^{e}tro.
+Out last source is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority M\^{e}tro.
 we have information at the travel flow-level.

### 2016-02-10 R/agg_flows.R

@@ -1,5 +1,7 @@
 # Aggregate trips
 # Aim: find the number of students by mode travelling on each home:school route in the data
+library(sp)
+library(rgeos)
 
 # od_flows with source("load_OD_2007")
 
@@ -7,9 +9,44 @@
 nrow(od_flow)
 sel_u <- !duplicated(od_flow[c("CO_O_X", "CO_O_Y")])
 od_flowu <- od_flow[sel_u,]
-cents_o <- zones
+# remove ones with na coords
+sel_nacoords <- is.na(od_flowu$CO_O_X) | is.na(od_flowu$CO_O_Y)
+sum(sel_nacoords)
+od_flowu <- od_flowu[!sel_nacoords,]
+
+cents_o <- gCentroid(zones, byid = T)
 cents_d <- schools
-head(zones)
+
+flow_o <- SpatialPoints(cbind(od_flowu$CO_O_X, od_flowu$CO_O_Y))
+flow_d <- SpatialPoints(cbind(od_flowu$CO_D_X, od_flowu$CO_D_Y))
+proj4string(flow_o) <- proj4string(flow_d) <- CRS("+init=epsg:29183")
+flow_o <- spTransform(flow_o, proj4string(schools))
+flow_d <- spTransform(flow_d, proj4string(schools))
+
+# distance to schools
+# sdists <- spDists(flow_o, schools, longlat = T)
+# mindists_o <- apply(sdists, 1, which.min)
+# sdists <- spDists(flow_d, schools, longlat = T)
+# mindists_d <- apply(sdists, 1, which.min)
+# summary(mindists_o) 
+# summary(mindists_d) # no discernable difference - select only destinations near schools
+
+# subset flows that go to the 2201 schools - 4 for edu. performance
+# for route analysis (stats on od_flowu)
+# for analysis of mode choice etc use od_flows
+library(stplanr)
+buff_school <- buff_geo(schools, width = 500)
+plot(buff_school)
+flow_d <- SpatialPointsDataFrame(flow_d, od_flowu)
+flow_s <- flow_d[buff_school,]
+summary(row.names(flow_s))
+od_school <- 
+
+# check the flows overlap
+plot(schools)
+points(flow_o)
+bbox(flow_dest)
+bbox(zones)
 
 
 library(stplanr)

### 2016-02-10 R/load_OD_2007.R

@@ -3,10 +3,10 @@
 
 library(raster)
 
-od_zone <- shapefile("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
-# od_zone <- shapefile("D://data-2-irc/sao-paulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
-od_flow <- read.table("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/OD_2007.txt", sep=",")
-# od_flow <- read.table("D://data-2-irc/sao-paulo/Data/OD-2007/OD_2007.txt", sep=",")
+# od_zone <- shapefile("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
+od_zone <- shapefile("D://data-2-irc/sao-paulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
+# od_flow <- read.table("C:/Users/juanita82/Dropbox/Robin-SaoPaulo/SaoPaulo/Data/OD-2007/OD_2007.txt", sep=",")
+od_flow <- read.table("D://data-2-irc/sao-paulo/Data/OD-2007/OD_2007.txt", sep=",")
 
[…]

### 2016-02-10 R/load_schools_data.R

@@ -10,22 +10,22 @@ proj4string(schools) <- CRS("+init=epsg:4326")
 zones <- shapefile("D://data-2-irc/sao-paulo/Data/OD-2007/Mapas OD/ZONAS_od07.shp")
 zones <- spTransform(zones, proj4string(schools))
 
-plot(zones)
-points(schools, col = "red")
-schools <- schools[zones,] # select only those in zones
-points(schools, col = "green")
-nrow(schools)
-df <- schools@data
-View(df)
-plot(df$ALE_1_4 + df$ALE_5_8, df$TOT_MATR)
-cor(df$ALE_1_4 + df$ALE_5_8, df$TOT_MATR, use = "complete.obs")
-
-# which schools to use
-
-# Visualising data
-library(tmap)
-library(leaflet)
-leaflet() %>% addTiles() %>% addCircles(data = schools)
-plot(schools)
-head
+# plot(zones)
+# points(schools, col = "red")
+# schools <- schools[zones,] # select only those in zones
+# points(schools, col = "green")
+# nrow(schools)
+# df <- schools@data
+# View(df)
+# plot(df$ALE_1_4 + df$ALE_5_8, df$TOT_MATR)
+# cor(df$ALE_1_4 + df$ALE_5_8, df$TOT_MATR, use = "complete.obs")
+# 
+# # which schools to use
+# 
+# # Visualising data
+# library(tmap)
+# library(leaflet)
+# leaflet() %>% addTiles() %>% addCircles(data = schools)
+# plot(schools)
+# head
 

### 2016-02-11 paper.tex

@@
 \urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \setlength{\parindent}{0pt}
@@
 \setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
 \setcounter{secnumdepth}{0}
@@
 
+% Redefines (sub)paragraphs to behave more like sections
+\ifx\paragraph\undefined\else
+\let\oldparagraph\paragraph
+\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+\let\oldsubparagraph\subparagraph
+\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
 \begin{document}
-
 \maketitle
@@
 \begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
+\tightlist
 \item
@@
   School accessibility
+\item
+  Literature review
 \end{itemize}
 
+References transport and social exclusion (Hernandez and Titheridge
+2015) and (Hernandez and Titheridge 2015) provides a general framework
+to undertand transport-related social exclusion in emerging economies
+
+Evidence on access inequalities in the Brazilian context (Maia et al.
+2016)
+
+Active travel children and adolescents in the SPMR (Sá et al. 2015)
 \ldots{}
@@
 
+\section{Area of study}\label{area-of-study}
+
+The São Paulo Metropolitan Region (SPMR) is a large metropolis located
+in the South West of Brazil. It extends for over 7,700 squared
+kilometers and groups 39 different minucipalities. The estimated
+population in 2010 was close to 21 million. \ldots{} ADD MAP, etc.
+
+\section{Data}\label{data}
+
+For building our school accessibility index, we use data from three
+different sources. The first one is the 2008 School Census of the
+Brazilian National Institute of Educational Studies and Surveys (INEP),
+which provides information on the (universe) of public educational
+institutions, including their exact location (postal address) and the
+number of students enrolled in primary and secondary education. A
+clarification about the educational system in Brazil is in place here.
+Primary education \textit{ensino fundamental} comprises nine years,
+divided on a first year of basic literacy
+\textit{clase de alfabetiza\c{c}\~{a}}, followed by eight `` series''.
+Primary education is mandatory for children aged 6-14. Secondary
+education lasts three years. In principle, the age range of secondary
+education students is 15-18. For each school, have geo-localized each
+postal adress for a total of XX public primary and XX secondary schools
+in the SPRM.
+
+Our second source is information the 2010 Population Census of Brazil at
+the \textit{clase de alfabetiza\c{c}\~{a}}, compiled and freely
+distributed by the Brazilian Institute of Statistics (IBGE). This
+spatial unit is equivalent to enumeration areas defined for surveying
+purposes, and contain on average 400 households. There is a total of
+30,815 areas in the SPMR. IBGE also provides the digital networks
+containing the boundaries of the enumeration areas. From the census
+microdata, we have data on the number of inhabitants by age for each
+enumeration area. As the data was compiled in 2010 and we are interested
+in the number of children in primary school age (ages 6-14), and
+adolescents in secondary school age (ages 15-18), we calculate the count
+of children in the age range 8-16 (who were 6 and 14 in 2008) and
+adolescents in the age range 17-20 (aged 15-18 in 2008). The underlying
+assumption is that the location of children and adolescents did not
+change dramatically or systematically between 2008 and 2010, which is a
+reasonable assumption given thatr residential mobility is relatively
+low. using this procedure, we obtain a total number of children in
+primary school age of 2,731,999 (approximately 13 percent of the
+population) and a total of 1,730,252 tenneagers in secondary school age
+(approximately 8 percent of the population).
+
+Out last source is the 2007 Origin-Destination Household Travel Survey
+(O-D Survey), carried out by the São Paulo Metropolitan Transport
+Authority M\^{}\{e\}tro. we have information at the travel flow-level.
+geographic coordinates are provided for the origin and destination of
+each trip, as well as additional information such as trip purpose
+(includng education), trip duration, number of transfers and mode. The
+O-D Survey also provides the exact location of schools. The survey also
+provides relevant information on individuals undertaking the trips, such
+as their age, income range and educational status (including student).
+We focus on trips made by primary and secondary schools with an
+indicated motive ``education'' at the destination. The total number of
+trips (excluding missing values) was 11,845. Of these, 5.5 percent were
+multimodal. In those cases, we assing the mode of the trip leg with the
+largest duration. We add all trip leg durations (in minutes) to obtain
+the total trip duration.
+
 \section{Method: measuring school
@@
 
-The cumulative-opportunity measure for mode \(M\) (Boisjoly and
-El-Geneidy, 2016) is defined as:
+The cumulative-opportunity measure for mode \(M\) (Boisjoly G. 2016) is
+defined as:
 
@@
 
+\hyperdef{}{ref-allen1993accesibility}{\label{ref-allen1993accesibility}}
 Allen, W Bruce, Dong Liu, and Scott Singer. 1993. ``Accesibility
@@
 
+\hyperdef{}{ref-boisjoly2016}{\label{ref-boisjoly2016}}
+Boisjoly G., El-Geneidy A. 2016. ``Is Complexity Better? Assessing
+Accessibility Measures That Account for Daily Fluctuations in Transit
+and Jobs Availability.'' 95th Annual Meeting of the Transportation
+Research Board, Washington D.C., USA.
+
+\hyperdef{}{ref-hernandez2015mobilities}{\label{ref-hernandez2015mobilities}}
+Hernandez, Daniel Oviedo, and Helena Titheridge. 2015. ``Mobilities of
+the Periphery: Informality, Access and Social Exclusion in the Urban
+Fringe in Colombia.'' \emph{Journal of Transport Geography}. Elsevier.
+
+\hyperdef{}{ref-ingram1971concept}{\label{ref-ingram1971concept}}
 Ingram, David R. 1971. ``The Concept of Accessibility: A Search for an
@@
 
+\hyperdef{}{ref-maia2016access}{\label{ref-maia2016access}}
+Maia, Maria Leonor, Karen Lucas, Geraldo Marinho, Enilson Santos, and
+Jessica Helena de Lima. 2016. ``Access to the Brazilian City---From the
+Perspectives of Low-Income Residents in Recife.'' \emph{Journal of
+Transport Geography}. Elsevier.
+
+\hyperdef{}{ref-de2015changes}{\label{ref-de2015changes}}
+Sá, Thiago Hérick de, Leandro Martin Totaro Garcia, Grégore Iven Mielke,
+Fabiana Maluf Rabacow, and Leandro Fórnias Machado de Rezende. 2015.
+``Changes in Travel to School Patterns Among Children and Adolescents in
+the São Paulo Metropolitan Area, Brazil, 1997--2007.'' \emph{Journal of
+Transport \& Health} 2 (2). Elsevier: 143--50.
+
 \end{document}

### 2016-02-11 paper.Rmd
Update references (all included now)

@@
 
-In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found that the cumulative-opportunity measure (using a 45 minute threshold) is highly correlated with a gravity-based accessibility measure ([@handy1997measuring], [@hansen1959accessibility]), defined as:
+In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found that the cumulative-opportunity measure (using a 45 minute threshold) is highly correlated with a gravity-based accessibility measure ([@handy_measuring_1997], [@hansen1959accessibility]), defined as:
 

### 2016-05-29 R/routing.R
Update distance matrix code

@@ -1,12 +1,17 @@
 # Aim: generate routes between ODs
-pkgs = c("tmap", "rgdal", "rgeos", "mapview", "stplanr", "geosphere", "dplyr")
+pkgs = c("sp", "tmap", "rgdal", "rgeos", "stplanr", "geosphere", "dplyr")
 lapply(pkgs, library, character.only = T)
-cents = read_shape("Data/AEP_2010_codes/aep10.shp")
+cents = read_shape("Data/aep10.shp")
+# cents = gCentroid(cents, byid = T)
 cents = spTransform(x = cents, CRSobj = CRS("+init=epsg:4326"))
 qtm(cents)
 mapview(cents)
 
 f = read.csv("Data/centroids_60min.csv")
+
+odf = points2odf(cents)
+brtime = strptime("2016-05-26 09:00:00", format = "%Y-%m-%d %H:%M:%S", tz = "BRT")
+
 
 l = od2line(flow = f, zones = cents)
 mapview(l)
@@ -31,3 +36,33 @@ r_out = cbind(l@data[1:20,], ldf[1:20,], rc@data[1:20,], rf@data)
 names(r_out)[c(9:10, 12:13)] = c("time_car", "dist_car", "time_foot", "dist_foot")
 summary(r_out)
 write.csv(r_out, "output-data/sample-time-od-out.csv")
+
+# The google dist matrix api way
+
+l@data = cbind(l@data, dists)
+class(l$from_addresses) 
+l@data[4:7] = NA
+lapply(l@data, class)
+class(l$from_addresses) = "character"
+class(l$to_addresses) = "character"
+class(l$distances) = "numeric"
+class(l$duration) = "numeric"
+
+# for(i in 1:nrow(l)){
+for(i in 1:100){
+  flag <- TRUE
+  tryCatch({
+    dists = dist_google(line2points(l[i,])[1,], line2points(l[i,])[2,],
+                        mode = "transit", arrival_time = brtime)
+    l$from_addresses[i] = as.character(dists$from_addresses)
+    l$to_addresses[i] = as.character(dists$to_addresses)
+    l$distances[i] = dists$distances
+    l$duration[i] = dists$duration
+  },
+    error=function(e) flag<<-FALSE
+  )
+  if (!flag) next
+
+}
+ldff = l@data
+write.csv(ldff, "output-data/sample-time-od-out.csv")

### 2016-05-31 paper.Rmd
Update quote marks to markdown style
Updates to paper

--- 2016-02-10 paper.Rmd
@@
 title: 'The unequal distribution of access to education and transport:
-  insights from for São Paulo'
+  insights from the São Paulo Metropolitan Region'
 author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação
@@
 and economic inequalities are embedded through unfair education and
-transport systems. Good schools are mostly concentrated in wealthy
-areas, while transport links to school are deficient in deprived areas,
-exacerbating the issue. Inequalities in educational and transport
+transport systems. Inequalities in educational and transport
 infrastructure are mutually reinforcing: the right to mobility is
@@
 contested. Our index combines information on the spatial distribution of
-children and adolescents by income categories, the location of existing
-schools, the travel patterns of students, and the travel infrastructure
-serving the school catchment area into a single measure. The index is
-used to measure school accessibility locally across São Paulo, using
+adolescents, the location of existing schools, the travel patterns of students, and the travel infrastructure serving the school catchment area into a single measure.
+The index is used to measure school accessibility locally across São Paulo, using
 data sources from Population and School Censuses, commonly available in
@@
 
-**Contents:**
+Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the São Paulo Metropolitan Region had gained a public transport subsidy in 2015.
+However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016.
+The change in policy was estimated to affect over 300,000 students who would be placed in schools far from their homes.
+As a response, in the following months students occupied over 200 schools.
+Protesters scored some victories: the school closings were postponed for one year, and the education secretary has resigned (The New York Times, 2015).
 
-- Motivations
-- General measures of accessibility
-- School accessibility
-- Literature review
+Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas.
+In this paper, we propose a new school accessibility index that combines information of the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure.
+We calculate a school accessibility index for 633 areas within the SPMR and use it to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility.
+We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.
+
+The proposed index is inspired by Shen's index, developed in the context of job accessibility.
+The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account bith the 'supply' and 'demand' of students in each area.
+Concretly, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available.
+As areas are part of a city, each local area is also subject to the inflow of potential students from other areas, as well of the outflow of students to other areas.
+The magnitude of the net flow will depend on the travel distance between all other areas and the area in question, which here we consider to be the public transport commuting time between areas.
+The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unewually across space.
+
+--Lit review
+
+The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
+This early work made the distinction between accessibility indeces that apply to zones or single 'desire lines':
+"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility].
+
+Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education.
+This is manifested by the overlap between recent protests against unwanted changes to public education and the social movements contesting increases in public transport fares.
+Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups.
+Conversely, poor accessibility to deprived area can reinforce social inequalities, with long-term implications.
+Based on this emerging context, this paper develops a school accessibility index for local areas.
 
 References transport and social exclusion
-[@hernandez2015mobilities] and [@hernandez2015mobilities] provides a general framework to undertand transport-related social exclusion in emerging economies
+[@jones2012] provides a general framework to understand transport-related social exclusion in emerging economies
 
-Evidence on access inequalities in the Brazilian context [@maia2016access]
+Evidence on access inequalities in the Brazilian context [@gomide2006]
 
-Active travel children and adolescents in the SPMR [@de2015changes]
+Active travel children and adolescents in the SPMR [@de2015]
+
+Previous literature for consolidated has found that school accessibility has a significant and positive impact on gradutation from upper secondary schools [@falch2013], and on post-compulsory education enrolment [@dickerson2012].
+For emerging economies, [@asahi2014] finds that improved metro accessibility had a negative impact on student's performance as measured by test scores.
 ...
-
-Travel to school is an everyday reality for millions of young people around the world.
-The mode, duration, safety, comfort and pollution levels of this trip has huge
-impacts on the future generation, yet has received relatively little academic
-attention.
-
-Travel options are vital for ensuring a more equitable supply of
-educational opportunity to diverse groups. Conversely, poor accessibility
-to deprived area can reinforce social inequalities, with long-term implications.
-Based on this emerging context, this paper develops a school accessibility index
-for local areas.
-
-The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
-
-This early work made the distinction between accessibility indeces that apply to zones or single 'desire lines':
-"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area"
-[@allen1993accesibility].
 
@@
 The São Paulo Metropolitan Region (SPMR) is a large metropolis located in the South West of Brazil.
-It extends for over 7,700 squared kilometers and groups 39 different minucipalities.
-The estimated population in 2010 was close to 21 million.
-...
-ADD MAP, etc.
+It extends for over 7,700 squared kilometers and groups 38 different minucipalities besides the municipality of São Paulo itself.
+The estimated population in 2010 was close to 19.5 million in 2010.
+The integrated transport system, to which users can access through an electronic card (Bilhete Único) is composed of a railway and bus network.
+The railways network comprises 78.4 kilometers in five subway lines, which are complemented by six suburban railway lines.
+Some of the bus network operates on dedicated bus lines, which amounted to 500 km in 2015.
+As can be seen in Figure
+1, the richer areas of the SPMR are better served by public transport, while an important number of lower income areas is underserved.
+
+```{r inc-trans, echo=FALSE, label="inc-trans"}
+knitr::include_graphics("income_pc.png")
+```
+
+<!-- ![Public transport system and income per capita, 2010](income_pc.png) -->
+
+Regarding the educational system, since the constitution of 1946, primary education is compulsory.
+In 2005, the legislation changed to increase the duration of the primary education, going from 8 to 9 years.
+In 2010, all regions finished the transition.
+In 2009, a constitutional amendment imposed compulsory education for the primary “fundamental” and secondary “médio” levels.
+The constitution imposes that this resolution should be 100% implemented by 2016.
+
+By law, the state/municipalities should guaranty the adequate provision (100% coverage of the demand).
+There are national and state educational funds that are distributed among the schools  according to the number of enrolled students and  existence of facilities (library , laboratories).
+
+The strucuture of the educational system is composed by both public and private schools.
+The public schools are divided in the folllowing levels: Pre school – 4-5 years (municipal administration); Fundamental – 6 – 14 (municipal administration); Secondary “medio” – 15-17 (state administration) and Superior (state and federal administration).
+
+According to the Censo Escolar 2010, there are 10,251 schools in the SPMR, out of which      54% are public (22% state / 32% municipal) and 46% are private.
+Geographically, the student is assigned to the closer school or to their house or their parent’s working place.
+In the secondary, the student can ask for a vacancy in a different school with a justification and under the condition of the availability of vacancy.
+
+The municipality is responsible for the organization of the pre-school and fundamental level.
+Poorer municipalities in SPMR have, in general, lower quality in terms of teachers and facilities.
+All the secondary schools are administrated by the state, nevertheless, differences exists depending of the location in the SPMR.
+In general, schools in peripheral locations show lower quality scores.
 
@@
 For building our school accessibility index, we use data from three different sources.
-The first one is the 2008 School Census of the Brazilian National Institute of Educational Studies and Surveys (INEP), which provides information on the (universe) of public educational institutions, including their exact location (postal address) and the number of students enrolled in primary and secondary education.
+The first one is the 2008 School Census of the Brazilian National Institute of Educational Studies and Surveys (INEP), which provides information on the (universe) of public educational institutions, including a unique identifier and the number of students enrolled in secondary education.
+We use the coordinates contained in the database provided by the CEM (Centro de Estudos da Metrópole) for 2001 to geolocate the schools in 2008 using the unique identifier (FRED TO EXPLAIN HOW WE GEO-LOCATED THE REST).
 A clarification about the educational system in Brazil is in place here.
-Primary education *ensino fundamental* comprises nine years, divided on a first year of basic literacy *clase de alfabetização*, followed by eight "series".
+Primary education \textit{ensino fundamental} comprises nine years, divided on a first year of basic literacy \textit{clase de alfabetização}, followed by eight `` series''.
 Primary education is mandatory for children aged 6-14.
@@
 
-Our second source is information the 2010 Population Census of Brazil at the *clase de alfabetização*, compiled and freely distributed by the Brazilian Institute of Statistics (IBGE).
+Our second source is information the 2010 Population Census of Brazil at the \textit{setor censitario} level, compiled and freely distributed by the Brazilian Institute of Statistics (IBGE).
 This spatial unit is equivalent to enumeration areas defined for surveying purposes, and contain on average 400 households.
@@
 From the census microdata, we have data on the number of inhabitants by age for each enumeration area.
-As the data was compiled in 2010 and we are interested in the number of children in primary school age (ages 6-14), and adolescents in secondary school age (ages 15-18), we calculate the count of children in the age range 8-16 (who were 6 and 14 in 2008) and adolescents in the age range 17-20 (aged 15-18 in 2008).
-The underlying assumption is that the location of children and adolescents did not change dramatically or systematically between 2008 and 2010, which is a reasonable assumption given thatr residential mobility is relatively low.
-using this procedure, we obtain a total number of children in primary school age of 2,731,999 (approximately 13 percent of the population) and a total of 1,730,252 tenneagers in secondary school age (approximately 8 percent of the population).
+The number of adolescents in secondary school age (ages 15-18) is 1'216.611 (approximately 13 percent of the population).
 
-Out last source is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority M\^{e}tro.
-we have information at the travel flow-level.
+Our last source is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority Mêtro.
+We have information at the travel flow-level.
 geographic coordinates are provided for the origin and destination of each trip, as well as additional information such as trip purpose (includng education), trip duration, number of transfers and mode.
@@
 The survey also provides relevant information on individuals undertaking the trips, such as their age, income range and educational status (including student).
-We focus on trips made by primary and secondary schools with an indicated motive ``education'' at the destination.
+We focus on trips made by secondary school students with an indicated motive ``education'' at the destination.
 The total number of trips (excluding missing values) was 11,845.
@@
 
+The Google Distance Matrix API was used to find the travel times and distances by public transport.
+This was implemented using an interface to the API, using the R package **stplanr** [@stplanr].
+To provide an example of how the method works, a reproducible example is illustrated below.
+
+```{r, message=FALSE, eval=FALSE}
+library(stplanr)
+o = c(-46.31813, -23.38584) # coordinates of origin
+d = c(-46.34078, -23.37746) # coordinates of origin
+result = dist_google(from = o, to = d, mode = "transit")
+```
+
+```{r, echo=FALSE}
+# saveRDS(result, "result.Rds")
+result = readRDS("result.Rds")
+```
+
+```{r}
+result[3:4]
+```
+
+The code above takes an origin (`o`) and destination (`d`) and finds the travel time and distance.
+This is reported as `r result[3]` metres and `r result[4]` seconds (`r round(result[4] / 60)` minutes) respectively.
+To ensure that the result was relevant to school travel, the times were calculated for arrival time at 9am on a weekday.
+The function was repeated for all origin-destinatio pairs.
+
+# Preliminary facts
+
+As can be seen by comparing Figures 1, 2 and 3, students disproportionally attend private secondary schools in higher income areas, while the opposite is true for public secondary schools.
+
+![Private secondary schools students home location](med_pr.png)
+![Public secondary schools students home location](med_pu.png)
+
+Enrollment in public secondary schools is more dispersed spatially, as in principle public shcools are not clearly geographically concentrated within the region.
+There is, however, an important degree of variation in the school seats in each sub-area, ranging from zero to 7,754.
+
+![Enrollment in public secondary schools](enrol.png)
+
+Table 1 shows the model split by type of secondary school student.
+Clearly, secondary school students attending private schools disproportionally travel by car, compared to public school students (9% versus 47%), even though higher income areas are better served by public transport.
+This is in line with the findings of [@de2015].
+Around 67% of public school students commute to school by active travel (mostly walking, since the percentage of biking is still relatively small), while 23% commute by public transport, which has much longer mean duration than other modes.
+
+```{r table-placeholder, echo=FALSE}
+
+```
+
+Although the mean duration of travel to school remained stable over the period 1997-2007 ([@de2015]), they could increase as a result of several factors relevant to the SPMR, including sub-urbanization of particular demographic groups, such as those with higher income; increase in the size of schools leading to an increase in school's catchment area; less than proportional increase in schools in high population growth areas with poor connectivity; and institutional changes towards more flexible school choice [@easton2015].
+
 # Method: measuring school accessibility
 
-We calculate three different accessibility measures for different modes: cumulative, gravitiy-based, and competitive.
+To define school accessibility, we start by defining the broad concept of accessibility, as used in the transport literature.
+According to (@geurs2004, p.
+128) accessibility is “the extent to which land-use and transport systems enable (groups of) individuals to reach activities or destinations by means of a (combination of) transport mode(s).” Extended to our case, school accessibility then measures the extent to which the existing built-environment and transport facilities enable children and adolescents in school age to reach schools.
 
-The cumulative-opportunity measure for mode $M$ [@boisjoly2016] is defined as:
+Ideally, a school accessibility measure would include information on the (almost exact) home location of every student and the location of every school they could potentially attend (as in [@andersson2012]), and some measure of commuting costs for any chosen mode (in terms of time and money).
+Given that this level of detail is not available in our data, we rely on area aggregates, implicitly assuming that the measure at the centroid of the area applies homogeneously within the area.
+A previous work studying the impact of metro expansions on test scores for Santiago de Chile measures school accessibility as the distance of every existing school to the nearest subway station [@asahi2014].
+This measure implicitly assumes that most or all students commute to school by subway, which is unlike to hold for the case of the SPMR.
+Furthermore, this measure does not take into account the uneven distribution of students of different ages within the city.
+
+Given these limitations, we first consider a cumulative-opportunity measure for mode $M$ [@boisjoly2016] is defined as:
 
@@
                 \begin{array}{ll}
-                  1 if C_{ij}<=t\\
-                  0 if C_{ij}<t
+                  1\quad if\quad C_{ij}<=t\\
+                  0\quad if\quad C_{ij}<t
                 \end{array}
@@
 
-This measure counts the number of schools available from one area within a certain travel time by mode $M$ threshold.
-$C_{ij}$ is the travel cost (measured in time) between the centroid of zone i and the centroid of zone j.
+This measure counts the number of schools available from one area within a certain travel time threshold by mode $M$.
+$C_{ij}$ is the travel cost (measured in time) between the centroid of zone $i$ and the centroid of zone $j$, and $f(C_{ij})$ is a weight function.
 
-Gravity-based accessibility measure
+This measure takes into account the sptail distribution of schooling opportunities, but not the local demand for schooling.
+This is particularly relevant for our analysis, since it could be the case that some areas are disproportionally served with respect to the number of potentail students living within a certain travel distance.
+In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (supply), and the sum of individuals within the school grade age-group living in each area (demand) in the following competitive accessibility measure, first proposed by [@shen1998location]:
 
-In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found that the cumulative-opportunity measure (using a 45 minute threshold) is highly correlated with a gravity-based accessibility measure ([@handy1997measuring], [@hansen1959accessibility]), defined as:
+$$ CA_{i}^M= \sum_{j=1}^n\frac{O_{j}f(C_{ij})}{\sum_{k=1}^n(t_{k}f(C_{kj}))}$$
 
-$$ A_{i}^M= \sum_{j=1}^n(S_{j}e^{\beta*C_{ij}}) $$
-
-Where $A_{i}^M$ is the gravity accessibility at the centroid of area $i$ to all schools at area $j$ using mode $M$, and $\beta$ is a negative exponential cost function.
-The cost function, based on a negative exponential decay curve, includes reported number of school trips in the 2007 Origin-Destination Survey.
-(Here we can also use information from the school census on number of students per school to know how many students the area attracts).
-The proposed accessibility measure captures the fact that more proximate schools (S) are more attractive than those located further away.
-
-Competitive accessibility measure
-
-The measures proposed above take into account the sptail distribution of schooling opportunities, but not the local demand for schooling.
-This is particularly relevant for our equity analysis, since it could be the case that higher areas are disproportionally served with respect to the number of potentail students living within a certain travel distance, whereas the opposite holds for lower-income areas.
-In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (the demand), and the sum of individuals within the school grade age-group living in each area (the supply) in the following competitive accessibility measure, first proposed by [@shen1998location] and adapted by El-Geneidy et al (2015):
-
-$$ CA_{i}^M= \sum_{j=1}^n\frac{P_{j}-f(C_{ij})}{\sum_{k=1}^n(Y_{k}-f(C_{kj}))}$$
-
-Where the numerator discounts the number of pupils in area $j$ $P_{j}$ by how far area $i$ is from area $j$ using the same function as before, and the denominator discounts the number of students (young adults) living in zone $k$ $Y_{k}$ by how far they are from area $j$.
-In this way, the discounted number of study places at each area is divided by the potential students available to fill those places, and then summed in order to obtain a single accessibility measure for each area $i$.
-
-In order to assess equality in access to schooling, we stratify our population of students in two student groups (we could do it for more income groups): the students living with a head of household who earns 1 minimum wage or less, and those living with a head of household who earns more than 1 minimum wage.
-The accessibility measures are based on student's travel times by different modes using data from the 2007 Origin-Destination survey.
-Alternatively, for the case of public transport, wet use actual transit times in 2015 for comparison (bilhete unico data?).
+Where the numerator discounts the number of "opportunities" (school seats) in area $j$ $P_{j}$ by how far area $i$ is from area $j$ using the same function as before, and the denominator discounts the number of teens in secondary school age living in zone $k$ $T_{k}$ by how far they are from area $j$.
+In other words, the numerator counts how many schools seats can be reached from an area, while the denominator counts how many teens can potentially reach the same area.
+In this way, the discounted number of school seats places at each area is divided by the potential students available to fill those places, and then summed in order to obtain a single accessibility measure for each area.
 
@@
 
-- Descriptive statistics about modal choice and travel time by AEPs and level of income
-- Spatial Distribution of Schools and Transit-Dependent Students at the AEP Level
+INLCUDE MAP COMPETITIVE ACCESIBILITY INDEX BY PUBLIC TRANSPORT, INTERPRETATION OF VALUES, SENSITIVITY ANALYSIS
+
+POLICY EXPERIMENT 
+To simulate the effects of the proposed educational reform on accessibility, assign
+students in areas with lower than average enrollment to the next area with higher
+than average enrollment
+INCLUDE TABLE MEAN ACCESSIBILITY BEFORE AND AFTER REDISTRIBUTION, INTERPRETATION
 

### 2016-05-31 R/routing.R

@@ -47,22 +47,53 @@ class(l$from_addresses) = "character"
 class(l$to_addresses) = "character"
 class(l$distances) = "numeric"
 class(l$duration) = "numeric"
+l$fare = NA
 
 # for(i in 1:nrow(l)){
-for(i in 1:100){
+for(i in 12001:nrow(l)){
   flag <- TRUE
   tryCatch({
     dists = dist_google(line2points(l[i,])[1,], line2points(l[i,])[2,],
-                        mode = "transit", arrival_time = brtime)
+                        mode = "transit", arrival_time = brtime,
+                        google_api = Sys.getenv("GOOGLEDIST"))
     l$from_addresses[i] = as.character(dists$from_addresses)
     l$to_addresses[i] = as.character(dists$to_addresses)
     l$distances[i] = dists$distances
     l$duration[i] = dists$duration
+    l$fare[i] = dists$fare
   },
     error=function(e) flag<<-FALSE
   )
   if (!flag) next
-
 }
+
 ldff = l@data
+l$dist_hav_km = l$dist_hav / 1000
+
+# plotting
+l$dist_band = cut(l$dist_hav_km, breaks = quantile(l$dist_hav_km))
+l$color = cut(l$dist_hav_km, breaks = quantile(l$dist_hav_km),
+              labels = c("green", "blue", "yellow", "red"))
+l$color = as.character(l$color)
+plot(ldff$dist_hav / 1000, ldff$distances / 1000,
+    ylab = "Public transport distance (km)",
+    xlab = "Euclidean distance (km)", col = l$color)
+
+
+
+library(tmap)
+tmap_mode("view")
+osm_tiles = tmap::read_osm(l)
+tm_shape(osm_tiles) +
+  tm_raster() +
+  tm_shape(l) +
+  tm_lines(col = "dist_band")
+
+
+
+
+
+
+qtm(l, line.col = "dist_hav_km")
+
 write.csv(ldff, "output-data/sample-time-od-out.csv")

### 2016-05-31 paper.Rmd
Update paper, no dependency on .Rds file

@@
 ---
-title: 'The unequal distribution of access to education and transport:
-  insights from the São Paulo Metropolitan Region'
+title: 'The unequal distribution of access to education and transport: insights from
+  the São Paulo Metropolitan Region'
 author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação
@@
 output:
+  html_document:
+    keep_md: yes
   pdf_document:
@@
 
-Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the São Paulo Metropolitan Region had gained a public transport subsidy in 2015.
+Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the São Paulo Metropolitan Region gained access to a public transport subsidy in 2015.
 However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016.
-The change in policy was estimated to affect over 300,000 students who would be placed in schools far from their homes.
+The change in policy was estimated to affect over 300,000 students, many whom would be placed in schools far from their homes.
 As a response, in the following months students occupied over 200 schools.
-Protesters scored some victories: the school closings were postponed for one year, and the education secretary has resigned (The New York Times, 2015).
+Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (The New York Times, 2015).
 
@@
 We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.
+(INCLUDE SOMETHING ABOUT CONTESTATION)
 
-The proposed index is inspired by Shen's index, developed in the context of job accessibility.
+The new index we propose here constitutes a methodological contribution to the measurement of  school accessibility.
+It is inspired by the index developed by Shen (1998) in the context of job accessibility.
 The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account bith the 'supply' and 'demand' of students in each area.
@@
 
+HERE REVIEW PREVIOUS SCHOOL ACCESSIBILITY MEASURES AND EXPLAIN THEIR LIMITATIONS VIS-A-VIS OUR INDEX
 --Lit review
@@
 The integrated transport system, to which users can access through an electronic card (Bilhete Único) is composed of a railway and bus network.
-The railways network comprises 78.4 kilometers in five subway lines, which are complemented by six suburban railway lines.
+The railway network comprises 78.4 kilometers in five subway lines and six suburban railway lines which provide less frequent and slower service than the subway.
 Some of the bus network operates on dedicated bus lines, which amounted to 500 km in 2015.
-As can be seen in Figure
-1, the richer areas of the SPMR are better served by public transport, while an important number of lower income areas is underserved.
+As can be seen in Figure 1, the richer areas of the SPMR are better served by public transport, while n important number of lower income areas is underserved.
 
-```{r inc-trans, echo=FALSE, label="inc-trans"}
+```{r inc-trans, echo=FALSE, out.width="80%"}
 knitr::include_graphics("income_pc.png")
 ```
-
-<!-- ![Public transport system and income per capita, 2010](income_pc.png) -->
 
@@
 
-```{r, echo=FALSE}
+```{r, echo=FALSE, eval=FALSE}
 # saveRDS(result, "result.Rds")
@@
 
-```{r}
-result[3:4]
+```
+##   distances duration
+## 1      2962     2118
 ```
@@
 The code above takes an origin (`o`) and destination (`d`) and finds the travel time and distance.
-This is reported as `r result[3]` metres and `r result[4]` seconds (`r round(result[4] / 60)` minutes) respectively.
+This is reported as 2962 metres and 2118 seconds (35 minutes) respectively.
 To ensure that the result was relevant to school travel, the times were calculated for arrival time at 9am on a weekday.
@@
 than average enrollment
+
 INCLUDE TABLE MEAN ACCESSIBILITY BEFORE AND AFTER REDISTRIBUTION, INTERPRETATION

### 2016-05-31 paper.Rmd
Fix typos

@@
 output:
+  word_document: default
+  pdf_document:
+    keep_tex: yes
   html_document:
     keep_md: yes
-  pdf_document:
-    keep_tex: yes
 bibliography: school-access-bib.bib
@@
 inequality nexus with reference to a new school accessibility measure
-applied São Paulo. By capturing both the unequal distribution of schools
+applied to São Paulo. By capturing both the unequal distribution of schools
 and transport services across space, the index allows embedded
@@
 It is inspired by the index developed by Shen (1998) in the context of job accessibility.
-The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account bith the 'supply' and 'demand' of students in each area.
-Concretly, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available.
+The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' of students in each area.
+Concretely, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available.
 As areas are part of a city, each local area is also subject to the inflow of potential students from other areas, as well of the outflow of students to other areas.
 The magnitude of the net flow will depend on the travel distance between all other areas and the area in question, which here we consider to be the public transport commuting time between areas.
-The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unewually across space.
+The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unequally across space.
 
@@
 The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
-This early work made the distinction between accessibility indeces that apply to zones or single 'desire lines':
+This early work made the distinction between accessibility indices that apply to zones or single 'desire lines':
 "relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility].
@@
 
-Previous literature for consolidated has found that school accessibility has a significant and positive impact on gradutation from upper secondary schools [@falch2013], and on post-compulsory education enrolment [@dickerson2012].
+Previous literature for consolidated has found that school accessibility has a significant and positive impact on graduation from upper secondary schools [@falch2013], and on post-compulsory education enrolment [@dickerson2012].
 For emerging economies, [@asahi2014] finds that improved metro accessibility had a negative impact on student's performance as measured by test scores.
@@
 The São Paulo Metropolitan Region (SPMR) is a large metropolis located in the South West of Brazil.
-It extends for over 7,700 squared kilometers and groups 38 different minucipalities besides the municipality of São Paulo itself.
+It extends for over 7,700 squared kilometers and groups 38 different municipalities besides the municipality of São Paulo itself.
 The estimated population in 2010 was close to 19.5 million in 2010.
@@
 Some of the bus network operates on dedicated bus lines, which amounted to 500 km in 2015.
-As can be seen in Figure 1, the richer areas of the SPMR are better served by public transport, while n important number of lower income areas is underserved.
+As can be seen in Figure 1, the richer areas of the SPMR are better served by public transport, while n important number of lower income areas is under-served.
 
@@
 
-The strucuture of the educational system is composed by both public and private schools.
-The public schools are divided in the folllowing levels: Pre school – 4-5 years (municipal administration); Fundamental – 6 – 14 (municipal administration); Secondary “medio” – 15-17 (state administration) and Superior (state and federal administration).
+The structure of the educational system is composed by both public and private schools.
+The public schools are divided in the following levels: Pre school – 4-5 years (municipal administration); Fundamental – 6 – 14 (municipal administration); Secondary “medio” – 15-17 (state administration) and Superior (state and federal administration).
 
@@
 In principle, the age range of secondary education students is 15-18.
-For each school, have geo-localized each postal adress for a total of XX public primary and XX secondary schools in the SPRM.
+For each school, have geo-localized each postal address for a total of XX public primary and XX secondary schools in the SPRM.
 
@@
 We have information at the travel flow-level.
-geographic coordinates are provided for the origin and destination of each trip, as well as additional information such as trip purpose (includng education), trip duration, number of transfers and mode.
+geographic coordinates are provided for the origin and destination of each trip, as well as additional information such as trip purpose (including education), trip duration, number of transfers and mode.
 The O-D Survey also provides the exact location of schools.
@@
 Of these, 5.5 percent were multimodal.
-In those cases, we assing the mode of the trip leg with the largest duration.
+In those cases, we assign the mode of the trip leg with the largest duration.
 We add all trip leg durations (in minutes) to obtain the total trip duration.
@@
 d = c(-46.34078, -23.37746) # coordinates of origin
-result = dist_google(from = o, to = d, mode = "transit")
+dist_google(from = o, to = d, mode = "transit")
 ```
@@
 To ensure that the result was relevant to school travel, the times were calculated for arrival time at 9am on a weekday.
-The function was repeated for all origin-destinatio pairs.
+The function was repeated for all origin-destination pairs.
 
@@
 
-Enrollment in public secondary schools is more dispersed spatially, as in principle public shcools are not clearly geographically concentrated within the region.
+Enrolment in public secondary schools is more dispersed spatially, as in principle public schools are not clearly geographically concentrated within the region.
 There is, however, an important degree of variation in the school seats in each sub-area, ranging from zero to 7,754.
@@
 
-This measure takes into account the sptail distribution of schooling opportunities, but not the local demand for schooling.
-This is particularly relevant for our analysis, since it could be the case that some areas are disproportionally served with respect to the number of potentail students living within a certain travel distance.
+This measure takes into account the spatial distribution of schooling opportunities, but not the local demand for schooling.
+This is particularly relevant for our analysis, since it could be the case that some areas are disproportionally served with respect to the number of potential students living within a certain travel distance.
 In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (supply), and the sum of individuals within the school grade age-group living in each area (demand) in the following competitive accessibility measure, first proposed by [@shen1998location]:
@@
 To simulate the effects of the proposed educational reform on accessibility, assign
-students in areas with lower than average enrollment to the next area with higher
-than average enrollment
+students in areas with lower than average enrolment to the next area with higher
+than average enrolment
 

### 2016-05-31 paper.Rmd
Reword introduction

@@
 Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas.
-In this paper, we propose a new school accessibility index that combines information of the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure.
+In this dynamic and highly political context, new quantitative evidence can help shed light on the the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing.
+Moreover, the results should help design policies that are more inclusive and equitable by highlighting areas where the provision of schools, travel to school opportunities, or both, are particularly restrictive.
+
+Motivated by this political context this paper presents a new school accessibility index.
+It combines information about the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure.
 We calculate a school accessibility index for 633 areas within the SPMR and use it to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility.
 We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.
-(INCLUDE SOMETHING ABOUT CONTESTATION)
 
-The new index we propose here constitutes a methodological contribution to the measurement of  school accessibility.
+<!-- (INCLUDE SOMETHING ABOUT CONTESTATION) -->
+
+The new index contributes to the measurement of school accessibility.
 It is inspired by the index developed by Shen (1998) in the context of job accessibility.

### 2016-05-31 paper.Rmd
Update figures with captions

@@
 output:
-  word_document: default
   pdf_document:
@@
     keep_md: yes
+  word_document: default
 bibliography: school-access-bib.bib
@@
 
-```{r inc-trans, echo=FALSE, out.width="80%"}
+```{r inc-trans, echo=FALSE, out.width="80%", fig.cap="Public transport system and income per capita in São Paulo, 2010."}
 knitr::include_graphics("income_pc.png")
@@
 
-#Data
+# Data
 
@@
 
-# Preliminary facts
+# Initial findings
 
@@
 
-![Private secondary schools students home location](med_pr.png)
-![Public secondary schools students home location](med_pu.png)
+```{r, echo=FALSE, fig.cap="The home location of students attending public secondary schools students in São Paulo.", out.width="70%"}
+knitr::include_graphics(c("med_pu.png"))
+```
+
+```{r, echo=FALSE, fig.cap="The home location of students attending private secondary schools students in São Paulo.", out.width="70%"}
+knitr::include_graphics(c("med_pr.png"))
+```
+
+<!-- ![](med_pr.png) -->
+<!-- ![Public secondary schools students home location](med_pu.png) -->
 

### 2016-06-02 paper.Rmd
Include formatted table
Latest paper update (from Ana)

@@
 
-In many large Latin American cities such as São Paulo, growing social
+In many large Latin American urban areas such as the São Paulo Metropolitan Region (SPMR), growing social
 and economic inequalities are embedded through unfair education and
@@
 education and the social movements contesting increases in public
-transport fares. Another manifestation is to be found in the concept of
-school accessibility. This paper sheds light on the transport-education
+transport fares. This paper sheds light on the transport-education
 inequality nexus with reference to a new school accessibility measure
-applied to São Paulo. By capturing both the unequal distribution of schools
-and transport services across space, the index allows embedded
-inequalities to be better understood and, with political will,
-contested. Our index combines information on the spatial distribution of
-adolescents, the location of existing schools, the travel patterns of students, and the travel infrastructure serving the school catchment area into a single measure.
-The index is used to measure school accessibility locally across São Paulo, using
-data sources from Population and School Censuses, commonly available in
-Latin American cities. The results illustrate how existing inequalities
+applied to the SPMR. Our index combines information on the spatial distribution of
+adolescents, the location of existing schools, and the travel infrastructure serving the school catchment area into a single measure.
+The index is used to measure school accessibility locally across 633 areas within the SPMR.
+The results illustrate how existing inequalities
 are amplified by variable accessibility to schools across income groups
-and geographical space. We conclude that extending the concept of local
-accessibility indicators to education can help to both contest and
-constructively tackle embedded social inequalities.
+and geographical space.
 
@@
 
-Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the São Paulo Metropolitan Region gained access to a public transport subsidy in 2015.
+Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education.
+This is manifested by the overlap between recent protests against unwanted changes to public education and the social movements contesting increases in public transport fares.
+Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups.
+Conversely, poor accessibility to deprived area can reinforce social inequalities, with long-term implications.
+Based on this emerging context, this paper develops a school accessibility index for local areas.
+
+The São Paulo Metropolitan Region (SPMR) presents an excellent opportunity to measure the extent of educational and transport infrastructure, and to study the impact of public policies on different socio-economic groups.
+Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the SPMR gained access to a public transport subsidy in 2015.
 However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016.
@@
 
-<!-- (INCLUDE SOMETHING ABOUT CONTESTATION) -->
+The first well-known attempt to define accessibility quantitatively was by [@ingram1971concept], which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
+This early work made the distinction between accessibility indices that apply to zones or single 'desire lines':
+"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility].
+The measurement of access to jobs by particular transport modes has received vast attention in the literature [@geurs2004], while studies on access to education has largely focused on the consequences of mode choice on socio-economic indicators and school outcomes (see for instance [@asahi2014], , [@falch2013] and [@dickerson2012] on the impact of accessibility on school outcomes, and [@andersson2012] and [@easton2015] on the effect of changes in travel-to-school patterns in developed economies).
 
 The new index contributes to the measurement of school accessibility.
-It is inspired by the index developed by Shen (1998) in the context of job accessibility.
+It is inspired by the index developed by [@shen1998location] in the context of job accessibility.
 The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' of students in each area.
@@
 The index thus captures the fact that students compete for school seats, which are limited in number, and distributed unequally across space.
+The inclusion of this competitive element to school accessibility, akin to that in job access, highlights the fact that under certain educational systems, the access to opportunities is mediated by competition.
 
-HERE REVIEW PREVIOUS SCHOOL ACCESSIBILITY MEASURES AND EXPLAIN THEIR LIMITATIONS VIS-A-VIS OUR INDEX
---Lit review
-
-The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
-This early work made the distinction between accessibility indices that apply to zones or single 'desire lines':
-"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility].
-
-Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education.
-This is manifested by the overlap between recent protests against unwanted changes to public education and the social movements contesting increases in public transport fares.
-Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups.
-Conversely, poor accessibility to deprived area can reinforce social inequalities, with long-term implications.
-Based on this emerging context, this paper develops a school accessibility index for local areas.
-
-References transport and social exclusion
-[@jones2012] provides a general framework to understand transport-related social exclusion in emerging economies
-
-Evidence on access inequalities in the Brazilian context [@gomide2006]
-
-Active travel children and adolescents in the SPMR [@de2015]
-
-Previous literature for consolidated has found that school accessibility has a significant and positive impact on graduation from upper secondary schools [@falch2013], and on post-compulsory education enrolment [@dickerson2012].
-For emerging economies, [@asahi2014] finds that improved metro accessibility had a negative impact on student's performance as measured by test scores.
-...
+Our work contributes to the literature on transport-related social exclusion in emerging economies [@jones2012] by providing a quantitative way to assess multiple dimensions into a single measure.
+It provides a benchmark to compare access inequalities in countries with unequal provision of public services [@gomide2006], [@pacione1989access].
+In the context of contested mobilities, by capturing both the unequal distribution of schools and transport services across space, the index allows embedded inequalities to be better understood and, with political will, contested.
 
@@
 
-```{r, echo=FALSE, fig.cap="The home location of students attending public secondary schools students in São Paulo.", out.width="70%"}
+```{r, echo=FALSE, fig.cap="The home location of students attending public secondary schools students in the SPMR.", out.width="70%"}
 knitr::include_graphics(c("med_pu.png"))
@@
 
-```{r, echo=FALSE, fig.cap="The home location of students attending private secondary schools students in São Paulo.", out.width="70%"}
+```{r, echo=FALSE, fig.cap="The home location of students attending private secondary schools students in the SPMR.", out.width="70%"}
 knitr::include_graphics(c("med_pr.png"))
 ```
-
-<!-- ![](med_pr.png) -->
-<!-- ![Public secondary schools students home location](med_pu.png) -->
 
@@
 
-![Enrollment in public secondary schools](enrol.png)
+```{r, echo=FALSE, fig.cap="Enrollment in public secondary schools in the SPMR.", out.width="70%"}
+knitr::include_graphics(c("enrol.png"))
+```
 
@@
 
-```{r table-placeholder, echo=FALSE}
-
+```{r table-placeholder, echo=FALSE, warning=FALSE}
+tbl = read.csv("res-table.csv")
+names(tbl) = c("Quantile", "Pre-reform", "Post-reform", "% change")
+knitr::kable(tbl, row.names = F, caption = "Mean accessibility before and after redistribution")
 ```
@@
 
-# Discussion
+# Conclusion
 
-# Conclusion
+ We conclude that extending the concept of local
+accessibility indicators to education can help to both contest and
+constructively tackle embedded social inequalities.
 

### 2016-06-06 paper.Rmd
New method figures, description for paper
Various updates to the paper

@@
 o = c(-46.31813, -23.38584) # coordinates of origin
-d = c(-46.34078, -23.37746) # coordinates of origin
+d = c(-46.34078, -23.37746) # coordinates of destination
 dist_google(from = o, to = d, mode = "transit")
@@
 The function was repeated for all origin-destination pairs.
+To ensure the correct OD data was being generated, the results were plotted on an interactive map (Figure xx).
+The results show the high variability of route distance vs Euclidean distance ('circuity'), providing further evidence of inequalities in public transport provision in the city (Figure xx).
+
+```{r , echo=FALSE, fig.cap="Visual representation of the 12,697 OD pairs routed through the Google Distance Matrix API on an interactive map in RStudio, an open source data analysis platform.", out.width="100%"}
+knitr::include_graphics("od-map.png")
+```
+
+```{r euclid, echo=FALSE, fig.cap="Euclidean (straight line) vs public transport distances resulting from routing the 12,697 OD pairs of interest through the Google Distance Matrix API.", out.width="100%"}
+knitr::include_graphics("public-vs-euclid-dists.png")
+```
 
@@
 Although the mean duration of travel to school remained stable over the period 1997-2007 ([@de2015]), they could increase as a result of several factors relevant to the SPMR, including sub-urbanization of particular demographic groups, such as those with higher income; increase in the size of schools leading to an increase in school's catchment area; less than proportional increase in schools in high population growth areas with poor connectivity; and institutional changes towards more flexible school choice [@easton2015].
+
+\newpage
 

### 2016-06-06 paper.Rmd
Various updates to the paper

@@
   the São Paulo Metropolitan Region'
-author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação
-  Getulio Vargas) and Robin Lovelace (University of Leeds)
+author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University of Leeds)  and Frederico R. Ramos (Fundação
+  Getulio Vargas) 
 output:
@@
 and economic inequalities are embedded through unfair education and
-transport systems. Inequalities in educational and transport
-infrastructure are mutually reinforcing: the right to mobility is
-intrinsically linked to the right to education. This is manifested by
-the overlap between recent protests against unwanted changes to public
-education and the social movements contesting increases in public
-transport fares. This paper sheds light on the transport-education
+transport systems. This paper sheds light on the transport-education
 inequality nexus with reference to a new school accessibility measure
@@
 Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education.
-This is manifested by the overlap between recent protests against unwanted changes to public education and the social movements contesting increases in public transport fares.
 Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups.
 Conversely, poor accessibility to deprived area can reinforce social inequalities, with long-term implications.
-Based on this emerging context, this paper develops a school accessibility index for local areas.
+Against this background, in this paper we propose a new way to masure school accessibility in local areas within an urban area, and aply it for the case of the São Paulo Metropolitan Region (SPMR).
 
-The São Paulo Metropolitan Region (SPMR) presents an excellent opportunity to measure the extent of educational and transport infrastructure, and to study the impact of public policies on different socio-economic groups.
+The SPMR presents an excellent opportunity to measure the extent of educational and transport infrastructure, and to study the impact of public policies on different socio-economic groups.
 Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the SPMR gained access to a public transport subsidy in 2015.
@@
 
-Motivated by this political context this paper presents a new school accessibility index.
-It combines information about the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure.
-We calculate a school accessibility index for 633 areas within the SPMR and use it to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility.
+Our proposed competitive accessibility measure combines information about the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single measure.
+We start from a cumulative opportunity measure that counts the number of schools seats that can be reached within a 30 minute journey by public transport, and then build a competitive measure that takes into account the place of residency of teens.
+Travel-to-school times by public transport are based on actual commuting times obtained through the Google Travel Distance API.
+We calculate these measures for 633 areas within the SPMR.
+We use the competitive meausre to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility.
 We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.
 
-The first well-known attempt to define accessibility quantitatively was by [@ingram1971concept], which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
-This early work made the distinction between accessibility indices that apply to zones or single 'desire lines':
-"relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility].
-The measurement of access to jobs by particular transport modes has received vast attention in the literature [@geurs2004], while studies on access to education has largely focused on the consequences of mode choice on socio-economic indicators and school outcomes (see for instance [@asahi2014], , [@falch2013] and [@dickerson2012] on the impact of accessibility on school outcomes, and [@andersson2012] and [@easton2015] on the effect of changes in travel-to-school patterns in developed economies).
+The first well-known attempt to define accessibility quantitatively was by @ingram1971concept, which presented a range of measures related to distance (Euclidean and network), barriers and different functions representing distance decay.
+This early work made the distinction between accessibility indices that apply to zones or single 'desire lines': "relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area" [@allen1993accesibility].
+In subsequent works, the measurement of job accessibility by particular transport modes has received vast attention in the literature [@geurs2004], while studies on access to education has largely focused on the consequences of mode choice on socio-economic indicators and school outcomes (see for instance @asahi2014, @falch2013 and @dickerson2012 on the impact of accessibility on school outcomes, and @andersson2012 and @easton2015 on the effect of changes in travel-to-school patterns in developed economies).
 
-The new index contributes to the measurement of school accessibility.
-It is inspired by the index developed by [@shen1998location] in the context of job accessibility.
-The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' of students in each area.
+Our index is inspired by the index developed by @shen1998location in the context of job accessibility.
+The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the 'supply' and 'demand' for schooling in each area.
 Concretely, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available.
@@
 Our work contributes to the literature on transport-related social exclusion in emerging economies [@jones2012] by providing a quantitative way to assess multiple dimensions into a single measure.
-It provides a benchmark to compare access inequalities in countries with unequal provision of public services [@gomide2006], [@pacione1989access].
-In the context of contested mobilities, by capturing both the unequal distribution of schools and transport services across space, the index allows embedded inequalities to be better understood and, with political will, contested.
+It also provides a benchmark to compare access inequalities in countries with unequal provision of public services [@gomide2006], [@pacione1989access].
+Considering conestation as the process through which social groups mobilize in an organized way in order to impede the implementation of unwanted policies, or force the negotiation of new conditions, our index allows embedded inequalities to be better understood.
 
@@
 Regarding the educational system, since the constitution of 1946, primary education is compulsory.
-In 2005, the legislation changed to increase the duration of the primary education, going from 8 to 9 years.
+In 2005, the legislation changed to increase the duration of the primary education, going from eight to nine years.
 In 2010, all regions finished the transition.
-In 2009, a constitutional amendment imposed compulsory education for the primary “fundamental” and secondary “médio” levels.
+In 2009, a constitutional amendment imposed compulsory education for the primary and secondary levels.
 The constitution imposes that this resolution should be 100% implemented by 2016.
-
 By law, the state/municipalities should guaranty the adequate provision (100% coverage of the demand).
-There are national and state educational funds that are distributed among the schools  according to the number of enrolled students and  existence of facilities (library , laboratories).
+There are national and state educational funds that are distributed among the schools according to the number of enrolled students and  existence of facilities (library, laboratories, etc.).
 
 The structure of the educational system is composed by both public and private schools.
-The public schools are divided in the following levels: Pre school – 4-5 years (municipal administration); Fundamental – 6 – 14 (municipal administration); Secondary “medio” – 15-17 (state administration) and Superior (state and federal administration).
+The public schools are divided in the following levels: Pre school (\textit{clase de alfabetização}) – 4-5 years (municipal administration); Fundamental (\textit{ensino fundamental}) – 6 – 14 (municipal administration); Secondary (\textit{ensino médio}) – 15-17 (state administration) and Post-secondary (\textit{ensino superior}) (state and federal administration).
 
 According to the Censo Escolar 2010, there are 10,251 schools in the SPMR, out of which  54% are public (22% state / 32% municipal) and 46% are private.
-Geographically, the student is assigned to the closer school or to their house or their parent’s working place.
-In the secondary, the student can ask for a vacancy in a different school with a justification and under the condition of the availability of vacancy.
+Geographically, students are assigned to the closest school to their place of residency or their parent’s working place.
+For secondary school, the student can ask for a vacancy in a different school after providing a valid justification and under the condition of the availability of vacancy.
 
@@
 We use the coordinates contained in the database provided by the CEM (Centro de Estudos da Metrópole) for 2001 to geolocate the schools in 2008 using the unique identifier (FRED TO EXPLAIN HOW WE GEO-LOCATED THE REST).
-A clarification about the educational system in Brazil is in place here.
-Primary education \textit{ensino fundamental} comprises nine years, divided on a first year of basic literacy \textit{clase de alfabetização}, followed by eight `` series''.
-Primary education is mandatory for children aged 6-14.
-Secondary education lasts three years.
-In principle, the age range of secondary education students is 15-18.
-For each school, have geo-localized each postal address for a total of XX public primary and XX secondary schools in the SPRM.
+For each school, we have geo-localized a total of 4,612 public schools in the SPRM.
 
-Our second source is information the 2010 Population Census of Brazil at the \textit{setor censitario} level, compiled and freely distributed by the Brazilian Institute of Statistics (IBGE).
-This spatial unit is equivalent to enumeration areas defined for surveying purposes, and contain on average 400 households.
-There is a total of 30,815 areas in the SPMR.
-IBGE also provides the digital networks containing the boundaries of the enumeration areas.
+Our second source is information the 2010 Population Census of Brazil, compiled and freely distributed by the Brazilian Institute of Statistics (IBGE).
+We aggregate the data by Area de Ponderacão Espacial (AEP) areas, a spatial unit defined for surveying purposes by IBGE.
+There is a total of 633 areas in the SPMR.
+IBGE provides the digital networks containing the boundaries of these areas.
 From the census microdata, we have data on the number of inhabitants by age for each enumeration area.
@@
 
-Our last source is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority Mêtro.
-We have information at the travel flow-level.
-geographic coordinates are provided for the origin and destination of each trip, as well as additional information such as trip purpose (including education), trip duration, number of transfers and mode.
-The O-D Survey also provides the exact location of schools.
-The survey also provides relevant information on individuals undertaking the trips, such as their age, income range and educational status (including student).
-We focus on trips made by secondary school students with an indicated motive ``education'' at the destination.
-The total number of trips (excluding missing values) was 11,845.
-Of these, 5.5 percent were multimodal.
-In those cases, we assign the mode of the trip leg with the largest duration.
-We add all trip leg durations (in minutes) to obtain the total trip duration.
-
-The Google Distance Matrix API was used to find the travel times and distances by public transport.
-This was implemented using an interface to the API, using the R package **stplanr** [@stplanr].
+Our last source of information is an Google Distance Matrix API.
+We use this API to find the travel times and distances by public transport.
+This was implemented using an interface to the API by means of the R package **stplanr** [@stplanr].
 To provide an example of how the method works, a reproducible example is illustrated below.
@@
 The function was repeated for all origin-destination pairs.
-To ensure the correct OD data was being generated, the results were plotted on an interactive map (Figure xx).
-The results show the high variability of route distance vs Euclidean distance ('circuity'), providing further evidence of inequalities in public transport provision in the city (Figure xx).
+To ensure the correct OD data was being generated, the results were plotted on an interactive map (Figure 1).
+The results show the high variability of route distance vs Euclidean distance ('circuity'), providing further evidence of inequalities in public transport provision in the city (Figure 2).
 
@@
 
-# Initial findings
+# Preliminary findings
 
-As can be seen by comparing Figures 1, 2 and 3, students disproportionally attend private secondary schools in higher income areas, while the opposite is true for public secondary schools.
+In this section we discuss a set of results that serve as motivation for the implementation of our index.
+We start by comparing the spatial distribution of public and private secondary school attendance.
+As can be seen by comparing Figures 4 and 5, with Figure 1, students in the SPMR disproportionally attend private secondary schools in higher income areas, while the opposite is true for public secondary schools.
 
@@
 Table 1 shows the model split by type of secondary school student.
-Clearly, secondary school students attending private schools disproportionally travel by car, compared to public school students (9% versus 47%), even though higher income areas are better served by public transport.
-This is in line with the findings of [@de2015].
+(FOOTNOTE the source of this information is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority Mêtro.
+We focus on trips made by secondary school students with an indicated motive ``education'' at the destination.
+The total number of trips (excluding missing values) was 11,845.
+Of these, 5.5 percent were multimodal.
+In those cases, we assign the mode of the trip leg with the largest duration.
+We add all trip leg durations (in minutes) to obtain the total trip duration.) Clearly, secondary school students attending private schools disproportionally travel by car, compared to public school students (9% versus 47%), even though higher income areas are better served by public transport.
+This is in line with the findings of @de2015.
 Around 67% of public school students commute to school by active travel (mostly walking, since the percentage of biking is still relatively small), while 23% commute by public transport, which has much longer mean duration than other modes.
@@
 ```{r table-placeholder, echo=FALSE, warning=FALSE}
-tbl = read.csv("res-table.csv")
-names(tbl) = c("Quantile", "Pre-reform", "Post-reform", "% change")
-knitr::kable(tbl, row.names = F, caption = "Mean accessibility before and after redistribution")
+tbl = read.csv("public.csv")
+names(tbl) = c("Mode", "Percent", "Duration (min)")
+knitr::kable(tbl, row.names = T, caption = "Modal split by type of secondary school")
 ```
 
-Although the mean duration of travel to school remained stable over the period 1997-2007 ([@de2015]), they could increase as a result of several factors relevant to the SPMR, including sub-urbanization of particular demographic groups, such as those with higher income; increase in the size of schools leading to an increase in school's catchment area; less than proportional increase in schools in high population growth areas with poor connectivity; and institutional changes towards more flexible school choice [@easton2015].
+Although the mean duration of travel to school remained stable over the period 1997-2007 [@de2015], they could increase as a result of several factors relevant to the SPMR, including sub-urbanization of particular demographic groups, such as those with higher income; increase in the size of schools leading to an increase in school's catchment area; less than proportional increase in schools in high population growth areas with poor connectivity; and institutional changes towards more flexible school choice [@easton2015].
 
@@
 
-To define school accessibility, we start by defining the broad concept of accessibility, as used in the transport literature.
-According to (@geurs2004, p.
+To define school accessibility, we start with the broad concept of accessibility, as used in the transport literature.
+According to @geurs2004 (p.
 128) accessibility is “the extent to which land-use and transport systems enable (groups of) individuals to reach activities or destinations by means of a (combination of) transport mode(s).” Extended to our case, school accessibility then measures the extent to which the existing built-environment and transport facilities enable children and adolescents in school age to reach schools.
 
-Ideally, a school accessibility measure would include information on the (almost exact) home location of every student and the location of every school they could potentially attend (as in [@andersson2012]), and some measure of commuting costs for any chosen mode (in terms of time and money).
+Ideally, a school accessibility measure would include information on the home location of every student and the location of every school they could potentially attend (as in @andersson2012), and some measure of commuting costs for any chosen mode (in terms of time and money).
 Given that this level of detail is not available in our data, we rely on area aggregates, implicitly assuming that the measure at the centroid of the area applies homogeneously within the area.
+
 A previous work studying the impact of metro expansions on test scores for Santiago de Chile measures school accessibility as the distance of every existing school to the nearest subway station [@asahi2014].
@@
 
-$$ CO_{i}^M= \sum_{j=1}^n(S_{j}f(C_{ij}) $$
+$$ CO_{i}^M= \sum_{j=1}^n(O_{j}f(C_{ij}) $$
 
@@
 
-This measure counts the number of schools available from one area within a certain travel time threshold by mode $M$.
+This measure counts the number of "opportunities" $O$, in our case schools available, from one area within a certain travel time threshold by mode $M$.
 $C_{ij}$ is the travel cost (measured in time) between the centroid of zone $i$ and the centroid of zone $j$, and $f(C_{ij})$ is a weight function.
@@
 This is particularly relevant for our analysis, since it could be the case that some areas are disproportionally served with respect to the number of potential students living within a certain travel distance.
-In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (supply), and the sum of individuals within the school grade age-group living in each area (demand) in the following competitive accessibility measure, first proposed by [@shen1998location]:
+In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (supply), and the sum of individuals within the school grade age-group living in each area (demand) in the following competitive accessibility measure, first proposed by @shen1998location:
 
-$$ CA_{i}^M= \sum_{j=1}^n\frac{O_{j}f(C_{ij})}{\sum_{k=1}^n(t_{k}f(C_{kj}))}$$
+$$ CA_{i}^M= \sum_{j=1}^n\frac{O_{j}f(C_{ij})}{\sum_{k=1}^n(T_{k}f(C_{kj}))}$$
 
-Where the numerator discounts the number of "opportunities" (school seats) in area $j$ $P_{j}$ by how far area $i$ is from area $j$ using the same function as before, and the denominator discounts the number of teens in secondary school age living in zone $k$ $T_{k}$ by how far they are from area $j$.
+Where the numerator discounts the number of school seats in area $j$ $O_{j}$ by how far area $i$ is from area $j$ using the same function as before, and the denominator discounts the number of teens in secondary school age living in zone $k$ $T_{k}$ by how far they are from area $j$.
 In other words, the numerator counts how many schools seats can be reached from an area, while the denominator counts how many teens can potentially reach the same area.
@@
 
+We calculate both measures for commutes to school by public transport.
+We set a threshold of 30 minutes based on the mean average travel times by this mode (see Table 1).
+We assign a minimum time of zero for commutes within the same area (i.e., schools within the same area are added to the index in all cases).
+
 # Results
 
-INLCUDE MAP COMPETITIVE ACCESIBILITY INDEX BY PUBLIC TRANSPORT, INTERPRETATION OF VALUES, SENSITIVITY ANALYSIS
+Figure XX shows the spatial distribution of the cumulative-opportunity measure.
+The index has a mean value of 3,777, which indicates that for teens residing in the average area, there are 3,777 school seats that can be reached within a journey by public transport of up to 30 minutes.
+The standard deviation is 3,485 indicates large variation in the level of accessibility across areas.
+As can be seen on the Figure XX, areas with low accessibility include not only the expaded financial business district located on the central south west area, where there are not many teens or schools to start with (see Figures XX and XX), but also areas with deficient public transport provision, such as some north and south eastern areas.
 
-POLICY EXPERIMENT 
-To simulate the effects of the proposed educational reform on accessibility, assign
-students in areas with lower than average enrolment to the next area with higher
-than average enrolment
+```{r, echo=FALSE, fig.cap="Cumulative opportunity index", out.width="70%"}
+knitr::include_graphics(c("cum_index.png"))
+```
 
-INCLUDE TABLE MEAN ACCESSIBILITY BEFORE AND AFTER REDISTRIBUTION, INTERPRETATION
+Figure XX shows the spatial distribution of the competitive accessibility measure.
+Although it largely follows the distribution of the cumulative index, it corrects the value in some areas based on the place of residence of students across the SPRM.
+To interpret the value of the index, recall that if all students in an area attended secondary schools in public secondary schools in their area (i.e., no student would travel outside her own area to attend school), the index would take the value of 1.
+Then, in areas with a value lower than one, the provision of public schools and public transport is such that there is local underprovision, in the sense that the number of schools seats that can be reached from the area is smaller than the number of teens that can potentially reach the area (the opposite is true for values higher than one).
+Note that some relatively close by areas, especially in the eastern part of the city, have striking differences in the level of accessibility.
+In some cases, an area can have twice the level of accessibility than the most proximate area.
+This can be due to differences in the local provision of public schools, public transport or the number of students living within the reach of the areas.
 
-# Conclusion
+```{r, echo=FALSE, fig.cap="Competitive accessibility index", out.width="70%"}
+knitr::include_graphics(c("comp_index.png"))
+```
 
- We conclude that extending the concept of local
-accessibility indicators to education can help to both contest and
-constructively tackle embedded social inequalities.
+To understand the effects of changes in public school provision on school accessiblity, we conduct a comparative static analysis using our competitive index.
+To simulate the effects of a policy that seeks to spatially concentrate students in larger public schools by closing down smaller schools, we assign students in areas with lower than average enrolment to the next area with higher than average enrolment, while keeping the distribution of teens and public transport constant.
+We then compare the changes in school accessibility for different quartiles before and after this change.
+Table XX summarizes the results.
+
+```{r , echo=FALSE, warning=FALSE}
+tbl = read.csv("res-table.csv")
+names(tbl) = c("Quantile","Pre-change", "Post-change", "Change (%)")
+knitr::kable(tbl, row.names = F, caption = "Mean accessibility before and after redistribution")
+```
+
+The policy reduces further the accessibility level of areas in the first and second quantiles of accessibility.
+Areas in the first quantile experience an average reduction of 21% in school accessibility levels as a result of the changes in the spatial distribution of schools.
+In the mean time, areas that had high levels of accesibility to start with, experience an increase.
+The results suggest that concentrating the provision of public secondary schools accentuates the already large differences in accesibility levels across areas.
+
+# Conclusion and discussion
+
+The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive, in the sense that teens living in areas that had the lowest levels of accessibility are the ones who experience the largest negative impact from the policy.
+They are likely to be worse off compared to an already difficult situation of low local schooling provision, low access by public transport and/or high competition by teens in other areas.
+In practice, such a change would probabily imply larger commuting times for students residing in areas with low accessibility, which can have further consequences in terms of attendance, drop-out likelihood and performance.
+
+Although the simulated scenario represents a rather extreme change in the distribution of public schooling, the way in which students with different levels of accessibility are impacted remains.
+In the absence of a public transport subsidy for school commutes, students that experience a significant increase in travel-to-school times may be force to change their commute mode from an active (and free of charge mode) to a paid-by mode, with the consequent increase in their monetary spent in commuting to school.
+The most negative effect is then felt by low-incomes students living in areas with low school accessibility, for whom the marginal increase in commuting costs represents a much larger share in their budget.
+
+The provision of a public transport subsidy for school commutes is unlikely to compensate students at the bottom of the income and accessibility ladder if the increase in commuting times has consequences in terms of motivation to stay in school and performance.
+In this sense, a “trade-off” logic where less schooling provision is compensated by public transport subsidies does not operate when existing inequalities in provision are large.
+
+Our study can be further extended to account for other modes of school commute, such as active travel.
+It can also benefit from a finer spatial dissagregation to capture within area inequalities, which may be substantial and are currently hidden away in area aggregates.
+A future extension of the index can incorporate a measure of school quality, which in so far assumed away.
+It will also be highly informative to calculate the proposed indices for other contexts and for different time periods in order to understand the true magnitude of the extent of inequality across individuals living in different areas of the SPMR.
+
+We conclude that extending the concept of local accessibility indicators to education can help to both contest and constructively tackle embedded social inequalities.
+Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities.
+Quantitative measures such as the ones proposed here help summarizing multiple dimensions of inequality into a single, intterpretable measure.
+Having this tool may be useful for both policy makers and social groups in discussing and negotiating policies with a highly sensitive social component and evident short and long-term consequences for those affected.
 

### 2016-06-06 paper.Rmd
Paper update from Ana
Comment out non-local data

@@
 ```{r table-placeholder, echo=FALSE, warning=FALSE}
-tbl = read.csv("public.csv")
-names(tbl) = c("Mode", "Percent", "Duration (min)")
-knitr::kable(tbl, row.names = T, caption = "Modal split by type of secondary school")
+# tbl = read.csv("public.csv")
+# names(tbl) = c("Mode", "Percent", "Duration (min)")
+# knitr::kable(tbl, row.names = T, caption = "Modal split by type of secondary school")
 ```
@@
 ```{r, echo=FALSE, fig.cap="Cumulative opportunity index", out.width="70%"}
-knitr::include_graphics(c("cum_index.png"))
+# knitr::include_graphics(c("cum_index.png"))
 ```
@@
 ```{r, echo=FALSE, fig.cap="Competitive accessibility index", out.width="70%"}
-knitr::include_graphics(c("comp_index.png"))
+# knitr::include_graphics(c("comp_index.png"))
 ```

### 2016-06-06 paper.Rmd
Fix typos

@@
   the São Paulo Metropolitan Region'
-author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University of Leeds)  and Frederico R. Ramos (Fundação
-  Getulio Vargas) 
+author: Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University
+  of Leeds)  and Frederico R. Ramos (Fundação Getulio Vargas)
 output:
+  word_document: default
   pdf_document:
@@
     keep_md: yes
-  word_document: default
 bibliography: school-access-bib.bib
@@
 Conversely, poor accessibility to deprived area can reinforce social inequalities, with long-term implications.
-Against this background, in this paper we propose a new way to masure school accessibility in local areas within an urban area, and aply it for the case of the São Paulo Metropolitan Region (SPMR).
+Against this background, in this paper we propose a new way to measure school accessibility in local areas within an urban area, and ply it for the case of the São Paulo Metropolitan Region (SPMR).
 
@@
 We calculate these measures for 633 areas within the SPMR.
-We use the competitive meausre to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility.
+We use the competitive measure to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the effect of a policy aiming at concentrating public secondary education provision on school accessibility.
 We find that closing down schools in areas with lower than average provision is highly regressive: it negatively impacts students with the lowest accessibility levels, accentuating existing inequalities.
@@
 To interpret the value of the index, recall that if all students in an area attended secondary schools in public secondary schools in their area (i.e., no student would travel outside her own area to attend school), the index would take the value of 1.
-Then, in areas with a value lower than one, the provision of public schools and public transport is such that there is local underprovision, in the sense that the number of schools seats that can be reached from the area is smaller than the number of teens that can potentially reach the area (the opposite is true for values higher than one).
+Then, in areas with a value lower than one, the provision of public schools and public transport is such that there is local under-provision, in the sense that the number of schools seats that can be reached from the area is smaller than the number of teens that can potentially reach the area (the opposite is true for values higher than one).
 Note that some relatively close by areas, especially in the eastern part of the city, have striking differences in the level of accessibility.
@@
 
-To understand the effects of changes in public school provision on school accessiblity, we conduct a comparative static analysis using our competitive index.
+To understand the effects of changes in public school provision on school accessibility, we conduct a comparative static analysis using our competitive index.
 To simulate the effects of a policy that seeks to spatially concentrate students in larger public schools by closing down smaller schools, we assign students in areas with lower than average enrolment to the next area with higher than average enrolment, while keeping the distribution of teens and public transport constant.
@@
 Areas in the first quantile experience an average reduction of 21% in school accessibility levels as a result of the changes in the spatial distribution of schools.
-In the mean time, areas that had high levels of accesibility to start with, experience an increase.
-The results suggest that concentrating the provision of public secondary schools accentuates the already large differences in accesibility levels across areas.
+In the mean time, areas that had high levels of accessibility to start with, experience an increase.
+The results suggest that concentrating the provision of public secondary schools accentuates the already large differences in accessibility levels across areas.
 
@@
 They are likely to be worse off compared to an already difficult situation of low local schooling provision, low access by public transport and/or high competition by teens in other areas.
-In practice, such a change would probabily imply larger commuting times for students residing in areas with low accessibility, which can have further consequences in terms of attendance, drop-out likelihood and performance.
+In practice, such a change would probably imply larger commuting times for students residing in areas with low accessibility, which can have further consequences in terms of attendance, drop-out likelihood and performance.
 
@@
 Our study can be further extended to account for other modes of school commute, such as active travel.
-It can also benefit from a finer spatial dissagregation to capture within area inequalities, which may be substantial and are currently hidden away in area aggregates.
+It can also benefit from a finer spatial disaggregation to capture within area inequalities, which may be substantial and are currently hidden away in area aggregates.
 A future extension of the index can incorporate a measure of school quality, which in so far assumed away.
@@
 Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities.
-Quantitative measures such as the ones proposed here help summarizing multiple dimensions of inequality into a single, intterpretable measure.
+Quantitative measures such as the ones proposed here help summarizing multiple dimensions of inequality into a single, interpretable measure.
 Having this tool may be useful for both policy makers and social groups in discussing and negotiating policies with a highly sensitive social component and evident short and long-term consequences for those affected.

### 2016-06-06 paper.Rmd
Add back Ana's content

@@
 However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016.
-The change in policy was estimated to affect over 300,000 students, many whom would be placed in schools far from their homes.
+The change in policy was estimated to affect more than 300,000 students, many whom would be placed in schools far from their homes.
 As a response, in the following months students occupied over 200 schools.
@@
 Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas.
-In this dynamic and highly political context, new quantitative evidence can help shed light on the the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing.
+In this dynamic and highly political context, new quantitative evidence can help shed light on the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing.
 Moreover, the results should help design policies that are more inclusive and equitable by highlighting areas where the provision of schools, travel to school opportunities, or both, are particularly restrictive.
@@
 
-Our last source of information is an Google Distance Matrix API.
-We use this API to find the travel times and distances by public transport.
+The final source of information was derived from the Google Distance Matrix API, which provided travel times and distances by public transport.
 This was implemented using an interface to the API by means of the R package **stplanr** [@stplanr].

### 2016-06-06 paper.Rmd
Add footnote

@@
 
-Table 1 shows the model split by type of secondary school student.
-(FOOTNOTE the source of this information is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority Mêtro.
+Table 1 shows the model split by type of secondary school student.^[The source of this information is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority Mêtro.
 We focus on trips made by secondary school students with an indicated motive ``education'' at the destination.
@@
 In those cases, we assign the mode of the trip leg with the largest duration.
-We add all trip leg durations (in minutes) to obtain the total trip duration.) Clearly, secondary school students attending private schools disproportionally travel by car, compared to public school students (9% versus 47%), even though higher income areas are better served by public transport.
+We add all trip leg durations (in minutes) to obtain the total trip duration.]
+Clearly, secondary school students attending private schools disproportionally travel by car, compared to public school students (9% versus 47%), even though higher income areas are better served by public transport.
 This is in line with the findings of @de2015.

### 2016-06-07 R/routing.R
Update plot

@@ -73,7 +73,7 @@ l$dist_hav_km = l$dist_hav / 1000
 # plotting
 l$dist_band = cut(l$dist_hav_km, breaks = quantile(l$dist_hav_km))
 l$color = cut(l$dist_hav_km, breaks = quantile(l$dist_hav_km),
-              labels = c("green", "blue", "yellow", "red"))
+              labels = c("green", "blue", "orange", "red"))
 l$color = as.character(l$color)
 plot(ldff$dist_hav / 1000, ldff$distances / 1000,
     ylab = "Public transport distance (km)",
@@ -97,3 +97,4 @@ tm_shape(osm_tiles) +
 qtm(l, line.col = "dist_hav_km")
 
 write.csv(ldff, "output-data/sample-time-od-out.csv")
+l = read.csv("output-data/sample-time-od-out.csv")

### 2017-02-26 paper.tex
Add latex paper

--- 2016-02-10 paper.tex
@@
 \usepackage[margin=1in]{geometry}
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{248,248,248}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
+\newcommand{\RegionMarkerTok}[1]{{#1}}
+\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
+\newcommand{\NormalTok}[1]{{#1}}
+\usepackage{longtable,booktabs}
+\usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
 \ifxetex
@@
             bookmarks=true,
-            pdfauthor={Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos (Fundação Getulio Vargas) and Robin Lovelace (University of Leeds)},
-            pdftitle={The unequal distribution of access to education and transport: insights from for São Paulo},
+            pdfauthor={Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)},
+            pdftitle={The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region},
             colorlinks=true,
@@
   \title{The unequal distribution of access to education and transport: insights
-from for São Paulo}
+from the São Paulo Metropolitan Region}
   \pretitle{\vspace{\droptitle}\centering\huge}
   \posttitle{\par}
-  \author{Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Frederico R. Ramos
-(Fundação Getulio Vargas) and Robin Lovelace (University of Leeds)}
+  \author{Ana I. Moreno-Monroy (Autonomous university of Barcelona), Robin Lovelace
+(University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)}
   \preauthor{\centering\large\emph}
@@
 
-In many large Latin American cities such as São Paulo, growing social
-and economic inequalities are embedded through unfair education and
-transport systems. Good schools are mostly concentrated in wealthy
-areas, while transport links to school are deficient in deprived areas,
-exacerbating the issue. Inequalities in educational and transport
-infrastructure are mutually reinforcing: the right to mobility is
-intrinsically linked to the right to education. This is manifested by
-the overlap between recent protests against unwanted changes to public
-education and the social movements contesting increases in public
-transport fares. Another manifestation is to be found in the concept of
-school accessibility. This paper sheds light on the transport-education
+In many large Latin American urban areas such as the São Paulo
+Metropolitan Region (SPMR), growing social and economic inequalities are
+embedded through high spatial inequality in the provision of education
+and transport systems. This paper sheds light on the transport-education
 inequality nexus with reference to a new school accessibility measure
-applied São Paulo. By capturing both the unequal distribution of schools
-and transport services across space, the index allows embedded
-inequalities to be better understood and, with political will,
-contested. Our index combines information on the spatial distribution of
-children and adolescents by income categories, the location of existing
-schools, the travel patterns of students, and the travel infrastructure
-serving the school catchment area into a single measure. The index is
-used to measure school accessibility locally across São Paulo, using
-data sources from Population and School Censuses, commonly available in
-Latin American cities. The results illustrate how existing inequalities
-are amplified by variable accessibility to schools across income groups
-and geographical space. We conclude that extending the concept of local
-accessibility indicators to education can help to both contest and
-constructively tackle embedded social inequalities.
+applied to the SPMR. Our index combines information on the spatial
+distribution of adolescents, the location of existing schools, and the
+travel infrastructure serving the school catchment area into a single
+measure. The index is used to measure school accessibility locally
+across 633 areas within the SPMR. We use the index to simulate the
+impact of a policiy aiming at increasing the centralization of public
+secondary education provision, and find that it negatively affects
+students with the lowest levels of accessibility. These results
+illustrate how existing inequalities are amplified by variable
+accessibility to schools across income groups and geographical space.
+
+\bigskip
+
+\textit{This version: January 2017. Preliminary version. Please do not cite}
 
@@
 
-\textbf{Contents:}
+Inequalities in educational and transport infrastructure are mutually
+reinforcing: the right to mobility is intrinsically linked to the right
+to education. Travel to school options are vital for ensuring a more
+equitable supply of educational opportunity to diverse groups.
+Conversely, poor accessibility to and from deprived areas can reinforce
+social inequalities, with long-term implications. Against this
+background, in this paper we propose a new way to measure school
+accessibility in local areas within an urban area, and apply it for the
+case of the São Paulo Metropolitan Region (SPMR).
 
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Motivations
-\item
-  General measures of accessibility
-\item
-  School accessibility
-\end{itemize}
+The SPMR presents an excellent opportunity to measure the extent of
+educational and transport infrastructure, and to study the impact of
+public policies on different socio-economic groups. Partly as a response
+to the pressures from the Free Pass Movement that started in March,
+2013, low income public school students residing in the SPMR gained
+access to a public transport subsidy in 2015. However, in October 2015,
+the São Paulo state government announced during a television interview
+that as part of a budgetary deficit reduction plan, dozens of secondary
+schools would be closed in 2016. The change in policy was estimated to
+affect more than 300,000 students, many whom would be placed in schools
+far from their homes. As a response, in the following months students
+occupied over 200 schools. Protesters scored some victories: the reform
+was postponed for one year, and the education secretary resigned
+(Ortellado 2015).
 
-\ldots{}
+Although this time the policy change was not carried out, in a context
+of high spatial inequalities in the provision of public transport and
+schooling, it is worth asking whether a public transport subsidy for
+students can compensate for the lower provision of public schools in
+some areas. In this dynamic and highly political context, new
+quantitative evidence can help shed light on the relationship between
+transport and educational inequalities and the extent to which they are
+mutually reinforcing. Moreover, the results should help design policies
+that are more inclusive and equitable by highlighting areas where the
+provision of schools, travel to school opportunities, or both, are
+particularly restrictive.
 
-Travel to school is an everyday reality for millions of young people
-around the world. The mode, duration, safety, comfort and pollution
-levels of this trip has huge impacts on the future generation, yet has
-received relatively little academic attention.
-
-Travel options are vital for ensuring a more equitable supply of
-educational opportunity to diverse groups. Conversely, poor
-accessibility to deprived area can reinforce social inequalities, with
-long-term implications. Based on this emerging context, this paper
-develops a school accessibility index for local areas.
+Our proposed competitive accessibility measure combines information
+about the place of residence of students, the spatial distribution of
+public schools, and public transport accessibility into a single
+measure. We start from a cumulative opportunity measure that counts the
+number of schools seats that can be reached within a 30 minute journey
+by public transport, and then build a competitive measure that takes
+into account the place of residency of teens. Travel-to-school times by
+public transport are based on actual commuting times obtained through
+the Google Matrix Distance API. We calculate these measures for 633
+areas within the SPMR. We use the competitive measure to simulate the
+impact of a policy change in the location of public secondary schools,
+in order to, first, understand the extent of spatial inequalities in
+school accessibility by public transport, and second, estimate the
+effect of a policy aiming at concentrating public secondary education
+provision on school accessibility. We find that closing down schools in
+areas with lower than average provision is highly regressive: it
+negatively impacts students with the lowest accessibility levels,
+accentuating existing inequalities.
 
@@
 distance (Euclidean and network), barriers and different functions
-representing distance decay.
+representing distance decay. This early work made the distinction
+between accessibility indices that apply to zones or single `desire
+lines': ``relative accessibility is defined as a measure of the effort
+of overcoming spatial separation between two points, while the integral
+accessibility is defined as a measure of the effort of overcoming
+spatial separation between a point and all other points within an area''
+(Allen, Liu, and Singer 1993). In subsequent works, the measurement of
+job accessibility by particular transport modes has received vast
+attention in the literature (Geurs and Van Wee 2004), while studies on
+access to education has largely focused on the consequences of mode
+choice on socio-economic indicators and school outcomes (see for
+instance Asahi (2014), Falch, Lujala, and Str{ø}m (2013) and Dickerson
+and McIntosh (2012) on the impact of accessibility on school outcomes,
+and Andersson, Malmberg, and {Ö}sth (2012) and Easton and Ferrari (2015)
+on the effect of changes in travel-to-school patterns in developed
+economies).
 
-This early work made the distinction between accessibility indeces that
-apply to zones or single `desire lines': ``relative accessibility is
-defined as a measure of the effort of overcoming spatial separation
-between two points, while the integral accessibility is defined as a
-measure of the effort of overcoming spatial separation between a point
-and all other points within an area'' (Allen, Liu, and Singer 1993).
+Our index is inspired by the index developed by Shen (1998) in the
+context of job accessibility. The main innovative element with respect
+to previous proposed measures of school accessibility is that it takes
+into account both the `supply' and `demand' for schooling in each area.
+Concretely, it is based on the idea that in each local area, there is a
+certain amount of students in (secondary) school age, and a certain
+quantity of (public) school seats available. As areas are part of a
+city, each local area is also subject to the inflow of potential
+students from other areas, as well of the outflow of students to other
+areas. The magnitude of the net flow will depend on the travel distance
+between all other areas and the area in question, which here we consider
+to be the public transport commuting time between areas. The index thus
+captures the fact that students compete for school seats, which are
+limited in number, and distributed unequally across space. The inclusion
+of this competitive element to school accessibility, akin to that in job
+access, highlights the fact that under certain educational systems, the
+access to opportunities is mediated by competition.
 
-\section{Method: measuring school
+Our work contributes to the literature on transport-related social
+exclusion (Jones and Lucas 2012) (Lucas et al. 2016) by providing a
+quantitative way to assess multiple dimensions into a single measure.
+Education and public transport are both essential public services, and
+thus are part of the set of fundamental and universal human rights. Lack
+of sufficient access to these services limits the way in which
+individuals use can their capacities and exersise their rights in a
+context of equal opportunities (Gomide 2003). Our index provides a
+benchmark to compare access inequalities in countries with unequal
+provision of public services (Gomide 2006), (Pacione 1989). Considering
+contestation as the process through which social groups mobilize in an
+organized way in order to impede the implementation of unwanted
+policies, or force the negotiation of new conditions, our index allows
+embedded inequalities to be better understood.
+
+After this introduction, we detail our data sources and definitions in
+Section 2. Section 3 describes the area of study and presents some
+preliminary findings. Section 4 describes the proposed methods for
+measuring school accessibility. Section 5 presents the results for the
+SPMR, and finally, Section 6 concludes.
+
+\section{2. Data}\label{data}
+
+z school accessibility index, we use data from three
+different sources. The first one is the 2008 School Census of the
+Brazilian National Institute of Educational Studies and Surveys (INEP),
+which provides information on the (universe) of public educational
+institutions, including a unique identifier and the number of students
+enrolled in secondary education. We use the coordinates contained in the
+database provided by the Center for Metropolitan Studies CEM
+(\textit{Centro de Estudos da Metrópole}) for 2001 to geolocate the
+schools in 2008 using the unique identifier. In order to include new
+schools built between 2001 and 2008, we used a dataset provided by the
+Fundação SEADE (São Paulo State Agency for Statistical Analysis)
+containing the goecoded location for all state administration schools at
+2008. The final dataset set was obtained merging this two data source
+using the standard ID used by the INEP. We geo-localized a total of
+4,612 public schools in the SPRM.
+
+Our second source is information the 2010 Population Census of Brazil,
+compiled and freely distributed by the Brazilian Institute of Statistics
+(IBGE). We aggregate the data by Area de Ponderacão Espacial (AEP)
+areas, a spatial unit defined for surveying purposes by IBGE. There is a
+total of 633 areas in the SPMR. IBGE provides the digital networks
+containing the boundaries of these areas. From the census microdata, we
+have data on the number of inhabitants by age for each enumeration area.
+The number of adolescents in secondary school age (ages 15-18) is
+1'216.611 (approximately 13 percent of the population).
+
+The final source of information was derived from the Google Distance
+Matrix API, which provided travel times and distances by public
+transport. This was implemented using an interface to the API by means
+of the R package \textbf{stplanr} (Lovelace et al. 2016). To provide an
+example of how the method works, a reproducible example is illustrated
+below.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\KeywordTok{library}\NormalTok{(stplanr)}
+\NormalTok{o =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.31813}\NormalTok{, -}\FloatTok{23.38584}\NormalTok{) }\CommentTok{# coordinates of origin}
+\NormalTok{d =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.34078}\NormalTok{, -}\FloatTok{23.37746}\NormalTok{) }\CommentTok{# coordinates of destination}
+\KeywordTok{dist_google}\NormalTok{(}\DataTypeTok{from =} \NormalTok{o, }\DataTypeTok{to =} \NormalTok{d, }\DataTypeTok{mode =} \StringTok{"transit"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##   distances duration
+## 1      2962     2118
+\end{verbatim}
+
+The code above takes an origin (\texttt{o}) and destination (\texttt{d})
+and finds the travel time and distance. This is reported as 2,962 metres
+and 2,118 seconds (35 minutes) respectively. To ensure that the result
+was relevant to school travel, the times were calculated for arrival
+time at 9am on a weekday. The function was repeated for all
+origin-destination pairs. To ensure the correct OD data was being
+generated, the results were plotted on an interactive map (Figure 2).
+The results show the high variability of route distance versus Euclidean
+distance (`circuity'), providing further evidence of inequalities in
+public transport provision in the city (Figure 3).
+\begin{figure}
+\includegraphics[width=1\linewidth]{od-map} \caption{Visual representation of the 12,697 OD pairs routed through the Google Distance Matrix API on an interactive map in RStudio, an open source data analysis platform.}\label{fig:unnamed-chunk-4}
+\end{figure}
+\begin{figure}
+\includegraphics[width=1\linewidth]{public-vs-euclid-dists} \caption{Euclidean (straight line) vs public transport distances resulting from routing the 12,697 OD pairs of interest through the Google Distance Matrix API.
+The colours represent the interquantile range of the Euclidean distances travelled to school.}\label{fig:euclid}
+\end{figure}
+
+\section{3. Area of study}\label{area-of-study}
+
+The São Paulo Metropolitan Region (SPMR) is a large metropolis located
+in the South West of Brazil. It extends for over 7,700 squared
+kilometers and groups 38 different municipalities besides the
+municipality of São Paulo itself. The estimated population in 2010 was
+close to 19.5 million in 2010. The integrated transport system, to which
+users can access through an electronic card (Bilhete Único) is composed
+of a railway and bus network. The railway network comprises 78.4
+kilometers in five subway lines and six suburban railway lines which
+provide less frequent and slower service than the subway. Some of the
+bus network operates on dedicated bus corridors, which amounted to 500
+km in 2015. As can be seen in Figure 1, the richer areas of the SPMR are
+better served by public transport, while an important number of lower
+income areas is under-served.
+
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{income_pc} \caption{Public transport system and income per capita in São Paulo, 2010.}\label{fig:inc-trans}
+\end{figure}
+
+Regarding the educational system, since the constitution of 1946,
+primary education is compulsory. In 2005, the legislation changed to
+increase the duration of the primary education, going from eight to nine
+years. In 2010, all regions finished the transition. In 2009, a
+constitutional amendment imposed compulsory education for the primary
+and secondary levels. The constitution imposes that this resolution
+should be fully implemented by 2016. By law, the state/municipalities
+should guaranty the adequate provision (100\% coverage of the demand).
+There are national and state educational funds that are distributed
+among the schools according to the number of enrolled students and
+existence of facilities (library, laboratories, etc.).
+
+The structure of the educational system is composed by both public and
+private schools. The public schools are divided in the following levels:
+Pre school (\textit{clase de alfabetização}) -- 4-5 year-olds (municipal
+administration); Fundamental (\textit{ensino fundamental}) comprises
+nine years, and the age range is usually 6-14 year-olds (municipal
+administration); Secondary (\textit{ensino médio}) lasts a minimum of
+three years, and the age range is 15-18 year-olds (state
+administration); and Post-secondary (\textit{ensino superior}) (state
+and federal administration).
+
+According to the School Census (\textit{Censo Escolar}) 2010\footnote{see
+  \url{http://portal.inep.gov.br/basica-censo} for details on the School
+  Census}, there are 10,251 schools in the SPMR, out of which 54\% are
+public (22\% state / 32\% municipal) and 46\% are private.
+Geographically, students are assigned to the closest school to their
+place of residency or their parent's working place. For secondary
+school, the student can ask for a vacancy in a different school after
+providing a valid justification and under the condition of the
+availability of vacancy.
+
+The municipality is responsible for the organization of the pre-school
+and fundamental level. Poorer municipalities in SPMR have, in general,
+lower quality in terms of teachers and facilities. All the secondary
+schools are administrated by the state, nevertheless, differences exists
+depending of the location in the SPMR. In general, schools in peripheral
+locations show lower quality scores.
+
+As can be seen by comparing Figures 4 and 5, with Figure 1, students in
+the SPMR disproportionally attend private secondary schools in higher
+income areas, while the opposite is true for public secondary schools.
+\begin{table}[]\centering \caption{Modal split by type of secondary school}
+\begin{tabular}{lcc} \hline
+\multicolumn{3}{c}{\textbf{Secondary Public School Students}}  \\  \hline
+ & Percent & Duration (min)  \\ \hline
+Public & 23 & 36.7 \\
+Private & 9 & 15.8 \\
+Active & 68 & 15.7 \\ \hline
+\multicolumn{3}{c}{\textbf{Secondary Private School Students}}  \\  \hline
+Public & 28 & 35.7 \\ 
+Private & 47 & 17.3 \\
+Active & 25 & 13.9 \\\hline
+\multicolumn{3}{l}{\small Public transport includes bus, metro and railway.} \\
+\multicolumn{3}{l}{\small Active transport includes walking and cycling.}  
+\end{tabular}
+\end{table}
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{med_pu} \caption{The home location of students attending public secondary schools students in the SPMR.}\label{fig:unnamed-chunk-5}
+\end{figure}\begin{figure}
+\includegraphics[width=0.7\linewidth]{med_pr} \caption{The home location of students attending private secondary schools students in the SPMR.}\label{fig:unnamed-chunk-6}
+\end{figure}
+
+As Figure 6 shows, enrolment in public secondary schools is more
+dispersed spatially, as in principle public schools are not clearly
+geographically concentrated within the region. There is, however, an
+important degree of variation in the school seats in each sub-area,
+ranging from zero to 7,754.
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{enrol} \caption{Enrollment in public secondary schools in the SPMR.}\label{fig:unnamed-chunk-7}
+\end{figure}
+Table 1 shows the model split by type of secondary school
+student.\footnote{The source of this information is the 2007
+  Origin-Destination Household Travel Survey (O-D Survey), carried out
+  by the São Paulo Metropolitan Transport Authority Mêtro. We focus on
+  trips made by secondary school students with an indicated motive
+  ``education'' at the destination. The total number of trips (excluding
+  missing values) was 11,845. Of these, 5.5 percent were multimodal. In
+  those cases, we assign the mode of the trip leg with the largest
+  duration. We add all trip leg durations (in minutes) to obtain the
+  total trip duration.} Clearly, secondary school students attending
+private schools disproportionally travel by car, compared to public
+school students (9\% versus 47\%), even though higher income areas are
+better served by public transport. This is in line with the findings of
+S{á} et al. (2015). Around 67\% of public school students commute to
+school by active travel (mostly walking, since the percentage of biking
+is still relatively small), while 23\% commute by public transport,
+which has much longer mean duration than other modes. Gomide (2003),
+using data from the 1997 SPMR Origin-Destination survey, reports that
+the motive ``education'' explained at least 60\% of trips in the SPMR of
+individuals earning up to one minimum wage a month. This proportion
+decreases as income increases, as other motives such as work and
+leisure become more relevant.
+
+Although the mean duration of travel to school remained stable over the
+period 1997-2007 (S{á} et al. 2015), travel-to-school times could
+increase as a result of several factors relevant to the SPMR, including
+sub-urbanization of particular demographic groups, such as those with
+higher income; increase in the size of schools leading to an increase in
+school's catchment area; less than proportional increase in schools in
+high population growth areas with poor connectivity; and institutional
+changes towards more flexible school choice (Easton and Ferrari 2015).
+
+\section{4. Method: measuring school
 accessibility}\label{method-measuring-school-accessibility}
 
-We calculate three different accessibility measures for different modes:
-cumulative, gravitiy-based, and competitive.
+To define school accessibility, we start with the broad concept of
+accessibility, as used in the transport literature. According to Geurs
+and Van Wee (2004) (p.~128) accessibility is ``the extent to which
+land-use and transport systems enable (groups of) individuals to reach
+activities or destinations by means of a (combination of) transport
+mode(s).'' Extended to our case, school accessibility then measures the
+extent to which the existing built-environment and transport facilities
+enable children and adolescents in school age to reach schools.
 
-The cumulative-opportunity measure for mode \(M\) (Boisjoly and
-El-Geneidy, 2016) is defined as:
+Ideally, a school accessibility measure would include information on the
+home location of every student and the location of every school they
+could potentially attend (as in Andersson, Malmberg, and {Ö}sth (2012)),
+and some measure of commuting costs for any chosen mode (in terms of
+time and money). Given that this level of detail is not available in our
+data, we rely on area aggregates, implicitly assuming that the measure
+at the centroid of the area applies homogeneously within the area.
 
-\[ CO_{i}^M= \sum_{j=1}^n(S_{j}f(C_{ij}) \]
+Given these limitations, we first consider a cumulative-opportunity
+measure for mode \(M\) (Boisjoly and El-Geneidy 2016) is defined as:
+
+\[ CO_{i}^M= \sum_{j=1}^nO_{j}f(C_{ij}) \]
 
@@
 
-This measure counts the number of schools available from one area within
-a certain travel time by mode \(M\) threshold. \(C_{ij}\) is the travel
-cost (measured in time) between the centroid of zone i and the centroid
-of zone j.
+This measure counts the number of ``opportunities'' \(O\), in our case
+schools available, from one area within a certain travel time threshold
+by mode \(M\). \(C_{ij}\) is the travel cost (measured in time) between
+the centroid of zone \(i\) and the centroid of zone \(j\), and
+\(f(C_{ij})\) is a weight function.
 
-Gravity-based accessibility measure
+This measure takes into account the spatial distribution of schooling
+opportunities, but not the local demand for schooling. This is
+particularly relevant for our analysis, since it could be the case that
+some areas are disproportionally served with respect to the number of
+potential students living within a certain travel distance. In order to
+assess the mismatch between the demand and supply for schooling, we use
+the sum of students in schools in each area (supply), and the sum of
+individuals within the school grade age-group living in each area
+(demand) in the following competitive accessibility measure (Shen 1998):
 
-In spite of its simplicity, Boisjoly and El-Geneidy (2016) have found
-that the cumulative-opportunity measure (using a 45 minute threshold) is
-highly correlated with a gravity-based accessibility measure
-(({\textbf{???}}), ({\textbf{???}})), defined as:
+\[ CA_{i}^M= \sum_{j=1}^n\frac{O_{j}f(C_{ij})}{\sum_{k=1}^n(T_{k}f(C_{kj}))}\]
 
-\[ A_{i}^M= \sum_{j=1}^n(S_{j}e^{\beta*C_{ij}}) \]
+Where the numerator discounts the number of school seats in area \(j\),
+\(O_{j}\), by how far area \(i\) is from area \(j\) using the same
+function as before, and the denominator discounts the number of teens in
+secondary school age living in zone \(k\), \(T_{k}\), by how far they
+are from area \(j\). In other words, the numerator counts how many
+schools seats can be reached from an area, while the denominator counts
+how many teens can potentially reach the same area. In this way, the
+discounted number of school seats places at each area is divided by the
+potential students available to fill those places, and then summed in
+order to obtain a single accessibility measure for each area.
 
-Where \(A_{i}^M\) is the gravity accessibility at the centroid of area
-\(i\) to all schools at area \(j\) using mode \(M\), and \(\beta\) is a
-negative exponential cost function. The cost function, based on a
-negative exponential decay curve, includes reported number of school
-trips in the 2007 Origin-Destination Survey. (Here we can also use
-information from the school census on number of students per school to
-know how many students the area attracts). The proposed accessibility
-measure captures the fact that more proximate schools (S) are more
-attractive than those located further away.
+We calculate both measures for commutes to school by public transport.
+We set a threshold of 30 minutes based on the mean average travel times
+by this mode (see Table 1). We assign a minimum time of zero for
+commutes within the same area (i.e., schools within the same area are
+added to the index in all cases).
 
-Competitive accessibility measure
+\section{5. Results}\label{results}
 
-The measures proposed above take into account the sptail distribution of
-schooling opportunities, but not the local demand for schooling. This is
-particularly relevant for our equity analysis, since it could be the
-case that higher areas are disproportionally served with respect to the
-number of potentail students living within a certain travel distance,
-whereas the opposite holds for lower-income areas. In order to assess
-the mismatch between the demand and supply for schooling, we use the sum
-of students in schools in each area (the demand), and the sum of
-individuals within the school grade age-group living in each area (the
-supply) in the following competitive accessibility measure, first
-proposed by ({\textbf{???}}) and adapted by El-Geneidy et al (2015):
+Figure 7 shows the spatial distribution of the cumulative-opportunity
+measure. The index has a mean value of 3,777, which indicates that for
+teens residing in the average area, there are 3,777 school seats that
+can be reached within a journey by public transport of up to 30 minutes.
+The standard deviation is 3,485 indicates large variation in the level
+of accessibility across areas.
 
-\[ CA_{i}^M= \sum_{j=1}^n\frac{P_{j}-f(C_{ij})}{\sum_{k=1}^n(Y_{k}-f(C_{kj}))}\]
+As can be seen on the Figure 7, areas with low accessibility include not
+only the expanded financial business district located on the central
+south west area, where higher-income teens mostly attend private school
+(see Figures 1 and 5), but also areas with deficient public transport
+provision, such as some north and south eastern areas.
 
-Where the numerator discounts the number of pupils in area \(j\)
-\(P_{j}\) by how far area \(i\) is from area \(j\) using the same
-function as before, and the denominator discounts the number of students
-(young adults) living in zone \(k\) \(Y_{k}\) by how far they are from
-area \(j\). In this way, the discounted number of study places at each
-area is divided by the potential students available to fill those
-places, and then summed in order to obtain a single accessibility
-measure for each area \(i\).
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{cum_index} \caption{Cumulative opportunity index}\label{fig:unnamed-chunk-8}
+\end{figure}
 
-In order to assess equality in access to schooling, we stratify our
-population of students in two student groups (we could do it for more
-income groups): the students living with a head of household who earns 1
-minimum wage or less, and those living with a head of household who
-earns more than 1 minimum wage. The accessibility measures are based on
-student's travel times by different modes using data from the 2007
-Origin-Destination survey. Alternatively, for the case of public
-transport, wet use actual transit times in 2015 for comparison (bilhete
-unico data?).
+Figure 8 shows the spatial distribution of the competitive accessibility
+measure. Although it largely follows the distribution of the cumulative
+index, it corrects the value in some areas based on the place of
+residence of students across the SPRM. To interpret the value of the
+index, recall that if all teens in an area attended secondary schools in
+public secondary schools in their area (i.e., no student would attend
+private school or travel outside her own area to attend school), the
+index would take the value of 1. Then, in areas with a value lower than
+one, the provision of public schools and public transport is such that
+there is local under-provision, in the sense that the number of schools
+seats that can be reached from the area is smaller than the number of
+teens that can potentially reach the area (the opposite is true for
+values higher than one). Note that some relatively close by areas,
+especially in the eastern part of the city, have striking differences in
+the level of accessibility. In some cases, an area can have twice the
+level of accessibility than the most proximate area. This can be due to
+differences in the local provision of public schools, public transport
+or the number of students living within the reach of the areas.
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{comp_index} \caption{Competitive accessibility index}\label{fig:unnamed-chunk-9}
+\end{figure}
+To understand the effects of changes in public school provision on
+school accessibility, we conduct a comparative static analysis using our
+competitive index. We compare the changes in school-accessibility by
+accessibility quantiles before and after the implementation of a policy
+that reallocates students attending school in areas with lower than
+average provision to schools in areas with higher than average
+provision. The experiment mimics a policy that aims at increasing scale
+economies in the provision of public education by centralizing provision
+in larger educational establishments, which is the rationale behind the
+policy proposal made by the São Paulo state government (Ortellado 2015).
 
-\section{Results}\label{results}
+To implement the policy experiment, we assign students in areas with
+lower than average enrolment to the next area with higher than average
+enrolment, while keeping the distribution of teens and public transport
+constant. In other words, total enrolment is the same before and after
+the change, but the distribution of public secondary schools increases
+in areas where it was already higher than average, and disappears in
+areas where it was lower than average. The change is equivalent to
+closing down 1,691 schools, or reallocating 163,231 students. Although
+seemingly extreme, it is worth noting again that the proposal of the São
+Paulo state government was estimated to impact 300,000 secondary
+students attending public school (Ortellado 2015). We then compare the
+changes in school accessibility for different quartiles before and after
+this change. Table 2 summarizes the results.
 
-\begin{itemize}
-\itemsep1pt\parskip0pt\parsep0pt
-\item
-  Descriptive statistics about modal choice and travel time by AEPs and
-  level of income
-\item
-  Spatial Distribution of Schools and Transit-Dependent Students at the
-  AEP Level
-\end{itemize}
+\begin{longtable}[c]{@{}lrrr@{}}
+\caption{Mean accessibility before and after
+redistribution}\tabularnewline
+\toprule
+Quantile & Pre-change & Post-change & Change (\%)\tabularnewline
+\midrule
+\endfirsthead
+\toprule
+Quantile & Pre-change & Post-change & Change (\%)\tabularnewline
+\midrule
+\endhead
+1st & 0.446 & 0.350 & -21.566\tabularnewline
+2nd & 0.844 & 0.822 & -2.592\tabularnewline
+3rd & 1.843 & 2.124 & 15.213\tabularnewline
+4th & 4.280 & 5.483 & 28.099\tabularnewline
+\bottomrule
+\end{longtable}
 
-\section{Discussion}\label{discussion}
+\newpage
 
-\section{Conclusion}\label{conclusion}
+The policy reduces further the accessibility level of areas in the first
+and second quantiles of accessibility. Areas in the first quantile
+experience an average reduction of 21\% in school accessibility levels
+as a result of the changes in the spatial distribution of schools. In
+the mean time, areas that had high levels of accessibility to start
+with, experience an increase. The results suggest that concentrating the
+provision of public secondary schools accentuates the already large
+differences in accessibility levels across areas.
+
+\section{6. Conclusion and discussion}\label{conclusion-and-discussion}
+
+The policy experiment results suggest that a policy that concentrates
+the provision of public schools spatially is regressive. Young people
+living in areas with the lowest levels of accessibility experience the
+largest negative impact from the policy, compounding issues of low local
+schooling provision, low access by public transport and/or high
+competition to access good schools. The policy would increase commuting
+times for students residing in areas with low accessibility. This would
+likely have further negative knock-on effects in terms of attendance,
+drop-out likelihood and performance.
+
+Although the simulated scenario represents a rather extreme change in
+the distribution of public schooling, the way in which students with
+different levels of accessibility are impacted remains, regardless of
+the level of concentration. In the absence of a public transport subsidy
+for school commutes, students who experience a substantial increase in
+travel-to-school times may be forced to change their commute mode from
+an active (and free of charge mode) to a paid-by mode, with the
+consequent increase in their monetary spent in commuting to school. The
+most negative effect is then felt by low-incomes students living in
+areas with low school accessibility, for whom the marginal increase in
+commuting costs represents a much larger share in their budget.
+
+The provision of a public transport subsidy for school commutes is
+unlikely to compensate students at the bottom of the income and
+accessibility ladder if the increase in commuting times has consequences
+in terms of motivation to stay in school and performance. In this sense,
+a ``trade-off'' logic where less schooling provision is compensated by
+public transport subsidies does not operate when existing inequalities
+in provision are large.
+
+Our study could be further extended to account for other modes of school
+commute, such as active travel. It would benefit from data with higher
+spatial resolution to capture within-area inequalities, which may be
+substantial but currently hidden away in area aggregates. A future
+extension of the index could also incorporate measures of school
+quality. Furthermore, it would be informative to calculate the proposed
+indices for other contexts and different time periods, to understand the
+extent to which the regressive nature of school agglomeration is a
+universal finding or specific to the SPMR.
+
+We conclude that extending the concept of local accessibility indicators
+to education can help to both contest and constructively tackle embedded
+social inequalities. Contestation can be seen as a powerful tool to
+prevent the implementation of regressive policies that threaten to
+accentuate already large disparities. Quantitative measures such as
+those proposed here can help demonstrate how multiple dimensions of
+inequality amplify inequalities, via a single interpretable measure. The
+methods demonstrated could be useful for policy makers and campaigners
+to negotiate policies with a highly sensitive social component and
+evident short and long-term consequences for those affected.
 
@@
 
+Andersson, Eva, Bo Malmberg, and John {Ö}sth. 2012. ``Travel-to-School
+Distances in Sweden 2000--2006: Changing School Geography with Equality
+Implications.'' \emph{Journal of Transport Geography} 23. Elsevier:
+35--43.
+
+Asahi, Kenzo. 2014. ``The Impact of Better School Accessibility on
+Student Outcomes.'' The London School of Economics; Political Science,
+SERC Discussion Paper.
+
+Boisjoly, Genevi{è}ve, and Ahmed El-Geneidy. 2016. ``Daily Fluctuations
+in Transit and Jobs Availability: A Comparative Assessment of
+Time-Sensitive Accessibility Measures.'' \emph{Journal of Transport
+Geography}, no. 52: 73--81.
+
+Dickerson, Andy, and Steven McIntosh. 2012. ``The Impact of Distance to
+Nearest Education Institution on the Post-Compulsory Education
+Participation Decision.'' \emph{Urban Studies}. SAGE Publications,
+0042098012455717.
+
+Easton, Sue, and Ed Ferrari. 2015. ``Children's Travel to School---the
+Interaction of Individual, Neighbourhood and School Factors.''
+\emph{Transport Policy} 44. Elsevier: 9--18.
+
+Falch, Torberg, P{ä}ivi Lujala, and Bjarne Str{ø}m. 2013. ``Geographical
+Constraints and Educational Attainment.'' \emph{Regional Science and
+Urban Economics} 43 (1). Elsevier: 164--76.
+
+Geurs, Karst T, and Bert Van Wee. 2004. ``Accessibility Evaluation of
+Land-Use and Transport Strategies: Review and Research Directions.''
+\emph{Journal of Transport Geography} 12 (2). Elsevier: 127--40.
+
+Gomide, Alexandre de {Á}vila. 2003. ``Transporte Urbano E Inclusão
+Social: Elementos Para Políticas Públicas.'' Instituto de Pesquisa
+Econômica Aplicada (Ipea).
+
+---------. 2006. ``Mobilidade Urbana, Iniqüidade E Políticas Sociais.''
+Instituto de Pesquisa Econômica Aplicada (Ipea).
+
 Ingram, David R. 1971. ``The Concept of Accessibility: A Search for an
@@
 
+Jones, Peter, and Karen Lucas. 2012. ``The Social Consequences of
+Transport Decision-Making: Clarifying Concepts, Synthesising Knowledge
+and Assessing Implications.'' \emph{Journal of Transport Geography} 21.
+Elsevier: 4--16.
+
+Lovelace, Robin, Richard Ellison, Barry Rowlingson, and Nick Bearman.
+2016. \emph{Stplanr: Sustainable Transport Planning}.
+\url{https://github.com/ropensci/stplanr}.
+
+Lucas, Karen, John Bates, Jos{é} Moore, and Juan Antonio Carrasco. 2016.
+``Modelling the Relationship Between Travel Behaviours and Social
+Disadvantage.'' \emph{Transportation Research Part A: Policy and
+Practice} 85. Elsevier: 157--73.
+
+Ortellado, Pablo. 2015. ``Brazil's Students Occupy Their Schools to Save
+Them.'' \emph{The New York Times}, 15 December 2015. Available:
+http://www.nytimes.com/2015/12/16/opinion/brazils-students-occupy-their-schools-to-save-them.html?\textsubscript{r}=1
+{[}Last accessed: 25 April 2016{]},
+
+Pacione, Michael. 1989. ``Access to Urban Services---the Case of
+Secondary Schools in Glasgow.'' \emph{The Scottish Geographical
+Magazine} 105 (1). Taylor \& Francis: 12--18.
+
+S{á}, Thiago H{é}rick de, Leandro Martin Totaro Garcia, Gr{é}gore Iven
+Mielke, Fabiana Maluf Rabacow, and Leandro F{ó}rnias Machado de Rezende.
+2015. ``Changes in Travel to School Patterns Among Children and
+Adolescents in the São Paulo Metropolitan Area, Brazil, 1997--2007.''
+\emph{Journal of Transport \& Health} 2 (2). Elsevier: 143--50.
+
+Shen, Qing. 1998. ``Location Characteristics of Inner-City Neighborhoods
+and Employment Accessibility of Low-Wage Workers.'' \emph{Environment
+and Planning B: Planning and Design} 25 (3). SAGE Publications: 345--65.
+
 \end{document}

### 2017-02-26 paper.tex
Reword data section

@@
 
-z school accessibility index, we use data from three
-different sources. The first one is the 2008 School Census of the
-Brazilian National Institute of Educational Studies and Surveys (INEP),
-which provides information on the (universe) of public educational
+The school accessibility index was created using data from three main
+sources: a 2008 School Census; school location data; and geographical
+route data generated by the Google Matrix API.
+
+The School Census data was provided by the
+Brazilian National Institute of Educational Studies and Surveys (INEP).
+The dataset contains information on all public educational
 institutions, including a unique identifier and the number of students
-enrolled in secondary education. We use the coordinates contained in the
+enrolled in secondary education. School coordinates contained in a
 database provided by the Center for Metropolitan Studies CEM
-(\textit{Centro de Estudos da Metrópole}) for 2001 to geolocate the
-schools in 2008 using the unique identifier. In order to include new
+(\textit{Centro de Estudos da Metrópole}) for 2001 were used to
+geolocate the schools in 2008 using the unique identifier.
+To include new
 schools built between 2001 and 2008, we used a dataset provided by the
@@
 containing the goecoded location for all state administration schools at
-2008. The final dataset set was obtained merging this two data source
+2008. The this dataset set was obtained merging this two data source
 using the standard ID used by the INEP. We geo-localized a total of
@@
 
-Our second source is information the 2010 Population Census of Brazil,
+The second source was the 2010 Population Census of Brazil,
 compiled and freely distributed by the Brazilian Institute of Statistics
-(IBGE). We aggregate the data by Area de Ponderacão Espacial (AEP)
-areas, a spatial unit defined for surveying purposes by IBGE. There is a
-total of 633 areas in the SPMR. IBGE provides the digital networks
+(IBGE). We aggregated the data by 633 administrative zones in the
+SPMR: Area de Ponderacão Espacial (AEP)
+areas, a spatial unit defined for surveying purposes by IBGE.
+IBGE provides geographical datasets
 containing the boundaries of these areas. From the census microdata, we
-have data on the number of inhabitants by age for each enumeration area.
-The number of adolescents in secondary school age (ages 15-18) is
-1'216.611 (approximately 13 percent of the population).
+have data on the number of inhabitants by age for each enumeration area,
+which totals more than one million secondary school students:
+1,216,611 people of secondary school age (15 - 18 years) were recorded
+in the Census dataset (approximately 13 percent of the population).
 

### 2017-02-26 paper.tex
Add Singleton reference, explain Google Matrix API

@@
 public transport are based on actual commuting times obtained through
-the Google Matrix Distance API. We calculate these measures for 633
+use of routing algorithms provided on-line through an application
+programming interface (API).
+
+For this work we used the Google Matrix Distance API, a multi-modal
+real-time travel data provider. 
+The API was used in preference to other options,
+such as Routino, described in a paper by Singleton (2014)
+because its
+ease of use and performance. The Google Distance Matrix
+API returns
+route distance, price and times for journeys between origins and
+destinations, provided either as Longitude/Latidue pairs or
+as text strings to be 'geocoded' (converted to geographical location).
+
+We calculate these measures for 633
 areas within the SPMR. We use the competitive measure to simulate the
@@
 
+Singleton, Alex.
+2014.
+``A GIS approach to modelling CO2 emissions associated with the pupil-school commute.'' \emph{International Journal of Geographical Information Science} 28, 256–273.
+doi:10.1080/13658816.2013.832765
+
 Shen, Qing. 1998. ``Location Characteristics of Inner-City Neighborhoods

### 2017-02-26 paper.tex
Finalise description of Google Matrix API

--- 2017-02-26 paper.tex
@@
 public transport are based on actual commuting times obtained through
-the Google Matrix Distance API. We calculate these measures for 633
+use of routing algorithms provided on-line through an on-line service.
+
+For this work we used the Google Matrix Distance API
+(API stands for `application programming interface' and is used for
+compter-computer communication over the web), a multi-modal
+real-time travel data provider (see https://developers.google.com/maps/documentation/distance-matrix/). 
+The API was used in preference to other options,
+such as Routino, described in a paper by Singleton (2014),
+and the OpenStreetMap Routing Machine (OSRM)
+because its
+ease of use (requiring no new software to be installed locally)
+and performance: the Google Distance Matrix
+API returns
+route distance, price and times for journeys between origins and
+destinations, provided either as Longitude/Latidue pairs or
+as text strings to be 'geocoded' (converted to geographical location).
+
+We calculate these measures for 633
 areas within the SPMR. We use the competitive measure to simulate the

### 2017-04-02 paper_v2.tex
Updated paper

--- 2017-02-26 paper.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\documentclass[3p,authoryear,preprint,review,12pt]{elsarticle} 
+\usepackage{lipsum}
+\makeatletter
+\def\ps@pprintTitle{%
+ \let\@oddhead\@empty
+ \let\@evenhead\@empty
+ \def\@oddfoot{}%
+ \let\@evenfoot\@oddfoot}
+\makeatother
+\usepackage{natbib,amssymb,amsmath,latexsym,makeidx,lscape,epic,overpic,graphicx,graphics,bm,pdflscape,rotating,multirow,bigdelim,color,epsfig,pgfpages} \usepackage{mathpazo}
+\usepackage{mathtools}
+\usepackage[portuguese, english]{babel}
 \usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-    \usepackage{xltxtra,xunicode}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
-  \newcommand{\euro}{€}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{{#1}}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{{#1}}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{{#1}}}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{{#1}}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{{#1}}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{{#1}}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{{#1}}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{{#1}}}
-\newcommand{\RegionMarkerTok}[1]{{#1}}
-\newcommand{\ErrorTok}[1]{\textbf{{#1}}}
-\newcommand{\NormalTok}[1]{{#1}}
+\usepackage[left]{eurosym}
+\usepackage{pdfsync,epstopdf,dblfloatfix,multicol,wrapfig,multirow}
+\usepackage{epstopdf}
+\usepackage{units}
+\epstopdfsetup{outdir=./}
+\usepackage[draft]{hyperref} % Drop [draft] to have references hyperlinked.
+\usepackage{float}
 \usepackage{longtable,booktabs}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\ifxetex
-  \usepackage[setpagesize=false, % page size defined by xetex
-              unicode=false, % unicode breaks when used with xetex
-              xetex]{hyperref}
-\else
-  \usepackage[unicode=true]{hyperref}
-\fi
-\hypersetup{breaklinks=true,
-            bookmarks=true,
-            pdfauthor={Ana I. Moreno-Monroy (Universitat Rovira i Virgili), Robin Lovelace (University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)},
-            pdftitle={The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region},
-            colorlinks=true,
-            citecolor=blue,
-            urlcolor=blue,
-            linkcolor=magenta,
-            pdfborder={0 0 0}}
-\urlstyle{same}  % don't use monospace font for urls
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\setcounter{secnumdepth}{0}
+\usepackage[toc,page]{appendix}
+\usepackage{url}
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
+\newcommand{\squeezeup}{\vspace{-2pt}}
+\setlength{\textfloatsep}{3pt plus 1.0pt minus 2.0pt}
+       \setlength{\abovecaptionskip }{-1ex}
 
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-  \title{The unequal distribution of access to education and transport: insights
-from the São Paulo Metropolitan Region}
-  \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-  \author{Ana I. Moreno-Monroy (Autonomous university of Barcelona), Robin Lovelace
-(University of Leeds) and Frederico R. Ramos (Fundação Getulio Vargas)}
-  \preauthor{\centering\large\emph}
-  \postauthor{\par}
-  \date{}
-  \predate{}\postdate{}
+%Standard settings for an Elsevier article. 
+\def\elsartstyle{\def\normalsize{\@setfontsize\normalsize\@xiipt{12.5}} \def\small{\@setfontsize\small\@xipt{13.6}} \let\footnotesize=\small
+   \def\large{\@setfontsize\large\@xivpt{18}} \def\Large{\@setfontsize\Large\@xviipt{22}}  \skip\@mpfootins = 18\p@ \@plus 2\p@ \normalsize }
 
@@
 
-\maketitle
+\begin{frontmatter}
+ \title{The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region}
 
-\section{Abstract}\label{abstract}
+  \author[uab]{Ana~I.~Moreno-Monroy}
+ \ead{anaisabel.moreno@uab.cat}
 
+  \author[uol]{Robin Lovelace}
+
+   \author[fgv]{Frederico R. Ramos}
+
+ \address[urv]{Applied Economics Department, Autonomous University of Barcelona}
+  \address[uol]{Intitute of Trasnport Studies, University of Leeds}
+ \address[fgv]{CEPESP, Fundação Getulio Vargas}
+
+\begin{abstract}
 In many large Latin American urban areas such as the São Paulo
-Metropolitan Region (SPMR), growing social and economic inequalities are
-embedded through high spatial inequality in the provision of education
-and transport systems. This paper sheds light on the transport-education
-inequality nexus with reference to a new school accessibility measure
-applied to the SPMR. Our index combines information on the spatial
-distribution of adolescents, the location of existing schools, and the
-travel infrastructure serving the school catchment area into a single
-measure. The index is used to measure school accessibility locally
-across 633 areas within the SPMR. We use the index to simulate the
-impact of a policiy aiming at increasing the centralization of public
-secondary education provision, and find that it negatively affects
-students with the lowest levels of accessibility. These results
-illustrate how existing inequalities are amplified by variable
-accessibility to schools across income groups and geographical space.
+Metropolitan Region (SPMR), growing social and economic inequalities are embedded through high spatial inequality in the provision of education and transport systems.
+This paper sheds light on the transport-education inequality nexus with reference to a new school accessibility measure applied to the SPMR.
+Our index combines information on the spatial distribution of adolescents, the location of existing schools, and the travel infrastructure serving the school catchment area into a single measure.
+The index is used to measure school accessibility locally across 633 areas within the SPMR.
+We use the index to simulate the impact of a policy aiming at increasing the centralization of public secondary education provision, and find that it negatively affects students with the lowest levels of accessibility.
+These results illustrate how existing inequalities are amplified by variable accessibility to schools across income groups and geographical space.
+\end{abstract}
 
-\bigskip
+\begin{keyword}
+Public transport, accessibility, schools, inequality, Latin America\\
+\end{keyword}
 
-\textit{This version: January 2017. Preliminary version. Please do not cite}
+\end{frontmatter}
 
+\newpage
 \section{Introduction}\label{introduction}
 
-Inequalities in educational and transport infrastructure are mutually
-reinforcing: the right to mobility is intrinsically linked to the right
-to education. Travel to school options are vital for ensuring a more
-equitable supply of educational opportunity to diverse groups.
-Conversely, poor accessibility to and from deprived areas can reinforce
-social inequalities, with long-term implications. Against this
-background, in this paper we propose a new way to measure school
-accessibility in local areas within an urban area, and apply it for the
-case of the São Paulo Metropolitan Region (SPMR).
+Inequalities in educational and transport infrastructure are mutually reinforcing: the right to mobility is intrinsically linked to the right to education.
+Travel to school options are vital for ensuring a more equitable supply of educational opportunity to diverse groups.Conversely, poor accessibility to and from deprived areas can reinforce social inequalities, with long-term implications.
+Against this background, in this paper we propose a new way to measure school accessibility in local areas within an urban area, and apply it for the case of the São Paulo Metropolitan Region (SPMR).
 
@@
 educational and transport infrastructure, and to study the impact of
-public policies on different socio-economic groups. Partly as a response
-to the pressures from the Free Pass Movement that started in March,
-2013, low income public school students residing in the SPMR gained
-access to a public transport subsidy in 2015. However, in October 2015,
-the São Paulo state government announced during a television interview
-that as part of a budgetary deficit reduction plan, dozens of secondary
-schools would be closed in 2016. The change in policy was estimated to
-affect more than 300,000 students, many whom would be placed in schools
-far from their homes. As a response, in the following months students
-occupied over 200 schools. Protesters scored some victories: the reform
-was postponed for one year, and the education secretary resigned
-(Ortellado 2015).
+public policies on different socio-economic groups.
+Partly as a response to the pressures from the Free Pass Movement that started in March, 2013, low income public school students residing in the SPMR gained access to a public transport subsidy in 2015.
+However, in October 2015, the São Paulo state government announced during a television interview that as part of a budgetary deficit reduction plan, dozens of secondary schools would be closed in 2016.
+The change in policy was estimated to affect more than 300,000 students, many whom would be placed in schools far from their homes.
+As a response, in the following months students occupied over 200 schools.
+Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (Ortellado 2015).
 
-Although this time the policy change was not carried out, in a context
-of high spatial inequalities in the provision of public transport and
-schooling, it is worth asking whether a public transport subsidy for
-students can compensate for the lower provision of public schools in
-some areas. In this dynamic and highly political context, new
-quantitative evidence can help shed light on the relationship between
-transport and educational inequalities and the extent to which they are
-mutually reinforcing. Moreover, the results should help design policies
-that are more inclusive and equitable by highlighting areas where the
+Although this time the policy change was not carried out, in a context of high spatial inequalities in the provision of public transport and schooling, it is worth asking whether a public transport subsidy for students can compensate for the lower provision of public schools in some areas.
+In this dynamic and highly political context, new quantitative evidence can help shed light on the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing.
+Moreover, the results should help design policies that are more inclusive and equitable by highlighting areas where the
 provision of schools, travel to school opportunities, or both, are
@@
 Our proposed competitive accessibility measure combines information
-about the place of residence of students, the spatial distribution of
-public schools, and public transport accessibility into a single
-measure. We start from a cumulative opportunity measure that counts the
-number of schools seats that can be reached within a 30 minute journey
-by public transport, and then build a competitive measure that takes
-into account the place of residency of teens. Travel-to-school times by
-public transport are based on actual commuting times obtained through
-the Google Matrix Distance API. We calculate these measures for 633
-areas within the SPMR. We use the competitive measure to simulate the
-impact of a policy change in the location of public secondary schools,
-in order to, first, understand the extent of spatial inequalities in
-school accessibility by public transport, and second, estimate the
-effect of a policy aiming at concentrating public secondary education
-provision on school accessibility. We find that closing down schools in
-areas with lower than average provision is highly regressive: it
+about the place of residence of students, the spatial distribution of public schools, and public transport accessibility into a single
+measure.
+We start from a cumulative opportunity measure that counts the number of schools seats that can be reached within a 30 minute journey by public transport, and then build a competitive measure that takes into account the place of residency of teens.
+Travel-to-school times by public transport are based on actual commuting times obtained through use of routing algorithms provided on-line through an on-line service.
+
+For this work we used the Google Matrix Distance API
+(API stands for `application programming interface' and is used for
+computer-computer communication over the web), a multi-modal
+real-time travel data provider (see \url{https://developers.google.com/maps/documentation/distance-matrix/}). 
+The API was used in preference to other options,
+such as Routino, described in a paper by Singleton (2014),
+and the OpenStreetMap Routing Machine (OSRM)
+because its
+ease of use (requiring no new software to be installed locally)
+and performance: the Google Distance Matrix
+API returns
+route distance, price and times for journeys between origins and
+destinations, provided either as Longitude/Latitude pairs or
+as text strings to be 'geocoded' (converted to geographical location).
+
+We calculate these measures for 633 areas within the SPMR.
+We use the competitive measure to simulate the impact of a policy change in the location of public secondary schools, in order to, first, understand the extent of spatial inequalities in school accessibility by public transport, and second, estimate the
+effect of a policy aiming at concentrating public secondary education provision on school accessibility.
+We find that closing down schools in areas with lower than average provision is highly regressive: it
 negatively impacts students with the lowest accessibility levels,
@@
 
-The first well-known attempt to define accessibility quantitatively was
-by Ingram (1971), which presented a range of measures related to
+The first well-known attempt to define accessibility quantitatively was by Ingram (1971), which presented a range of measures related to
 distance (Euclidean and network), barriers and different functions
@@
 between accessibility indices that apply to zones or single `desire
-lines': ``relative accessibility is defined as a measure of the effort
-of overcoming spatial separation between two points, while the integral
-accessibility is defined as a measure of the effort of overcoming
-spatial separation between a point and all other points within an area''
-(Allen, Liu, and Singer 1993). In subsequent works, the measurement of
-job accessibility by particular transport modes has received vast
-attention in the literature (Geurs and Van Wee 2004), while studies on
-access to education has largely focused on the consequences of mode
-choice on socio-economic indicators and school outcomes (see for
-instance Asahi (2014), Falch, Lujala, and Str{ø}m (2013) and Dickerson
-and McIntosh (2012) on the impact of accessibility on school outcomes,
-and Andersson, Malmberg, and {Ö}sth (2012) and Easton and Ferrari (2015)
-on the effect of changes in travel-to-school patterns in developed
-economies).
+lines': ``relative accessibility is defined as a measure of the effort of overcoming spatial separation between two points, while the integral accessibility is defined as a measure of the effort of overcoming spatial separation between a point and all other points within an area'' (Allen, Liu, and Singer 1993).
+In subsequent works, the measurement of job accessibility by particular transport modes has received vast attention in the literature (Geurs and Van Wee 2004), while studies on access to education has largely focused on the consequences of mode choice on socio-economic indicators and school outcomes (see for instance Asahi (2014), Falch, Lujala, and Str{ø}m (2013) and Dickerson and McIntosh (2012) on the impact of accessibility on school outcomes, and Andersson, Malmberg, and {Ö}sth (2012) and Easton and Ferrari (2015) on the effect of changes in travel-to-school patterns in developed economies).
 
 Our index is inspired by the index developed by Shen (1998) in the
-context of job accessibility. The main innovative element with respect
-to previous proposed measures of school accessibility is that it takes
-into account both the `supply' and `demand' for schooling in each area.
-Concretely, it is based on the idea that in each local area, there is a
-certain amount of students in (secondary) school age, and a certain
-quantity of (public) school seats available. As areas are part of a
-city, each local area is also subject to the inflow of potential
-students from other areas, as well of the outflow of students to other
-areas. The magnitude of the net flow will depend on the travel distance
-between all other areas and the area in question, which here we consider
-to be the public transport commuting time between areas. The index thus
+context of job accessibility.
+The main innovative element with respect to previous proposed measures of school accessibility is that it takes into account both the `supply' and `demand' for schooling in each area.
+Concretely, it is based on the idea that in each local area, there is a certain amount of students in (secondary) school age, and a certain quantity of (public) school seats available.
+As areas are part of a city, each local area is also subject to the inflow of potential students from other areas, as well of the outflow of students to other areas.
+The magnitude of the net flow will depend on the travel distance between all other areas and the area in question, which here we consider to be the public transport commuting time between areas.
+The index thus
 captures the fact that students compete for school seats, which are
-limited in number, and distributed unequally across space. The inclusion
-of this competitive element to school accessibility, akin to that in job
-access, highlights the fact that under certain educational systems, the
-access to opportunities is mediated by competition.
+limited in number, and distributed unequally across space.
+The inclusion of this competitive element to school accessibility, akin to that in job access, highlights the fact that under certain educational systems, the access to opportunities is mediated by competition.
 
@@
 quantitative way to assess multiple dimensions into a single measure.
-Education and public transport are both essential public services, and
-thus are part of the set of fundamental and universal human rights. Lack
-of sufficient access to these services limits the way in which
-individuals use can their capacities and exersise their rights in a
-context of equal opportunities (Gomide 2003). Our index provides a
-benchmark to compare access inequalities in countries with unequal
-provision of public services (Gomide 2006), (Pacione 1989). Considering
-contestation as the process through which social groups mobilize in an
-organized way in order to impede the implementation of unwanted
-policies, or force the negotiation of new conditions, our index allows
-embedded inequalities to be better understood.
+Education and public transport are both essential public services, and thus are part of the set of fundamental and universal human rights.
+Lack of sufficient access to these services limits the way in which individuals use can their capacities and exercise their rights in a context of equal opportunities (Gomide 2003).
+Our index provides a benchmark to compare access inequalities in countries with unequal provision of public services (Gomide 2006), (Pacione 1989).
+Considering contestation as the process through which social groups mobilize in an organized way in order to impede the implementation of unwanted policies, or force the negotiation of new conditions, our index allows embedded inequalities to be better understood.
 
-After this introduction, we detail our data sources and definitions in
-Section 2. Section 3 describes the area of study and presents some
-preliminary findings. Section 4 describes the proposed methods for
-measuring school accessibility. Section 5 presents the results for the
-SPMR, and finally, Section 6 concludes.
+After this introduction, we detail our data sources and definitions in Section 2.
+Section 3 describes the area of study and presents some preliminary findings.
+Section 4 describes the proposed methods for measuring school accessibility.
+Section 5 presents the results for the SPMR, and finally, Section 6 concludes.
 
-\section{2. Data}\label{data}
+\section{Data}\label{data}
 
-The school accessibility index was created using data from three main
-sources: a 2008 School Census; school location data; and geographical
-route data generated by the Google Matrix API.
+The school accessibility index was created using data from the following sources: 
+\begin{itemize}
+\item Demographic Census – 2010 – IBGE
+\item Origen – Destination Survey – 2007 – Metro
+\item School Census – 2008 – INEP
+\item Geocoded Schools – CEM/CEBRAP – 2001
+\item Geocoded Public Schools – SEADE – 2008
+\item Google Distance Matrix API 
+\end{itemize}
 
-The School Census data was provided by the
-Brazilian National Institute of Educational Studies and Surveys (INEP).
+The School Census data was provided by the Brazilian National Institute of Educational Studies and Surveys (INEP).
 The dataset contains information on all public educational
-institutions, including a unique identifier and the number of students
-enrolled in secondary education. School coordinates contained in a
-database provided by the Center for Metropolitan Studies CEM
-(\textit{Centro de Estudos da Metrópole}) for 2001 were used to
-geolocate the schools in 2008 using the unique identifier.
-To include new
-schools built between 2001 and 2008, we used a dataset provided by the
-Fundação SEADE (São Paulo State Agency for Statistical Analysis)
-containing the goecoded location for all state administration schools at
-2008. The this dataset set was obtained merging this two data source
-using the standard ID used by the INEP. We geo-localized a total of
-4,612 public schools in the SPRM.
+institutions, including a unique identifier and the number of students enrolled in secondary education.
+School coordinates contained in a database provided by the Center for Metropolitan Studies CEM (\textit{Centro de Estudos da Metrópole}) for 2001 were used to geolocate the schools in 2008 using the unique identifier.
+To include new schools built between 2001 and 2008, we used a dataset provided by the Fundação SEADE (São Paulo State Agency for Statistical Analysis) containing the geocoded location for all state administration schools at 2008.
+The this dataset set was obtained merging this two data source using the standard ID used by the INEP.
+We geo-localized a total of 4,612 public schools in the SPMR.
 
 The second source was the 2010 Population Census of Brazil,
-compiled and freely distributed by the Brazilian Institute of Statistics
-(IBGE). We aggregated the data by 633 administrative zones in the
-SPMR: Area de Ponderacão Espacial (AEP)
+compiled and freely distributed by the Brazilian Institute of Statistics (IBGE). We aggregated the data by 633 administrative zones in the SPMR: Area de Ponderacão Espacial (AEP)
 areas, a spatial unit defined for surveying purposes by IBGE.
-IBGE provides geographical datasets
-containing the boundaries of these areas. From the census microdata, we
-have data on the number of inhabitants by age for each enumeration area,
-which totals more than one million secondary school students:
-1,216,611 people of secondary school age (15 - 18 years) were recorded
-in the Census dataset (approximately 13 percent of the population).
+IBGE provides geographical datasets containing the boundaries of these areas.
+From the census microdata, we have data on the number of inhabitants by age for each enumeration area, which totals more than one million secondary school students: 1,216,611 people of secondary school age (15 - 18 years) were recorded in the Census dataset (approximately 13 percent of the population).
 
@@
 Matrix API, which provided travel times and distances by public
-transport. This was implemented using an interface to the API by means
-of the R package \textbf{stplanr} (Lovelace et al. 2016). To provide an
-example of how the method works, a reproducible example is illustrated
-below.
+transport.
+This was implemented using an interface to the API by means of the R package \textbf{stplanr} (Lovelace et al.
+2016).
+To provide an example of how the method works, a reproducible example is illustrated below.
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\KeywordTok{library}\NormalTok{(stplanr)}
-\NormalTok{o =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.31813}\NormalTok{, -}\FloatTok{23.38584}\NormalTok{) }\CommentTok{# coordinates of origin}
-\NormalTok{d =}\StringTok{ }\KeywordTok{c}\NormalTok{(-}\FloatTok{46.34078}\NormalTok{, -}\FloatTok{23.37746}\NormalTok{) }\CommentTok{# coordinates of destination}
-\KeywordTok{dist_google}\NormalTok{(}\DataTypeTok{from =} \NormalTok{o, }\DataTypeTok{to =} \NormalTok{d, }\DataTypeTok{mode =} \StringTok{"transit"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
+\begin{figure}[H]
+\includegraphics[width=1\linewidth]{box} \label{fig:box}
+\end{figure}
 \begin{verbatim}
@@
 
-The code above takes an origin (\texttt{o}) and destination (\texttt{d})
-and finds the travel time and distance. This is reported as 2,962 metres
-and 2,118 seconds (35 minutes) respectively. To ensure that the result
-was relevant to school travel, the times were calculated for arrival
-time at 9am on a weekday. The function was repeated for all
-origin-destination pairs. To ensure the correct OD data was being
-generated, the results were plotted on an interactive map (Figure 2).
-The results show the high variability of route distance versus Euclidean
-distance (`circuity'), providing further evidence of inequalities in
-public transport provision in the city (Figure 3).
-\begin{figure}
-\includegraphics[width=1\linewidth]{od-map} \caption{Visual representation of the 12,697 OD pairs routed through the Google Distance Matrix API on an interactive map in RStudio, an open source data analysis platform.}\label{fig:unnamed-chunk-4}
+The code above takes an origin (\texttt{o}) and destination (\texttt{d}) and finds the travel time and distance.
+This is reported as 2,962 metres and 2,118 seconds (35 minutes) respectively.
+To ensure that the result was relevant to school travel, the times were calculated for arrival time at 9am on a weekday.
+The function was repeated for all origin-destination pairs.
+To ensure the correct OD data was being generated, the results were plotted on an interactive map (Figure \ref{fig:pairs}).
+The results show the high variability of route distance versus Euclidean distance (`circuity'), providing further evidence of inequalities in public transport provision in the city (Figure \ref{fig:euclid}).
+\begin{figure}[H]
+\includegraphics[width=1\linewidth]{od-map} \caption{Visual representation of the 12,697 OD pairs routed through the Google Distance Matrix API on an interactive map in RStudio, an open source data analysis platform.}\label{fig:pairs}
 \end{figure}
-\begin{figure}
-\includegraphics[width=1\linewidth]{public-vs-euclid-dists} \caption{Euclidean (straight line) vs public transport distances resulting from routing the 12,697 OD pairs of interest through the Google Distance Matrix API.
+
+\begin{figure}[H]
+\includegraphics[width=0.5\linewidth]{public-vs-euclid-dists} \caption{Euclidean (straight line) vs public transport distances resulting from routing the 12,697 OD pairs of interest through the Google Distance Matrix API.
 The colours represent the interquantile range of the Euclidean distances travelled to school.}\label{fig:euclid}
@@
 
-\section{3. Area of study}\label{area-of-study}
+\section{Area of study}\label{area-of-study}
 
-The São Paulo Metropolitan Region (SPMR) is a large metropolis located
-in the South West of Brazil. It extends for over 7,700 squared
-kilometers and groups 38 different municipalities besides the
-municipality of São Paulo itself. The estimated population in 2010 was
-close to 19.5 million in 2010. The integrated transport system, to which
-users can access through an electronic card (Bilhete Único) is composed
-of a railway and bus network. The railway network comprises 78.4
-kilometers in five subway lines and six suburban railway lines which
-provide less frequent and slower service than the subway. Some of the
-bus network operates on dedicated bus corridors, which amounted to 500
-km in 2015. As can be seen in Figure 1, the richer areas of the SPMR are
-better served by public transport, while an important number of lower
-income areas is under-served.
+The São Paulo Metropolitan Region (SPMR) groups 38 different municipalities besides the municipality of São Paulo itself.
+The estimated population in 2010 was close to 19.5 million inhabitants.
+It has an integrated transport system, to which users can access through an electronic card (Bilhete Único) composed of a railway and bus network.
+The railway network comprises 78.4 kilometers in five subway lines and six suburban railway lines which provide less frequent and slower service than the subway.
+Some of the bus network operates on dedicated bus corridors, which amounted to 500 km in 2015.
+As can be seen in Figure \ref{fig:inc-trans}, the richer areas of the SPMR are better served by public transport, while an important number of lower income areas is under-served.
 
-\begin{figure}
+\begin{figure}[H]
 \includegraphics[width=0.8\linewidth]{income_pc} \caption{Public transport system and income per capita in São Paulo, 2010.}\label{fig:inc-trans}
@@
 
-Regarding the educational system, since the constitution of 1946,
-primary education is compulsory. In 2005, the legislation changed to
-increase the duration of the primary education, going from eight to nine
-years. In 2010, all regions finished the transition. In 2009, a
-constitutional amendment imposed compulsory education for the primary
-and secondary levels. The constitution imposes that this resolution
-should be fully implemented by 2016. By law, the state/municipalities
-should guaranty the adequate provision (100\% coverage of the demand).
-There are national and state educational funds that are distributed
-among the schools according to the number of enrolled students and
-existence of facilities (library, laboratories, etc.).
-
-The structure of the educational system is composed by both public and
-private schools. The public schools are divided in the following levels:
-Pre school (\textit{clase de alfabetização}) -- 4-5 year-olds (municipal
-administration); Fundamental (\textit{ensino fundamental}) comprises
-nine years, and the age range is usually 6-14 year-olds (municipal
-administration); Secondary (\textit{ensino médio}) lasts a minimum of
-three years, and the age range is 15-18 year-olds (state
-administration); and Post-secondary (\textit{ensino superior}) (state
-and federal administration).
+Regarding the educational system, primary education is compulsory.
+In 2005, the legislation changed to increase the duration of the primary education, going from eight to nine years.
+In 2010, all regions finished the transition.
+In 2009, a constitutional amendment imposed compulsory education for the primary and secondary levels.
+The constitution imposes that this resolution should be fully implemented by 2016.
+By law, the state/municipalities should guaranty the adequate provision (100\% coverage of the demand).
+There are national and state educational funds that are distributed among the schools per number of enrolled students and existence of facilities (library, laboratories, etc.).
+The structure of the educational system is composed by both public and private schools.
+The public schools are divided in the following levels: Pre school (classe de alfabetização) – 4-5 year-olds (municipal administration); Fundamental (ensino fundamental) comprises nine years, and the age range is usually 6-14 year-olds (municipal administration); Secondary (ensino médio) lasts a minimum of three years, and the age range is 15-18 year-olds (state administration); and Post-secondary (ensino superior) (state and federal administration).
 
@@
   \url{http://portal.inep.gov.br/basica-censo} for details on the School
-  Census}, there are 10,251 schools in the SPMR, out of which 54\% are
-public (22\% state / 32\% municipal) and 46\% are private.
+  Census}, there are 10,251 schools in the SPMR, out of which 54\% are public (22\% state / 32\% municipal) and 46\% are private.
 Geographically, students are assigned to the closest school to their
 place of residency or their parent's working place. For secondary
-school, the student can ask for a vacancy in a different school after
-providing a valid justification and under the condition of the
+school, the student can ask for a vacancy in a different school after providing a valid justification and under the condition of the
 availability of vacancy.
 
-The municipality is responsible for the organization of the pre-school
-and fundamental level. Poorer municipalities in SPMR have, in general,
-lower quality in terms of teachers and facilities. All the secondary
-schools are administrated by the state, nevertheless, differences exists
-depending of the location in the SPMR. In general, schools in peripheral
-locations show lower quality scores.
+The municipality is responsible for the organization of the pre-school and fundamental level.
+Poorer municipalities in SPMR have, in general, lower quality in terms of teachers and facilities.
+All the secondary schools are administrated by the state, nevertheless, differences exists depending of the location in the SPMR.
+In general, schools in peripheral locations show lower quality scores.
 
-As can be seen by comparing Figures 4 and 5, with Figure 1, students in
-the SPMR disproportionally attend private secondary schools in higher
-income areas, while the opposite is true for public secondary schools.
-\begin{table}[]\centering \caption{Modal split by type of secondary school}
+As can be seen by comparing Figures \ref{fig:inc-trans} and \ref{fig:students}, students in the SPMR disproportionally attend private secondary schools in higher income areas, while the opposite is true for public secondary schools.
+
+\begin{figure}[H]
+\includegraphics[width=1\linewidth]{med_pu} \caption{The home location of students attending public secondary schools students in the SPMR.}\label{fig:students}
+\end{figure}
+
+As Figure 5 shows, enrolment in public secondary schools is more
+dispersed spatially, as in principle public schools are not clearly
+geographically concentrated within the region. There is, however, an
+important degree of variation in the school seats in each sub-area,
+ranging from zero to 7,754.
+\begin{figure}
+\includegraphics[width=0.7\linewidth]{enrol} \caption{Enrollment in public secondary schools in the SPMR.}\label{fig:unnamed-chunk-7}
+\end{figure}
+Table 1 shows the modal split by type of secondary school student.\footnote{The source of this information is the 2007 Origin-Destination Household Travel Survey (O-D Survey), carried out by the São Paulo Metropolitan Transport Authority Mêtro.
+We focus on   trips made by secondary school students with an indicated motive   ``education'' at the destination.
+The total number of trips (excluding missing values) was 11,845.
+Of these, 5.5 percent were multimodal.
+In those cases, we assign the mode of the trip leg with the largest duration.
+We add all trip leg durations (in minutes) to obtain the total trip duration.}
+
+\begin{table}[H]\centering \caption{Modal split by type of secondary school}
 \begin{tabular}{lcc} \hline
@@
 \end{tabular}
-\end{table}
-\begin{figure}
-\includegraphics[width=0.7\linewidth]{med_pu} \caption{The home location of students attending public secondary schools students in the SPMR.}\label{fig:unnamed-chunk-5}
-\end{figure}\begin{figure}
-\includegraphics[width=0.7\linewidth]{med_pr} \caption{The home location of students attending private secondary schools students in the SPMR.}\label{fig:unnamed-chunk-6}
-\end{figure}
+\end{table} \label{split}
 
-As Figure 6 shows, enrolment in public secondary schools is more
-dispersed spatially, as in principle public schools are not clearly
-geographically concentrated within the region. There is, however, an
-important degree of variation in the school seats in each sub-area,
-ranging from zero to 7,754.
-\begin{figure}
-\includegraphics[width=0.7\linewidth]{enrol} \caption{Enrollment in public secondary schools in the SPMR.}\label{fig:unnamed-chunk-7}
-\end{figure}
-Table 1 shows the model split by type of secondary school
-student.\footnote{The source of this information is the 2007
-  Origin-Destination Household Travel Survey (O-D Survey), carried out
-  by the São Paulo Metropolitan Transport Authority Mêtro. We focus on
-  trips made by secondary school students with an indicated motive
-  ``education'' at the destination. The total number of trips (excluding
-  missing values) was 11,845. Of these, 5.5 percent were multimodal. In
-  those cases, we assign the mode of the trip leg with the largest
-  duration. We add all trip leg durations (in minutes) to obtain the
-  total trip duration.} Clearly, secondary school students attending
-private schools disproportionally travel by car, compared to public
-school students (9\% versus 47\%), even though higher income areas are
-better served by public transport. This is in line with the findings of
+Clearly, secondary school students attending private schools disproportionally travel by car, compared to public school students (47\% versus 9\%), even though higher income areas are better served by public transport.
+This is in line with the findings of
 S{á} et al. (2015). Around 67\% of public school students commute to
-school by active travel (mostly walking, since the percentage of biking
-is still relatively small), while 23\% commute by public transport,
-which has much longer mean duration than other modes. Gomide (2003),
-using data from the 1997 SPMR Origin-Destination survey, reports that
-the motive ``education'' explained at least 60\% of trips in the SPMR of
-individuals earning up to one minimum wage a month. This proportion
-decreases as income increases, as other motives such as work and
-leisure become more relevant.
+school by active travel (mostly walking, since the percentage of biking is still relatively small), while 23\% commute by public transport, which has much longer mean duration than other modes.
+Gomide (2003), using data from the 1997 SPMR Origin-Destination survey, reports that the motive ``education'' explained at least 60\% of trips in the SPMR of individuals earning up to one minimum wage a month.
+This proportion decreases as income increases, as other motives such as work and leisure become more relevant.
 
-Although the mean duration of travel to school remained stable over the
-period 1997-2007 (S{á} et al. 2015), travel-to-school times could
-increase as a result of several factors relevant to the SPMR, including
-sub-urbanization of particular demographic groups, such as those with
-higher income; increase in the size of schools leading to an increase in
-school's catchment area; less than proportional increase in schools in
-high population growth areas with poor connectivity; and institutional
-changes towards more flexible school choice (Easton and Ferrari 2015).
+Although the mean duration of travel to school remained stable over the period 1997-2007 (S{á} et al.
+2015), travel-to-school times could increase as a result of several factors relevant to the SPMR, including sub-urbanization of particular demographic groups, such as those with higher income; increase in the size of schools leading to an increase in school's catchment area; less than proportional increase in schools in high population growth areas with poor connectivity; and institutional changes towards more flexible school choice (Easton and Ferrari 2015).
 
-\section{4. Method: measuring school
+\section{Method: measuring school
 accessibility}\label{method-measuring-school-accessibility}
@@
 To define school accessibility, we start with the broad concept of
-accessibility, as used in the transport literature. According to Geurs
-and Van Wee (2004) (p.~128) accessibility is ``the extent to which
-land-use and transport systems enable (groups of) individuals to reach
-activities or destinations by means of a (combination of) transport
-mode(s).'' Extended to our case, school accessibility then measures the
-extent to which the existing built-environment and transport facilities
-enable children and adolescents in school age to reach schools.
+accessibility, as used in the transport literature.
+According to Geurs and Van Wee (2004) (p.~128) accessibility is ``the extent to which land-use and transport systems enable (groups of) individuals to reach activities or destinations by means of a (combination of) transport mode(s).'' Extended to our case, school accessibility then measures the extent to which the existing built-environment and transport facilities enable children and adolescents in school age to reach schools.
 
-Ideally, a school accessibility measure would include information on the
-home location of every student and the location of every school they
-could potentially attend (as in Andersson, Malmberg, and {Ö}sth (2012)),
-and some measure of commuting costs for any chosen mode (in terms of
-time and money). Given that this level of detail is not available in our
-data, we rely on area aggregates, implicitly assuming that the measure
-at the centroid of the area applies homogeneously within the area.
+Ideally, a school accessibility measure would include information on the home location of every student and the location of every school they could potentially attend (as in Andersson, Malmberg, and {Ö}sth (2012)), and some measure of commuting costs for any chosen mode (in terms of time and money).
+Given that this level of detail is not available in our data, we rely on area aggregates, implicitly assuming that the measure at the centroid of the area applies homogeneously within the area.
 
@@
                   1\quad if\quad C_{ij}<=t\\
-                  0\quad if\quad C_{ij}<t
+                  0\quad if\quad C_{ij}>t
                 \end{array}
@@
 
-This measure counts the number of ``opportunities'' \(O\), in our case
-schools available, from one area within a certain travel time threshold
-by mode \(M\). \(C_{ij}\) is the travel cost (measured in time) between
-the centroid of zone \(i\) and the centroid of zone \(j\), and
-\(f(C_{ij})\) is a weight function.
+This measure counts the number of ``opportunities'' \(O\), in our case schools available, from one area within a certain travel time threshold by mode \(M\).
+\(C_{ij}\) is the travel cost (measured in time) between the centroid of zone \(i\) and the centroid of zone \(j\), and \(f(C_{ij})\) is a weight function.
 
-This measure takes into account the spatial distribution of schooling
-opportunities, but not the local demand for schooling. This is
-particularly relevant for our analysis, since it could be the case that
-some areas are disproportionally served with respect to the number of
-potential students living within a certain travel distance. In order to
-assess the mismatch between the demand and supply for schooling, we use
-the sum of students in schools in each area (supply), and the sum of
-individuals within the school grade age-group living in each area
-(demand) in the following competitive accessibility measure (Shen 1998):
+This measure takes into account the spatial distribution of schooling opportunities, but not the local demand for schooling.
+This is particularly relevant for our analysis, since it could be the case that some areas are disproportionally served with respect to the number of potential students living within a certain travel distance.
+In order to assess the mismatch between the demand and supply for schooling, we use the sum of students in schools in each area (supply), and the sum of individuals within the school grade age-group living in each area (demand) in the following competitive accessibility measure (Shen 1998):
 
@@
 
-Where the numerator discounts the number of school seats in area \(j\),
-\(O_{j}\), by how far area \(i\) is from area \(j\) using the same
-function as before, and the denominator discounts the number of teens in
-secondary school age living in zone \(k\), \(T_{k}\), by how far they
-are from area \(j\). In other words, the numerator counts how many
-schools seats can be reached from an area, while the denominator counts
-how many teens can potentially reach the same area. In this way, the
-discounted number of school seats places at each area is divided by the
-potential students available to fill those places, and then summed in
-order to obtain a single accessibility measure for each area.
+Where the numerator discounts the number of school seats in area \(j\), \(O_{j}\), by how far area \(i\) is from area \(j\) using the same function as before, and the denominator discounts the number of teens in secondary school age living in zone \(k\), \(T_{k}\), by how far they are from area \(j\).
+In other words, the numerator counts how many schools seats can be reached from an area, while the denominator counts how many teens can potentially reach the same area.
+In this way, the discounted number of school seats places at each area is divided by the potential students available to fill those places, and then summed in order to obtain a single accessibility measure for each area.
 
 We calculate both measures for commutes to school by public transport.
-We set a threshold of 30 minutes based on the mean average travel times
-by this mode (see Table 1). We assign a minimum time of zero for
-commutes within the same area (i.e., schools within the same area are
-added to the index in all cases).
+We set a threshold of 30 minutes based on the mean average travel times by this mode (see Table \ref{split}).
+As a robustness check, in the result sections we show results for 35 and 40 minutes thresholds.
+We assign a minimum time of zero for commutes within the same area (i.e., schools within the same area are added to the index in all cases).
 
-\section{5. Results}\label{results}
+\section{Results}\label{results}
 
-Figure 7 shows the spatial distribution of the cumulative-opportunity
-measure. The index has a mean value of 3,777, which indicates that for
-teens residing in the average area, there are 3,777 school seats that
-can be reached within a journey by public transport of up to 30 minutes.
-The standard deviation is 3,485 indicates large variation in the level
-of accessibility across areas.
+\subsection{School accessibility}\label{results}
 
-As can be seen on the Figure 7, areas with low accessibility include not
+Figure \ref{fig:cum_ind} shows the spatial distribution of the cumulative-opportunity measure.
+The index has a mean value of 3,777, which indicates that for teens residing in the average area, there are 3,777 school seats that can be reached within a journey by public transport of up to 30 minutes.
+The standard deviation is 3,485 indicates large variation in the level of accessibility across areas.
+
+The results show that areas with low accessibility include not
 only the expanded financial business district located on the central
-south west area, where higher-income teens mostly attend private school
-(see Figures 1 and 5), but also areas with deficient public transport
-provision, such as some north and south eastern areas.
+south west area, where higher-income teens mostly attend private school (see Figures \ref{fig:inc-trans} and \ref{fig:students}), but also areas with deficient public transport provision, such as some north and south eastern areas.
 
-\begin{figure}
-\includegraphics[width=0.7\linewidth]{cum_index} \caption{Cumulative opportunity index}\label{fig:unnamed-chunk-8}
+\begin{figure}[H]
+\includegraphics[width=0.7\linewidth]{cum_index} \caption{Cumulative opportunity index}\label{fig:cum_ind}
 \end{figure}
 
-Figure 8 shows the spatial distribution of the competitive accessibility
-measure. Although it largely follows the distribution of the cumulative
-index, it corrects the value in some areas based on the place of
-residence of students across the SPRM. To interpret the value of the
-index, recall that if all teens in an area attended secondary schools in
-public secondary schools in their area (i.e., no student would attend
-private school or travel outside her own area to attend school), the
-index would take the value of 1. Then, in areas with a value lower than
-one, the provision of public schools and public transport is such that
-there is local under-provision, in the sense that the number of schools
-seats that can be reached from the area is smaller than the number of
-teens that can potentially reach the area (the opposite is true for
-values higher than one). Note that some relatively close by areas,
-especially in the eastern part of the city, have striking differences in
-the level of accessibility. In some cases, an area can have twice the
-level of accessibility than the most proximate area. This can be due to
-differences in the local provision of public schools, public transport
-or the number of students living within the reach of the areas.
-\begin{figure}
-\includegraphics[width=0.7\linewidth]{comp_index} \caption{Competitive accessibility index}\label{fig:unnamed-chunk-9}
+Figure \ref{fig:comp_ind} shows the spatial distribution of the competitive accessibility measure.
+Although it largely follows the distribution of the cumulative index, it corrects the value in some areas based on the place of residence of students across the SPMR.
+To interpret the value of the index, recall that if all teens in an area attended secondary schools in public secondary schools in their area (i.e., no student would attend private school or travel outside her own area to attend school), the index would take the value of 1.
+Then, in areas with a value lower than one, the provision of public schools and public transport is such that there is local under-provision, in the sense that the number of schools seats that can be reached from the area is smaller than the number of teens that can potentially reach the area (the opposite is true for values higher than one).
+Note that some relatively close by areas, especially in the eastern part of the city, have striking differences in the level of accessibility.
+In some cases, an area can have twice the level of accessibility than the most proximate area.
+This can be due to differences in the local provision of public schools, public transport or the number of students living within the reach of the areas.
+
+\begin{figure}[H]
+\includegraphics[width=0.7\linewidth]{comp_index} \caption{Competitive accessibility index}\label{fig:comp_ind}
 \end{figure}
+
+\subsection{Policy experiment}\label{results}
+
 To understand the effects of changes in public school provision on
-school accessibility, we conduct a comparative static analysis using our
-competitive index. We compare the changes in school-accessibility by
-accessibility quantiles before and after the implementation of a policy
-that reallocates students attending school in areas with lower than
-average provision to schools in areas with higher than average
-provision. The experiment mimics a policy that aims at increasing scale
-economies in the provision of public education by centralizing provision
-in larger educational establishments, which is the rationale behind the
-policy proposal made by the São Paulo state government (Ortellado 2015).
+school accessibility, we conduct a comparative static analysis using our competitive index.
+The experiment mimics a policy that aims at increasing scale economies in the provision of public education by centralizing provision in larger educational establishments, which is the rationale behind the policy proposal made by the São Paulo state government (Ortellado 2015).
+In this spirit, instead of selecting a specific number of schools to be closed or students to be reallocated, we compare the changes in school-accessibility before and after the implementation of a policy that closes down schools which fall under a specified size threshold.
+The threshold  choice is based on the quantiles of the school size distribution of all schools in the area.
+Figure \ref{density} shows the density plot of the number of students and three different thresholds: 15\%, 25\% and 50\%.
+
+\begin{figure}[H]
+\includegraphics[width=0.7\linewidth]{density} \caption{Competitive accessibility index}\label{fig:density}
+\end{figure}
+
+Table \ref{thresholds} summarizes the size threshold corresponding to each scenario, the number of schools which would close and the number of students which would be reallocated in each case.
+
+\begin{table}[H]\caption{Policy experiment scenarios}\label{thresholds}
+\begin{tabular}[c]{lccc}\hline
+Scenario & Size threshold & \# Schools closed & \# Students reallocated \\ \hline
+15\% & 337 & 343 & 8,002 \\
+25\% & 572 & 711 & 36,807 \\
+50\% & 1,061 & 1691 & 163,231 \\\hline
+\end{tabular}
+\end{table}
 
 To implement the policy experiment, we assign students in areas with
-lower than average enrolment to the next area with higher than average
-enrolment, while keeping the distribution of teens and public transport
-constant. In other words, total enrolment is the same before and after
-the change, but the distribution of public secondary schools increases
-in areas where it was already higher than average, and disappears in
-areas where it was lower than average. The change is equivalent to
-closing down 1,691 schools, or reallocating 163,231 students. Although
-seemingly extreme, it is worth noting again that the proposal of the São
-Paulo state government was estimated to impact 300,000 secondary
-students attending public school (Ortellado 2015). We then compare the
-changes in school accessibility for different quartiles before and after
-this change. Table 2 summarizes the results.
+lower than average enrolment to the next area with higher than average enrolment, while keeping the distribution of teens and public transport constant.
+In other words, total enrolment is the same before and after the change, but the distribution of public secondary schools increases in areas where it was already higher than average, and disappears in areas where it was lower than average.
+In the most extreme scenario, the change is equivalent to closing down 1,691 schools, or reallocating 163,231 students.
+Although seemingly massive, it is worth noting again that the proposal of the São Paulo state government was estimated to impact 300,000 secondary students attending public school (Ortellado 2015).
+We then compare the changes in school accessibility for different quartiles before and after this change.
+Table \ref{policy_change} summarizes the results for a 30 minute threshold in the Competitive Accessibility Index.
+Table \ref{sen1} in the Appendix presents the results for 35 and 40 minutes thresholds.
 
-\begin{longtable}[c]{@{}lrrr@{}}
+\begin{longtable}[H]{@{}lrrr@{}} 
 \caption{Mean accessibility before and after
@@
 \bottomrule
-\end{longtable}
+\end{longtable} \label{policy_change}
 
-\newpage
+The policy reduces further the accessibility level of areas in the first and second quantiles of accessibility.
+Areas in the first quantile experience an average reduction of 21\% in school accessibility levels as a result of the changes in the spatial distribution of schools.
+The average reduction is 17\% and 14\% when 35 and 40 minutes thresholds are used instead of a 30 minute threshold (see Table \ref{sen1} in the Appendix).
+In
+the mean time, areas that had high levels of accessibility to start
+with, experience an increase, which varies between a maximum of 28\% for the fourth quantile when a 30 minute threshold is chosen, to 19\% when a 40 minute threshold is used instead.
+The results suggest that concentrating the provision of public secondary schools accentuates the already large differences in accessibility levels across areas.
 
-The policy reduces further the accessibility level of areas in the first
-and second quantiles of accessibility. Areas in the first quantile
-experience an average reduction of 21\% in school accessibility levels
-as a result of the changes in the spatial distribution of schools. In
-the mean time, areas that had high levels of accessibility to start
-with, experience an increase. The results suggest that concentrating the
-provision of public secondary schools accentuates the already large
-differences in accessibility levels across areas.
+\section{Conclusion and discussion}\label{conclusion-and-discussion}
 
-\section{6. Conclusion and discussion}\label{conclusion-and-discussion}
-
-The policy experiment results suggest that a policy that concentrates
-the provision of public schools spatially is regressive. Young people
-living in areas with the lowest levels of accessibility experience the
-largest negative impact from the policy, compounding issues of low local
-schooling provision, low access by public transport and/or high
-competition to access good schools. The policy would increase commuting
-times for students residing in areas with low accessibility. This would
-likely have further negative knock-on effects in terms of attendance,
+The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive.
+Young people living in areas with the lowest levels of accessibility experience the largest negative impact from the policy, compounding issues of low local schooling provision, low access by public transport and/or high competition to access good schools.
+The policy would increase commuting times for students residing in areas with low accessibility.
+This would likely have further negative knock-on effects in terms of attendance,
 drop-out likelihood and performance.
 
-Although the simulated scenario represents a rather extreme change in
-the distribution of public schooling, the way in which students with
-different levels of accessibility are impacted remains, regardless of
-the level of concentration. In the absence of a public transport subsidy
-for school commutes, students who experience a substantial increase in
-travel-to-school times may be forced to change their commute mode from
-an active (and free of charge mode) to a paid-by mode, with the
-consequent increase in their monetary spent in commuting to school. The
-most negative effect is then felt by low-incomes students living in
-areas with low school accessibility, for whom the marginal increase in
-commuting costs represents a much larger share in their budget.
+Although the simulated scenario represents a rather extreme change in the distribution of public schooling, the way in which students with different levels of accessibility are impacted remains, regardless of the level of concentration.
+In the absence of a public transport subsidy for school commutes, students who experience a substantial increase in travel-to-school times may be forced to change their commute mode from an active (and free of charge mode) to a paid-by mode, with the consequent increase in their monetary spent in commuting to school.
+The most negative effect is then felt by low-incomes students living in areas with low school accessibility, for whom the marginal increase in commuting costs represents a much larger share in their budget.
 
@@
 unlikely to compensate students at the bottom of the income and
-accessibility ladder if the increase in commuting times has consequences
-in terms of motivation to stay in school and performance. In this sense,
-a ``trade-off'' logic where less schooling provision is compensated by
-public transport subsidies does not operate when existing inequalities
-in provision are large.
+accessibility ladder if the increase in commuting times has consequences in terms of motivation to stay in school and performance.
+In this sense, a ``trade-off'' logic where less schooling provision is compensated by public transport subsidies does not operate when existing inequalities in provision are large.
 
-Our study could be further extended to account for other modes of school
-commute, such as active travel. It would benefit from data with higher
-spatial resolution to capture within-area inequalities, which may be
-substantial but currently hidden away in area aggregates. A future
-extension of the index could also incorporate measures of school
-quality. Furthermore, it would be informative to calculate the proposed
-indices for other contexts and different time periods, to understand the
-extent to which the regressive nature of school agglomeration is a
-universal finding or specific to the SPMR.
+Our study could be further extended to account for other modes of school commute, such as active travel.
+It would benefit from data with higher spatial resolution to capture within-area inequalities, which may be substantial but currently hidden away in area aggregates.
+A future extension of the index could also incorporate measures of school quality.
+Furthermore, it would be informative to calculate the proposed indices for other contexts and different time periods, to understand the extent to which the regressive nature of school agglomeration is a universal finding or specific to the SPMR.
 
-We conclude that extending the concept of local accessibility indicators
-to education can help to both contest and constructively tackle embedded
-social inequalities. Contestation can be seen as a powerful tool to
-prevent the implementation of regressive policies that threaten to
-accentuate already large disparities. Quantitative measures such as
-those proposed here can help demonstrate how multiple dimensions of
-inequality amplify inequalities, via a single interpretable measure. The
-methods demonstrated could be useful for policy makers and campaigners
-to negotiate policies with a highly sensitive social component and
-evident short and long-term consequences for those affected.
+We conclude that extending the concept of local accessibility indicators to education can help to both contest and constructively tackle embedded social inequalities.
+Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities.
+Quantitative measures such as those proposed here can help demonstrate how multiple dimensions of inequality amplify inequalities, via a single interpretable measure.
+The methods demonstrated could be useful for policy makers and campaigners to negotiate policies with a highly sensitive social component and evident short and long-term consequences for those affected.
+
+\section*{Appendix}
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+\setcounter{table}{0}
+\renewcommand{\thetable}{A\arabic{table}}
+
+\begin{table}[H]\caption{Mean accessibility before and after
+redistribution, different time thresholds}\label{sen1}
+\begin{tabular}[c]{lcccccc}\hline
+ & \multicolumn{2}{c}{Pre-change} & \multicolumn{2}{c}{Post-change} & \multicolumn{2}{c}{ Change (\%)} \\ \hline
+Quantile & 35 min & 40 min & 35 min & 40 min& 35 min & 40 min \\ \hline 
+1st & 0.56 & 0.638 & 0.46 & 0.55 & -16.92 & -13.61 \tabularnewline
+2nd & 1.06 & 1.23 & 1.05 & 1.209 & -0.54 & -1.85 \tabularnewline
+3rd & 2.22 & 2.51 & 2.52 & 2.773 & 13.64 & 10.72 \tabularnewline
+4th& 5.26 & 6.69 & 7.22 & 7.928 & 37.29 & 18.55 \tabularnewline
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]\caption{Mean accessibility before and after
+redistribution, different allocation rule}\label{sen2}
+\begin{tabular}[c]{lccccccc}\hline
+ & \multicolumn{2}{c}{Pre-change} & \multicolumn{2}{c}{Post-change} & \multicolumn{2}{c}{ Change (\%)} \\ \hline
+Percentile & < 0.15  & < 0.25 &  < 0.15  & < 0.25 & < 0.15  & < 0.25 \\ \hline 
+1st & 0.446 & 0.421 &  0.443 & 0.397 & -0.70 & -5.59 \tabularnewline
+2nd & 0.844 & 0.832 &  0.838 & 0.820 & -0.70 & -1.42 \tabularnewline
+3rd & 1.843 & 1.92 &  1.859 & 2.00 & 0.83 & 4.37 \tabularnewline
+4th & 4.280 & 4.30 &  4.280 & 4.32 & 0 & 0.37 \tabularnewline
+\bottomrule
+\end{tabular}
+\end{table}
 
@@
 
-Allen, W Bruce, Dong Liu, and Scott Singer. 1993. ``Accesibility
+Allen, W Bruce, Dong Liu, and Scott Singer. 1993. ``Accessibility
 Measures of US Metropolitan Areas.'' \emph{Transportation Research Part

### 2017-04-02 paper_v2.tex
Improve text

@@
 For this work we used the Google Matrix Distance API
-(API stands for `application programming interface' and is used for
-computer-computer communication over the web), a multi-modal
-real-time travel data provider (see \url{https://developers.google.com/maps/documentation/distance-matrix/}). 
+(application programming interface), a multi-modal
+real-time travel data provider (see \href{https://developers.google.com/maps/documentation/distance-matrix/}{developers.google.com/maps/}). 
 The API was used in preference to other options,
-such as Routino, described in a paper by Singleton (2014),
+such as Routino --- see Singleton (2014) for a used case involving large-scale accessibility analysis ---
 and the OpenStreetMap Routing Machine (OSRM)

### 2017-04-02 paper_v2.tex
Improve discussion of Ingram (1971)

@@
 
-The first well-known attempt to define accessibility quantitatively was by Ingram (1971), which presented a range of measures related to
-distance (Euclidean and network), barriers and different functions
-representing distance decay. This early work made the distinction
+The first well-known quantitative definition of accessibility was by Ingram (1971).
+This seminal paper presented a range of measures determined by
+distances to destinations (Euclidean and network), natural barriers and 
+distance decay functions. This early work made the distinction
 between accessibility indices that apply to zones or single `desire

### 2017-04-02 paper_v2.tex
Update title
Update abstract

@@
 \begin{frontmatter}
- \title{The unequal distribution of access to education and transport: insights from the São Paulo Metropolitan Region}
+ \title{Public transport and school location impacts on educational inequalities: insights from São Paulo}
 
@@
 In many large Latin American urban areas such as the São Paulo
-Metropolitan Region (SPMR), growing social and economic inequalities are embedded through high spatial inequality in the provision of education and transport systems.
-This paper sheds light on the transport-education inequality nexus with reference to a new school accessibility measure applied to the SPMR.
-Our index combines information on the spatial distribution of adolescents, the location of existing schools, and the travel infrastructure serving the school catchment area into a single measure.
+Metropolitan Region (SPMR), growing social and economic inequalities are embedded through high spatial inequality in the provision of state schools and affordable public transport to these schools.
+This paper sheds light on the transport-education inequality nexus with reference to school accessibility in the SPMR.
+To assess school accessibility, we develop an accessibility index which combines information on the spatial distribution of adolescents, the location of existing schools, and the public transport provision serving the school catchment area into a single measure.
 The index is used to measure school accessibility locally across 633 areas within the SPMR.

### 2017-04-02 paper_v2.tex
Add sentence to abstract

@@
 We use the index to simulate the impact of a policy aiming at increasing the centralization of public secondary education provision, and find that it negatively affects students with the lowest levels of accessibility.
-These results illustrate how existing inequalities are amplified by variable accessibility to schools across income groups and geographical space.
+These results illustrate how existing inequalities can be amplified by variable accessibility to schools across income groups and geographical space.
+The research suggests that educational inequality impacts of school agglomeration policies should be considered before centralisation takes place and highlights the links between transport planning social injustice that.
 \end{abstract}

### 2017-04-02 paper_v2.tex
Add new reference
Veitch, J., A. Carver, J. Salmon, G. Abbott, K. Ball, D. Crawford, V. Cleland, and A. Timperio. 2017. “What Predicts Children’s Active Transport and Independent Mobility in Disadvantaged Neighborhoods?” Health & Place 44 (March): 103–9. doi:10.1016/j.healthplace.2017.02.003.

@@
 Metropolitan Region (SPMR), growing social and economic inequalities are embedded through high spatial inequality in the provision of state schools and affordable public transport to these schools.
-This paper sheds light on the transport-education inequality nexus with reference to school accessibility in the SPMR.
+This paper sheds light on the transport-education inequality nexus with reference to school accessibility by public transport in the SPMR.
 To assess school accessibility, we develop an accessibility index which combines information on the spatial distribution of adolescents, the location of existing schools, and the public transport provision serving the school catchment area into a single measure.
 The index is used to measure school accessibility locally across 633 areas within the SPMR.
-We use the index to simulate the impact of a policy aiming at increasing the centralization of public secondary education provision, and find that it negatively affects students with the lowest levels of accessibility.
+We use the index to simulate the impact of a policy aiming at increasing the centralization of public secondary education provision, and find that it negatively affects public transport accessibility for students with the lowest levels of accessibility.
 These results illustrate how existing inequalities can be amplified by variable accessibility to schools across income groups and geographical space.
@@
 The SPMR presents an excellent opportunity to measure the extent of
-educational and transport infrastructure, and to study the impact of
+educational and public transport provision, and to study the impact of
 public policies on different socio-economic groups.
@@
 Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (Ortellado 2015).
+It was this policy context that informed the paper's focus on public transport, rather than the topic of active travel to school, which has gained recent prominence in Australia (e.g.
+Veitch et al., 2017) the USA and Europe
 
@@
 
+Veitch, J., A.
+Carver, J.
+Salmon, G.
+Abbott, K.
+Ball, D.
+Crawford, V.
+Cleland, and A.
+Timperio.
+2017.
+``What Predicts Children’s Active Transport and Independent Mobility in Disadvantaged Neighborhoods?'' \emph{Health \& Place} 44 (March): 103–9.
+doi:10.1016/j.healthplace.2017.02.003.
+
 \end{document}

### 2017-04-02 paper_v2.tex
Justify emphasis on PT

@@
 Protesters scored some victories: the reform was postponed for one year, and the education secretary resigned (Ortellado 2015).
-It was this policy context that informed the paper's focus on public transport, rather than the topic of active travel to school, which has gained recent prominence in Australia (e.g.
-Veitch et al., 2017) the USA and Europe
 
@@
 In this dynamic and highly political context, new quantitative evidence can help shed light on the relationship between transport and educational inequalities and the extent to which they are mutually reinforcing.
-Moreover, the results should help design policies that are more inclusive and equitable by highlighting areas where the
-provision of schools, travel to school opportunities, or both, are
-particularly restrictive.
+Moreover, the results should help design school location and public transport policies that are more inclusive and equitable by highlighting areas where
+school location and public transport opportunities conspire to exacerbate educational inequalities.
+
+The recent experience of São Paulo and other cities informed the paper's public transport emphasis. 
+This could be seen as a contrast to the topic of active travel to school, which has gained recent prominence in Australia (e.g.
+Veitch et al., 2017) the USA (Lee et al., 2017) and Europe (e.g.
+Macdonald et al., 2016).
+However, we see great potential synergies between provision of public and active transport opportunities and, by focussing on public transport, do not wish to endorse an `either or' approach to adequate school travel opportunities.
+The focus on public transport is especially appropriate in a Latin American context where parents' perception of road danger may (rightly) be greater and where parents cannot be expected automatically to be able to have the resources to buy and maintain a bike for children to cycle to school (public provision of safe walking and cycling infrastructure and bicycles are areas of research worthy of attention beyond the scope of this paper).
 
@@
 
+Lee, Richard J., Ipek N. Sener, and S. Nathan Jones. 2017. ``Understanding the Role of Equity in Active Transportation Planning in the United States.'' \emph{Transport Reviews} 37 (2): 211–26.
+
 Lovelace, Robin, Richard Ellison, Barry Rowlingson, and Nick Bearman.
@@
 Practice} 85. Elsevier: 157--73.
+
+Macdonald, Laura, Paul McCrorie, Natalie Nicholls, and Anne Ellaway. 2016. ``Walkability around Primary Schools and Area Deprivation across Scotland.'' \emph{BMC Public Health} 16: 328.
 

### 2017-04-02 paper_v2.tex
Add why not using price

@@
 API returns
-route distance, price and times for journeys between origins and
+route distance, price (but not price for travel to school) and times for journeys between origins and
 destinations, provided either as Longitude/Latitude pairs or
@@
 In this sense, a ``trade-off'' logic where less schooling provision is compensated by public transport subsidies does not operate when existing inequalities in provision are large.
+The use of travel times rather than economic cost raises the question: is affordability or accessibility more important in affecting educational inequalities?
+The focus on the time data from the Google Matrix API reflects the greater robustness of this measure (prices can fluctuate rapidly) and the fact that, thanks to the Free Pass Movement, there is a widespread perception that public transport to schools should be free.
+The analysis in this paper therefore assumes very low or zero financial cost for public transport to school.
 

### 2019-03-18 paper.Rmd

@@
 
-The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive, in the sense that teens living in areas that had the lowest levels of accessibility are the ones who experience the largest negative impact from the policy.
-They are likely to be worse off compared to an already difficult situation of low local schooling provision, low access by public transport and/or high competition by teens in other areas.
-In practice, such a change would probably imply larger commuting times for students residing in areas with low accessibility, which can have further consequences in terms of attendance, drop-out likelihood and performance.
+The policy experiment results suggest that a policy that concentrates the provision of public schools spatially is regressive.
+Young people living in areas with the lowest levels of accessibility experienced the largest negative impact from the policy, compounding issues of low local schooling provision, low access by public transport and/or high competition to access good schools.
+The policy would increase commuting times for students residing in areas with low accessibility.
+This would likely have further negative knock-on effects in terms of attendance, drop-out likelihood and performance.
 
-Although the simulated scenario represents a rather extreme change in the distribution of public schooling, the way in which students with different levels of accessibility are impacted remains.
-In the absence of a public transport subsidy for school commutes, students that experience a significant increase in travel-to-school times may be force to change their commute mode from an active (and free of charge mode) to a paid-by mode, with the consequent increase in their monetary spent in commuting to school.
+Although the simulated scenario represents a rather extreme change in the distribution of public schooling, the way in which students with different levels of accessibility are impacted remains, regardless of the level of concentration.
+In the absence of a public transport subsidy for school commutes, students who experience a substantial increase in travel-to-school times may be forced to change their commute mode from an active (and free of charge mode) to a paid-by mode, with the consequent increase in their monetary spent in commuting to school.
 The most negative effect is then felt by low-incomes students living in areas with low school accessibility, for whom the marginal increase in commuting costs represents a much larger share in their budget.
@@
 
-Our study can be further extended to account for other modes of school commute, such as active travel.
-It can also benefit from a finer spatial disaggregation to capture within area inequalities, which may be substantial and are currently hidden away in area aggregates.
-A future extension of the index can incorporate a measure of school quality, which in so far assumed away.
-It will also be highly informative to calculate the proposed indices for other contexts and for different time periods in order to understand the true magnitude of the extent of inequality across individuals living in different areas of the SPMR.
+Our study could be further extended to account for other modes of school commute, such as active travel.
+It would benefit from data with higher spatial resolution to capture within-area inequalities, which may be substantial but currently hidden away in area aggregates.
+A future extension of the index could also incorporate measures of school quality.
+Furthermore, it would be informative to calculate the proposed indices for other contexts and different time periods, to understand the extent to which the regressive nature of school agglomeration is a universal finding or specific to the SPMR.
 
@@
 Contestation can be seen as a powerful tool to prevent the implementation of regressive policies that threaten to accentuate already large disparities.
-Quantitative measures such as the ones proposed here help summarizing multiple dimensions of inequality into a single, interpretable measure.
-Having this tool may be useful for both policy makers and social groups in discussing and negotiating policies with a highly sensitive social component and evident short and long-term consequences for those affected.
+Quantitative measures such as those proposed here can help demonstrate how multiple dimensions of inequality amplify inequalities, via a single interpretable measure.
+The methods demonstrated could be useful for policy makers and campaigners to negotiate policies with a highly sensitive social component and evident short and long-term consequences for those affected.
 
